```{r include=FALSE}
my_colors <- list(blue = "#4a5F70",
                  beige = "#7f825f",
                  mauve = "#c2ae95",
                  red = "#824e4e",
                  grey = "#66777d")
```

# Probability {#probability}

## Principles

Here are three rules that come up all the time.

* $Pr(A \cup B) = Pr(A)+Pr(B) - Pr(AB)$.  This rule generalizes to 
$Pr(A \cup B \cup C)=Pr(A)+Pr(B)+Pr(C)-Pr(AB)-Pr(AC)-Pr(BC)+Pr(ABC)$.

* $Pr(A|B) = \frac{P(AB)}{P(B)}$

* If A and B are independent, $Pr(A \cap B) = Pr(A)Pr(B)$, and $Pr(A|B)=Pr(A)$. 

Uniform distributions on finite sample spaces often reduce to counting the elements of *A* and the sample space *S*, a process called combinatorics.  Here are three important combinatorial rules.

**Multiplication Rule**.  $|S|=|S_1 |⋯|S_k|$.

*How many outcomes are possible from a sequence of 4 coin flips and 2 rolls of a die?*
$|S|=|S_1| \cdot |S_2| \dots |S_6| = 2 \cdot 2 \cdot 2 \cdot 2 \cdot 6 \cdot 6 = 288$.

*How many subsets are possible from a set of n=10 elements?*
In each subset, each element is either included or not, so there are $2^n = 1024$ subsets.

*How many subsets are possible from a set of n=10 elements taken k at a time with replacement?*
Each experiment has $n$ possible outcomes and is repeated $k$ times, so there are $n^k$ subsets.

**Permutations**.  The number of *ordered* arrangements (permutations) of a set of $|S|=n$ items taken $k$ at a time *without* replacement has $n(n-1) \dots (n-k+1)$ subsets because each draw is one of k experiments with decreasing number of possible outcomes.  

$$_nP_k = \frac{n!}{(n-k)!}$$

Notice that if $k=0$ then there is 1 permutation; if $k=1$ then there are $n$ permutations; if $k=n$ then there are $n!$ permutations.

*How many ways can you distribute 4 jackets among 4 people?*
$_nP_k = \frac{4!}{(4-4)!} = 4! = 24$

*How many ways can you distribute 4 jackets among 2 people?*
$_nP_k = \frac{4!}{(4-2)!} = 12$

**Subsets**.  The number of *unordered* arrangements (combinations) of a set of $|S|=n$ items taken $k$ at a time *without* replacement has 

$$_nC_k = {n \choose k} = \frac{n!}{k!(n-k)!}$$

combinations and is called the binomial coefficient.  The binomial coefficient is the number of different subsets.  Notice that if k=0 then there is 1 subset; if k=1 then there are n subsets; if k=n then there is 1 subset.  The connection with the permutation rule is that there are $n!/(n-k)!$ permutations and each permutation has $k!$ permutations.  

*How many subsets of 7 people can be taken from a set of 12 persons?*
$_{12}C_7 = {12 \choose 7} = \frac{12!}{7!(12-7)!} = 792$

*If you are dealt five cards, what is the probability of getting a "full-house" hand containing three kings and two aces (KKKAA)?*
$$P(F) = \frac{{4 \choose 3} {4 \choose 2}}{{52 \choose 5}}$$

**Distinguishable permutations**.  The number of *unordered* arrangements (distinguishable permutations) of a set of $|S|=n$ items in which $n_1$ are of one type, $n_2$ are of another type, etc., is 

$${n \choose {n_1, n_2, \dots, n_k}} = \frac{n!}{n_1! n_2! \dots n_k!}$$

*How many ordered arrangements are there of the letters in the word PHILIPPINES?*  There are n=11 objects.  $|P|=n_1=3$;  $|H|=n_2=1$; $|I|=n_3=3$; $|L|=n_4=1$; $|N|=n_5=1$; $|E|=n_6=1$; $|S|=n_7=1$.

$${n \choose {n_1, n_2, \dots, n_k}} = \frac{11!}{3! 1! 3! 1! 1! 1! 1!} = 1,108,800$$

*How many ways can a research pool of 15 subjects be divided into three equally sized test groups?*

$${n \choose {n_1, n_2, \dots, n_k}} = \frac{15!}{5! 5! 5!} = 756,756$$


## Likelihood

The *likelihood function* is the likelihood of a parameter $\theta$ given an observed value.  The likelihood function is identical to the probability distribution function, except that it reverses which variable is considered fixed.  E.g., the binomial distribution $Bin(n, \pi)$ expresses the probability that $X = x$ given a sample of size $n$ from a binomial distribution with success probability $\pi$.

$$f(x|\theta) = P(X=x|\pi) = \frac{{n!}}{{x!(n-x)!}} p^x (1-p)^{n-x}.$$

The corresponding *likelihood* function expresses the probability that $\pi = p$ given the observed value $x$.

$$L(\theta|x) = L(\pi|x) = \frac{{n!}}{{x!(n-x)!}} \pi^x (1-\pi)^{n-x} \propto \pi^x (1-\pi)^{n-x}.$$

Any multiplicative constant which does not depend on $\theta$ is irrevelant and can be discarded, so

$$L(\pi|x) =  \pi^x (1-\pi)^{n-x}$$

In most cases, you work with the *loglikelihood*, $l(\theta|x) = \log L(\theta|x)$.  So for the binomial distribution, $l(\pi|x) = x \log \pi + (n - x) \log (1 - \pi)$.  Frequently you derive loglikelihood from a sample.  The overall likelihood is the product of the individual likelihoods, and the overall loglikelihood is the log of the overall likelihood.

$$l(\theta|x) = \log \prod_{i=1}^n f(x_i|\theta)$$

You usually want to know the *maximum* of the likelihood function.  The value of $\theta$ that maximizes $L$ is the maximum-likelihood estimator of $\theta$.  Suppose an experiment had $n=5$ Bernoulli trials, and $X = 3$ where successes.  $X \sim Bin(5, p)$.  What is the MLE?  A plot of $L(p|x) = p^3(1 - p)^2$ is maximized at $p = 0.6$.

```{r}
library(tidyverse)
p <- 0:100*.01
L <- p^3 * (1 - p)^2
dat <- data.frame(L, p)
ggplot(dat, aes(x = p, y = L)) +
  geom_line() +
  geom_vline(xintercept = 0.6)
```

Calculating the MLE is usually simplified using the loglikelihood function.


## Discrete Distributions

### Binomial

If $X$ is the count of successful events in $n$ identical and independent Bernoulli trials of success probability $\pi$, then $X$ is a random variable with a binomial distribution $X \sim  Bin(n,\pi)$ with mean $\mu=n\pi$ and variance $\sigma^2 = n\pi(1-\pi)$. The probability of $X=x$ successes in $n$ trials is 

$$P(X=x) = \frac{{n!}}{{x!(n-x)!}} \pi^x (1-\pi)^{n-x}.$$

*What is the probability 2 out of 10 coin flips are heads if the probability of heads is 0.3?*

Function `dbinom()` calculates the binomial probability.

```{r}
dbinom(x = 2, size = 10, prob = 0.3)
```

A simulation of n = 10,000 random samples of size 10 gives a similar result.  `rbinom()` generates a random sample of numbers from the binomial distribution.

```{r message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
library(tidyverse)

data.frame(cnt = rbinom(n = 10000, size = 10, prob = 0.3)) %>%
  count(cnt) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         X_eq_x = cnt == 2) %>%
  ggplot(aes(x = as.factor(cnt), y = n, fill = X_eq_x, label = pct)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c(my_colors$grey, my_colors$red)) +
  geom_label(aes(label = round(pct, 2)), size = 3, alpha = .6) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Binomial Distribution", 
       subtitle = paste0("P(X=2) successes in 10 trials when p = 0.3 is ", round(dbinom(2, 10, 0.3), 4), "."),
       x = "Successes",
       y = "Count",
       caption = "Simulation from n = 10,000 binomial random samples.") 
```

*What is the probability of <=2 heads in 10 coin flips where probability of heads is 0.3?*

The cumulative probability is the sum of the first three bars in the simulation above.  Function `pbinom()` calculates the *cumulative* binomial probability.

```{r}
pbinom(q = 2, size = 10, prob = 0.3, lower.tail = TRUE)
```

*What is the expected number of heads in 25 coin flips if the probability of heads is 0.3?*  

The expected value, $\mu = np$, is `r 25 * .3`.  Here's an empirical test from 10,000 samples.

```{r}
mean(rbinom(n = 10000, size = 25, prob = .3))
```

The variance, $\sigma^2 = np (1 - p)$, is `r 25 * .3 * (1 - .3)`.  Here's an empirical test.

```{r}
var(rbinom(n = 10000, size = 25, prob = .3))
```

*Suppose X and Y are independen random variables distributed X ~ b(10, .6) and Y ~ b(10, .7).  What is the probability that either variable is <=4?*

Let $P(A) = P(X<=4)$ and $P(B) = P(Y<=4)$.  Then $P(A|B) = P(A) + P(B) - P(AB)$, and because the events are independent, $P(AB) = P(A)P(B)$.

```{r}
p_a <- pbinom(q = 4, size = 10, prob = 0.6, lower.tail = TRUE)
p_b <- pbinom(q = 4, size = 10, prob = 0.7, lower.tail = TRUE)
p_a + p_b - (p_a * p_b)
```

Here's an empirical test.

```{r}
df <- data.frame(
  x = rbinom(10000, 10, 0.6),
  y = rbinom(10000, 10, 0.7)
  )
mean(if_else(df$x <= 4 | df$y <= 4, 1, 0))
```


### Negative-Binomial

If $X$ is the count of trials required to reach a target number $r$ of successful events in identical and independent Bernoulli trials of success probability $p$, then $X$ is a random variable with a negative-binomial distribution $X \sim  nb(r,p)$ with mean $\mu=r/p$ and variance $\sigma^2 = r(1-p)/p^2$. The probability of $X=x$ trials prior to $r$ successes is 

$$P(X=x) = {{x - 1} \choose {r - 1}} p^r (1-p)^{x-r}.$$

*An oil company has a p = 0.20 chance of striking oil when drilling a well.  What is the probability the company drills x = 7 wells to strike oil r = 3 times?*

$$P(X=7) = {{7 - 1} \choose {3 - 1}} (0.2)^3 (1-0.2)^{(7-3)} = 0.049.$$

Function `dnbinom()` calculates the negative-binomial probability.  Parameter `x` equals the number of failures, $x - r$.

```{r}
dnbinom(x = 4, size = 3, prob = 0.2)
```

Here is a simulation of n = 10,000 random samples.  `rnbinom()` generates a random sample of numbers from the negative-binomial distribution.

```{r message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
data.frame(cnt = rnbinom(n = 10000, size = 3, prob = 0.2)) %>%
  count(cnt) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         X_eq_x = cnt == 7-3,
         cnt = cnt + 3) %>%
  filter(cnt < 15) %>%
  ggplot(aes(x = as.factor(cnt), y = n, fill = X_eq_x, label = pct)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c(my_colors$grey, my_colors$red)) +
  geom_label(aes(label = round(pct, 2)), size = 3, alpha = .6, check_overlap = TRUE) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Negative-Binomial Distribution", 
       subtitle = paste0("P(X=7) trials to reach 3 successes when p = 0.2 is ", round(dnbinom(4, 3, 0.2), 4), "."),
       x = "Trials",
       y = "Count",
       caption = "Simulation from n = 10,000 negative-binomial random samples.") 
```

### Geometric

If $X$ is the count of independent Bernoulli trials of success probability $p$ required to achieve the first successful trial, then $X$ is a random variable with a geometric distribution $X \sim  G(p)$ with mean $\mu=\frac{{n}}{{p}}$ and variance $\sigma^2 = \frac{{(1-p)}}{{p^2}}$ .  The probability of $X=n$ trials is 

$$f(X=n) = p(1-p)^{n-1}.$$

The probability of $X<=n$ trials is 

$$F(X=n) = 1 - (1-p)^n.$$ 

*Example.  A sports marketer randomly selects persons on the street until he encounters someone who attended a game last season. What is the probability the marketer encounters x = 3 people who did not attend a game before the first success if p = 0.20 of the population attended a game?*

Function `pgeom()` calculates the geometric distribution probability.

```{r}
dgeom(x = 3, prob = 0.20)
```

```{r message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
data.frame(cnt = rgeom(n = 10000, prob = 0.20)) %>%
  count(cnt) %>%
  top_n(n = 15, wt = n) %>%
  ungroup() %>%
  mutate(pct = round(n / sum(n), 2),
         X_eq_x = cnt == 3) %>%
  ggplot(aes(x = as.factor(cnt), y = n, fill = X_eq_x, label = pct)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c(my_colors$grey, my_colors$red)) +
  geom_text(size = 3) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of trials prior to first success",
       subtitle = paste("P(X = 3) | X ~ G(.2) = ", round(dgeom(3, .2), 4)),
       x = "Unsuccessful trials",
       y = "Count",
       caption = "simulation of n = 10,000 samples from geometric dist.") 
```

## Continuous Distributions

### Normal

Random variable $X$ is distributed $X \sim N(\mu, \sigma^2)$ if

$$f(X)=\frac{{1}}{{\sigma \sqrt{{2\pi}}}}e^{-.5(\frac{{x-\mu}}{{\sigma}})^2}$$.

#### Example {-}

*IQ scores are distributed $X \sim N(100, 16^2$. What is the probability a randomly selected person's IQ is <90?*

```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x = 90
# exact
pnorm(q = my_x, mean = my_mean, sd = my_sd, lower.tail = TRUE)
# simulated
mean(rnorm(n = 10000, mean = my_mean, sd = my_sd) <= my_x)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > 0 & x <= my_x, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Example
*IQ scores are distributed *$X \sim N(100, 16^2$*. What is the probability a randomly selected person's IQ is >140?*
```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x = 140
# exact
pnorm(q = my_x, mean = my_mean, sd = my_sd, lower.tail = FALSE)
# simulated
mean(rnorm(n = 10000, mean = my_mean, sd = my_sd) > my_x)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > my_x & x < 1000, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```

### Example
*IQ scores are distributed *$X \sim N(100, 16^2$*. What is the probability a randomly selected person's IQ is between 92 and 114?*
```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x_l = 92
my_x_h = 114
# exact
pnorm(q = my_x_h, mean = my_mean, sd = my_sd, lower.tail = TRUE) -
  pnorm(q = my_x_l, mean = my_mean, sd = my_sd, lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > my_x_l & x <= my_x_h, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Example
*Class scores are distributed *$X \sim N(70, 10^2$*. If the instructor wants to give A's to >=85th percentile and B's to 75th-85th percentile, what are the cutoffs?*
```{r message=FALSE, warning=FALSE}
my_mean = 70
my_sd = 10
my_pct_l = .75
my_pct_h = .85

qnorm(p = my_pct_l, mean = my_mean, sd = my_sd, lower.tail = TRUE)
qnorm(p = my_pct_h, mean = my_mean, sd = my_sd, lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1000 / 10, 
           prob = pnorm(q = 0:1000 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(prob > my_pct_l & prob <= my_pct_h, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<=x) = ['~.(my_pct_l)~','~.(my_pct_h)~'] when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Normal Approximation to Binomial

The CLT implies that certain distributions can be approximated by the normal distribution.  

The binomial distribution $X \sim B(n,p)$ is approximately normal with mean $\mu = n p$ and variance $\sigma^2=np(1-p)$.  The approximation is useful when the expected number of successes and failures is at least 5:  $np>=5$ and $n(1-p)>=5$.

### Example
*A measure requires p>=50% popular to pass.  A sample of n=1,000 yields x=460 approvals. What is the probability that the overall population approves, P(X)>0.5?*
```{r message=FALSE, warning=FALSE}
my_x = 460
my_p = 0.50
my_n = 1000

my_mean = my_p * my_n
my_sd = round(sqrt(my_n * my_p * (1 - my_p)), 1)

# Exact binomial
pbinom(q = my_x, size = my_n, prob = my_p, lower.tail = TRUE)

# Normal approximation
pnorm(q = my_x, mean = my_p * my_n, sd = sqrt(my_n * my_p * (1 - my_p)), lower.tail = TRUE)

library(dplyr)
library(ggplot2)
library(tidyr)

data.frame(x = 400:600, 
           Normal = pnorm(q = 400:600, 
                        mean = my_p * my_n, 
                        sd = sqrt(my_n * my_p * (1 - my_p)), 
                        lower.tail = TRUE),
           Binomial = pbinom(q = 400:600, 
                        size = my_n, 
                        prob = my_p, 
                        lower.tail = TRUE)) %>%
  gather(key = "Distribution", value = "cdf", c(-x)) %>%
  ggplot(aes(x = x, y = cdf, color = Distribution)) +
  geom_line() +
  labs(title = bquote('X~B(n='~.(my_n)~', p='~.(my_p)~'),  '~'X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = "Normal approximation to the binomial",
       x = "x",
       y = "Probability") 
```


The Poisson distribution $x~P(\lambda)$ is approximately normal with mean $\mu = \lambda$ and variance $\sigma^2 = \lambda$, for large values of $\lambda$.

### Example
*The annual number of earthquakes registering at least 2.5 on the Richter Scale and having an epicenter within 40 miles of downtown Memphis follows a Poisson distribution with mean *$\lambda=6.5$*. What is the probability that at least *$x>=9$* such earthquakes will strike next year?*
```{r message=FALSE, warning=FALSE}
my_x = 9
my_lambda = 6.5
my_sd = round(sqrt(my_lambda), 2)

# Exact Poisson
ppois(q = my_x - 1, lambda = my_lambda, lower.tail = FALSE)

# Normal approximation
pnorm(q = my_x - 0.5, mean = my_lambda, sd = my_sd, lower.tail = FALSE)

library(dplyr)
library(ggplot2)
library(tidyr)

data.frame(x = 0:200 / 10, 
           Normal = pnorm(q = 0:200 / 10, 
                        mean = my_lambda, 
                        sd = my_sd, 
                        lower.tail = TRUE),
           Poisson = ppois(q = 0:200 / 10, 
                        lambda = my_lambda, 
                        lower.tail = TRUE)) %>%
  gather(key = "Distribution", value = "cdf", c(-x)) %>%
  ggplot(aes(x = x, y = cdf, color = Distribution)) +
  geom_line() +
  labs(title = bquote('X~P('~lambda~'='~.(my_lambda)~'),  '~'X~N('~mu==.(my_lambda)~','~sigma^{2}==.(my_lambda)~')'),
       subtitle = "Normal approximation to the Poisson",
       x = "x",
       y = "Probability") 
```

### From Sample to Population

*Suppose a person's blood pressure typically measures 160?20 mm.  If one takes n=5 blood pressure readings, what is the probability the average will be <=150?*
```{r message=FALSE, warning=FALSE}
my_mu = 160
my_sigma = 20
my_n = 5
my_x = 150

my_se = round(my_sigma / sqrt(my_n), 1)

pnorm(q = my_x, mean = my_mu, sd = my_sigma / sqrt(my_n), lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 1000:2000 / 10, 
           prob = pnorm(q = 1000:2000 / 10, 
                        mean = my_mu, 
                        sd = my_sigma / sqrt(my_n), 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > 0 & x <= my_x, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mu)~','~sigma^{2}==.(my_se)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mu)~' and variance is'~sigma~'/sqrt(n)'~.(my_se)^{2}~'.'),
       x = "x",
       y = "Probability") 
```

```{r}
knitr::include_app("https://mpfoley73.shinyapps.io/shiny_dist/", 
  height = "600px")
```

