# Probability {#probability}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

Some _significant_ applications are demonstrated in this chapter.

## Normal
Random variable $X$ is distributed $X \sim N(\mu, \sigma^2)$ if $f(X)=\frac{{1}}{{\sigma \sqrt{{2\pi}}}}e^{-.5(\frac{{x-\mu}}{{\sigma}})^2}$.

R function `dnorm(x, mean, sd)` is the probability of `x` when the mean is `mean` and the standard deviation is `sd`.  R function `pnorm(q, mean, sd, lower.tail)` is the cumulative probability (`lower.tail = TRUE` for left tail, `lower.tail = FALSE` for right tail) of less than or equal to value `q`.  R function rnorm(n, mean, sd) returns `n` random numbers from the normal distribution `X~N(mu, sigma^2)`.  R function qnorm(p, mean, sd, lower.tail) is the value of `x` at the qth percentile (`lower.tail = TRUE`).

### Example
IQ scores are distributed $X \sim N(100, 16^2$. What is the probability a randomly selected person's IQ is <90?
```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x = 90
# exact
pnorm(q = my_x, mean = my_mean, sd = my_sd, lower.tail = TRUE)
# simulated
mean(rnorm(n = 10000, mean = my_mean, sd = my_sd) <= my_x)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > 0 & x <= my_x, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Example
*IQ scores are distributed *$X \sim N(100, 16^2$*. What is the probability a randomly selected person's IQ is >140?*
```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x = 140
# exact
pnorm(q = my_x, mean = my_mean, sd = my_sd, lower.tail = FALSE)
# simulated
mean(rnorm(n = 10000, mean = my_mean, sd = my_sd) > my_x)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > my_x & x < 1000, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```

### Example
*IQ scores are distributed *$X \sim N(100, 16^2$*. What is the probability a randomly selected person's IQ is between 92 and 114?*
```{r message=FALSE, warning=FALSE}
my_mean = 100
my_sd = 16
my_x_l = 92
my_x_h = 114
# exact
pnorm(q = my_x_h, mean = my_mean, sd = my_sd, lower.tail = TRUE) -
  pnorm(q = my_x_l, mean = my_mean, sd = my_sd, lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1500 / 10, 
           prob = pnorm(q = 0:1500 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > my_x_l & x <= my_x_h, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Example
*Class scores are distributed *$X \sim N(70, 10^2$*. If the instructor wants to give A's to >=85th percentile and B's to 75th-85th percentile, what are the cutoffs?*
```{r message=FALSE, warning=FALSE}
my_mean = 70
my_sd = 10
my_pct_l = .75
my_pct_h = .85

qnorm(p = my_pct_l, mean = my_mean, sd = my_sd, lower.tail = TRUE)
qnorm(p = my_pct_h, mean = my_mean, sd = my_sd, lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 0:1000 / 10, 
           prob = pnorm(q = 0:1000 / 10, 
                        mean = my_mean, 
                        sd = my_sd, 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(prob > my_pct_l & prob <= my_pct_h, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = bquote('P(X<=x) = ['~.(my_pct_l)~','~.(my_pct_h)~'] when mean is'~.(my_mean)~' and variance is'~.(my_sd)^{2}~'.'),
       x = "x",
       y = "Probability") 
```


### Normal Approximation to Binomial

The CLT implies that certain distributions can be approximated by the normal distribution.  

The binomial distribution $X \sim B(n,p)$ is approximately normal with mean $\mu = n p$ and variance $\sigma^2=np(1-p)$.  The approximation is useful when the expected number of successes and failures is at least 5:  $np>=5$ and $n(1-p)>=5$.

### Example
*A measure requires p>=50% popular to pass.  A sample of n=1,000 yields x=460 approvals. What is the probability that the overall population approves, P(X)>0.5?*
```{r message=FALSE, warning=FALSE}
my_x = 460
my_p = 0.50
my_n = 1000

my_mean = my_p * my_n
my_sd = round(sqrt(my_n * my_p * (1 - my_p)), 1)

# Exact binomial
pbinom(q = my_x, size = my_n, prob = my_p, lower.tail = TRUE)

# Normal approximation
pnorm(q = my_x, mean = my_p * my_n, sd = sqrt(my_n * my_p * (1 - my_p)), lower.tail = TRUE)

library(dplyr)
library(ggplot2)
library(tidyr)

data.frame(x = 400:600, 
           Normal = pnorm(q = 400:600, 
                        mean = my_p * my_n, 
                        sd = sqrt(my_n * my_p * (1 - my_p)), 
                        lower.tail = TRUE),
           Binomial = pbinom(q = 400:600, 
                        size = my_n, 
                        prob = my_p, 
                        lower.tail = TRUE)) %>%
  gather(key = "Distribution", value = "cdf", c(-x)) %>%
  ggplot(aes(x = x, y = cdf, color = Distribution)) +
  geom_line() +
  labs(title = bquote('X~B(n='~.(my_n)~', p='~.(my_p)~'),  '~'X~N('~mu==.(my_mean)~','~sigma^{2}==.(my_sd)^{2}~')'),
       subtitle = "Normal approximation to the binomial",
       x = "x",
       y = "Probability") 
```


The Poisson distribution $x~P(\lambda)$ is approximately normal with mean $\mu = \lambda$ and variance $\sigma^2 = \lambda$, for large values of $\lambda$.

### Example
*The annual number of earthquakes registering at least 2.5 on the Richter Scale and having an epicenter within 40 miles of downtown Memphis follows a Poisson distribution with mean *$\lambda=6.5$*. What is the probability that at least *$x>=9$* such earthquakes will strike next year?*
```{r message=FALSE, warning=FALSE}
my_x = 9
my_lambda = 6.5
my_sd = round(sqrt(my_lambda), 2)

# Exact Poisson
ppois(q = my_x - 1, lambda = my_lambda, lower.tail = FALSE)

# Normal approximation
pnorm(q = my_x - 0.5, mean = my_lambda, sd = my_sd, lower.tail = FALSE)

library(dplyr)
library(ggplot2)
library(tidyr)

data.frame(x = 0:200 / 10, 
           Normal = pnorm(q = 0:200 / 10, 
                        mean = my_lambda, 
                        sd = my_sd, 
                        lower.tail = TRUE),
           Poisson = ppois(q = 0:200 / 10, 
                        lambda = my_lambda, 
                        lower.tail = TRUE)) %>%
  gather(key = "Distribution", value = "cdf", c(-x)) %>%
  ggplot(aes(x = x, y = cdf, color = Distribution)) +
  geom_line() +
  labs(title = bquote('X~P('~lambda~'='~.(my_lambda)~'),  '~'X~N('~mu==.(my_lambda)~','~sigma^{2}==.(my_lambda)~')'),
       subtitle = "Normal approximation to the Poisson",
       x = "x",
       y = "Probability") 
```

### From Sample to Population

*Suppose a person's blood pressure typically measures 160?20 mm.  If one takes n=5 blood pressure readings, what is the probability the average will be <=150?*
```{r message=FALSE, warning=FALSE}
my_mu = 160
my_sigma = 20
my_n = 5
my_x = 150

my_se = round(my_sigma / sqrt(my_n), 1)

pnorm(q = my_x, mean = my_mu, sd = my_sigma / sqrt(my_n), lower.tail = TRUE)

library(dplyr)
library(ggplot2)

data.frame(x = 1000:2000 / 10, 
           prob = pnorm(q = 1000:2000 / 10, 
                        mean = my_mu, 
                        sd = my_sigma / sqrt(my_n), 
                        lower.tail = TRUE)) %>%
  mutate(cdf = ifelse(x > 0 & x <= my_x, prob, 0)) %>%
ggplot() +
  geom_line(aes(x = x, y = prob)) +
  geom_area(aes(x = x, y = cdf), alpha = 0.3) +
  labs(title = bquote('X~N('~mu==.(my_mu)~','~sigma^{2}==.(my_se)^{2}~')'),
       subtitle = bquote('P(X<='~.(my_x)~') when mean is'~.(my_mu)~' and variance is'~sigma~'/sqrt(n)'~.(my_se)^{2}~'.'),
       x = "x",
       y = "Probability") 
```

```{r}
knitr::include_app("https://mpfoley73.shinyapps.io/shiny_dist/", 
  height = "600px")
```

