```{r include=FALSE}
library(tidyverse)
library(mfstylr)
```

# Discrete Analysis {#discrete_analysis}


## Chi-Square Test

These notes rely on [PSU STAT 500](https://newonlinecourses.science.psu.edu/stat500/node/56/),  [Wikipedia](https://en.wikipedia.org/wiki/Chi-squared_test), and [Disha M](http://www.yourarticlelibrary.com/project-reports/chi-square-test/chi-square-test-meaning-applications-and-uses-statistics/92394).

The chi-square test compares observed categorical variable frequency counts $O$ with their expected values $E$.  The test statistic $X^2 = \sum (O - E)^2 / E$ is distributed $\chi^2$.  $H_0: O = E$ and $H_a$ is at least one pair of frequency counts differ. The chi-square test relies on the central limit theorem, so it is valid for independent, normally distributed samples, typically affirmed with at least 5 successes and failures in each cell.  There a small variations in the chi-square for its various applications.

* The **chi-square goodness-of-fit test** tests whether observed frequency counts $O_j$ of the $j \in (0, 1, \cdots k)$ levels of a single categorical variable differ from expected frequency counts $E_j$. $H_0$ is $O_j = E_j$.

* The **chi-square independence test** tests whether observed joint frequency counts $O_{ij}$ of the $i \in (0, 1, \cdots I)$ levels of categorical variable $Y$ and the $j \in (0, 1, \cdots J)$ levels of categorical variable $Z$ differ from expected frequency counts $E_{ij}$ under the *independence model* where $\pi_{ij} = \pi_{i+} \pi_{+j}$, the joint densities. $H_0$ is $O_{ij} = E_{ij}$.

* The **chi-square homogeneity test** tests whether frequency counts of the $R$ levels of a categorical variable are distributed identically across $C$ different populations.

## One-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/60/).

A one-way table is a frequency table for a single categorical variable.  You usually construct a one-way table to test whether the frequency counts differ from a hypothesized distribution using the chi-square goodness-of-fit test.  You may also simply want to construct a confidence interval around a proportion.

Here is an example.  A researcher crosses tall cut-leaf tomatoes with dwarf potato-leaf tomatoes, then classifies the ($n = 1,611$) offspring phenotypes.

```{r}
o <- c(926, 288, 293, 104)
cell_names <- c("tall cut-leaf", "tall potato-leaf", "dwarf cut-leaf", "dwarf potato-leaf")
names(o) <- cell_names
print(o)
```

The four phenotypes are expected to occur with relative frequencies 9:3:3:1.

```{r}
pi <- c(9, 3, 3, 1) / (9 + 3 + 3 + 1)
print(pi)
```

```{r}
e <- sum(o) * pi
names(e) <- cell_names
print(e)
```


```{r fig.height=4}
data.frame(O = o, E = e) %>%
  rownames_to_column(var = "i") %>%
  pivot_longer(cols = -i, values_to = "freq") %>%
  group_by(name) %>%
  mutate(pct = freq / sum(freq)) %>%
  ungroup() %>%
  ggplot(aes(x = i, y = freq, fill = name, 
             label = paste0(round(freq, 0), "\n", 
                            scales::percent(pct, accuracy = 0.1)))
         ) +
  geom_col(position = position_dodge()) +
  geom_text(position = position_dodge(width = 0.9), size = 2.8) +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Observed vs Expected", fill = "")
```

Do the observed phenotype counts conform to the expected proportions?  This is a goodness-of-fit question because you are comparing frequencies from a single categorical variable to a set of hypothesized frequencies.


### Chi-Square Goodness-of-Fit Test

The **chi-square goodness-of-fit test** tests whether observed frequency counts $O_j$ of the $J$ levels of a categorical variable differ from expected frequency counts $E_j$ in a sample. $H_0$ is $O_j = E_j$.

There are two possible test statistics for this test, Pearson $X^2$ and deviance $G^2$.  The sampling distributions of $X^2$ and $G^2$ approach the $\chi_{J-1}^2$ as the sample size $n \rightarrow \infty$.  It's a good idea to calculate both test statistics.

The Pearson goodness-of-fit statistic is

$$X^2 = \sum \frac{(O_j - E_j)^2}{E_j}$$

where $O_j = p_j n$ and $E_j = \pi_j n$.  There is a variation of the $X^2$ statistic that corrects for small cell counts by subtracting 0.5 from each cell, the Yates Continuity Correction.  

$$X^2 = \sum \frac{(O_j - E_j - 0.5)^2}{E_j}$$

The deviance statistic, aka likelihood-ratio chi-square test statistic, is 

$$G^2 = 2 \sum O_j \log \left[ \frac{O_j}{E_j} \right]$$

If the saturated model (the observed data represent the fit of the saturated model, the most complex model possible with the data) proportions $p_j$ equal equal the expected proportions $\pi_j$, $X^2$ and $G^2$ will equal zero.  Large values indicate the data do not agree well with the proposed model.

You can perform a chi-square test of significance with the $G^2$ and $X^2$ test statistics with $dof$ degrees of freedom (d.f.).  The chi-square test is reliable when at least 80% of $E_j >= 5$.  

```{r echo=FALSE}
x2 <- sum((o - e)^2 / e)
g2 <- 2 * sum(o * log(o / e))
dof <- length(o) - 1
```

Calculate $X^2$ as `x2 <- sum((o - e)^2 / e) = ` `r  x2` and the $G^2$ as `g2 <- 2 * sum(o * log(o / e)) = ` `r g2`. The degrees of freedom are `length(o) - 1 = ` `r dof`.  The chi-sq test p-values are nearly identical.

```{r}
pchisq(q = x2, df = dof, lower.tail = FALSE)
pchisq(q = g2, df = dof, lower.tail = FALSE)
```

`chisq.test()` performs the chi-square test of the Pearson test statistic.

```{r}
chisq.test(o, p = pi)
```

The p-values based on the $\chi^2$ distribution with 3 d.f. are about 0.69, so the test fails to reject the null hypothesis that the observed frequencies are consistent with the theory. The plot of the chi-squared distribution shows $X^2$  well outside the $\alpha = 0.05$ range of rejection.

```{r warning=FALSE, message=FALSE, fig.height=4}
alpha <- 0.05
dof <- length(e) - 1
lrr = -Inf
p_val <- pchisq(q = x2, df = length(o) - 1, lower.tail = FALSE)
urr = qchisq(p = alpha, df = dof, lower.tail = FALSE)
 data.frame(chi2 = seq(from = 0, to = 20, by = .1)) %>%
   mutate(density = dchisq(x = chi2, df = dof)) %>%
   mutate(rr = ifelse(chi2 < lrr | chi2 > urr, density, 0)) %>%
 ggplot() +
   geom_line(aes(x = chi2, y = density), color = mf_pal(12)(12)[12], size = 0.8) +
   geom_area(aes(x = chi2, y = rr), fill = mf_pal(12)(12)[2], alpha = 0.8) +
   geom_vline(aes(xintercept = x2), color = mf_pal(12)(12)[11], size = 0.8) +
   labs(title = bquote("Chi-Square Goodness-of-Fit Test"),
        subtitle = paste0("X^2=", round(x2,2), ", ",
                          "Critical value=", round(urr,2), ", ",
                          "p-value=", round(p_val,3), "."
                          ),
        x = "chisq",
        y = "Density") +
   theme(legend.position="none") +
   theme_mf()
```

If you reject $H_0$, you can inspect the residuals to learn which differences may have lead to rejecting the rejection.  $X^2$ and $G^2$ are sums of squared cell comparisons, or "residuals".  The expected value of a $\chi^2$ random variable is its d.f., $k - 1$, so the average residual size is $(k - 1) / k$. The typical residual should be within 2 $\sqrt{(k - 1) / k}$.

```{r}
e2_res <- sqrt((o - e)^2 / e)
g2_res <- sign(o - e) * sqrt(abs(2 * o * log(o / e)))
```

```{r fig.height=4}
data.frame(e2_res) %>%
  rownames_to_column() %>%
  # pivot_longer(cols = e2_res:g2_res) %>%
  ggplot(aes(x = rowname, y = e2_res)) +
  geom_point(size = 3, color = mf_pal(12)(12)[2], alpha = 0.8) +
  theme_mf() +
  labs(title = "X^2 Residuals by Cell", color = "", x = "", y = "")
```

If you want to test whether the data conform to a particular distribution instead of some set of theoretical values, the test is nearly the same except for an adjustment to the d.f.  Your first step is the estimate the distribution's parameter(s).  Then you perform the goodness of fit test, but with degrees of freedom reduced for each estimated parameter.

For example, suppose you sample $n = 100$ families and count the number of children.  The count of children should be a Poisson random variable, $J \sim  Pois(\lambda)$.  

```{r}
dat <- data.frame(j = 0:5, o = c(19, 26, 29, 13, 10, 3))
```

The ML estimate for $\lambda$ is

$$\hat{\lambda} = \frac{j_0 O_0 + j_1 O_1, + \cdots j_k O_k}{O}$$

```{r}
lambda_hat <- sum(dat$j * dat$o) / sum(dat$o)
print(lambda_hat)
```

The probabilities for each possible count are 

$$f(j; \lambda) = \frac{e^{-\hat{\lambda}} \hat{\lambda}^j}{j!}.$$

```{r}
f <- exp(-lambda_hat) * lambda_hat^dat$j / factorial(dat$j)
E <- f * sum(dat$o)
dat <- cbind(dat, e = E)
```

```{r fig.height=4}
dat %>%
  rename(pois = e) %>%
  pivot_longer(cols = -j, values_to = "freq") %>%
  group_by(name) %>%
  mutate(pct = freq / sum(freq)) %>%
  ungroup() %>%
  ggplot(aes(x = fct_inseq(as.factor(j)), y = freq, fill = name, 
             label = paste0(round(freq, 0), "\n", 
                            scales::percent(pct, accuracy = 0.1)))
         ) +
  geom_col(position = position_dodge()) +
  geom_text(position = position_dodge(width = 0.9), size = 2.8) +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Observed vs Expected", fill = "", x = "children in family")
```

Compare the expected values to the observed values with the $\chi^2$ goodness of fit test.  In this case, $df = 6 - 1 - 1$ because the estimated paramater $\lambda$ reduces d.f. by 1.

```{r}
(X2 <- sum((dat$o - dat$e)^2 / dat$e))
(dof <- nrow(dat) - 1 - 1)
pchisq(q = X2, df = dof)
```

Be careful of this adjustment to the d.f. because `chisq.test()` does not take this into account, and you cannot override the d.f..

```{r}
chisq.test(dat$o, p = dat$e / sum(dat$e))
```


### Proportion Test

A special case of the one-way table is the $2 \times 1$ table for a binomial random variable.  When you calculate a single proportion $p$, you can compare it to a hypothesized $\pi_0$, or create a confidence interval around the estimate.

Suppose a company claims to resolve at least 70% of maintenance requests within 24 hours. In a random sample of $n = 50$ repair requests, the company resolves $O_1 = 33$ ($p_1 = 66\%)$ within 24 hours. At a 5% level of significance, is the maintenance company’s claim valid?

```{r}
o <- c(33, 17)
n <- sum(o)
cell_names <- c("resolved", "not resolved")
names(o) <- cell_names
print(o)
```

The null hypothesis is that the maintenance company resolves $\pi_0 = 0.70$ of requests within 24 hours, $H_0: \pi = \pi_0$ with alternative hypothesis $H_a: \pi < \pi_0$. This is a left-tailed test with an $\alpha = 0.05$ level of significance.

```{r}
pi_0 <- 0.70
alpha <- 0.05
```

The sample is independently drawn without replacement from <10% of the population (by assumption) and there were >=5 successes, so you can use the Clopper-Pearson exact binomial test. *Clopper-Pearson inverts two single-tailed binomial tests at the desired alpha*.

```{r}
binom.test(x = o, p = pi_0, alternative = "less", conf.level = 1 - alpha)
```

There is insufficient evidence (p = 0.3161) to reject $H_0$ that true probability of success is less than 0.7.

```{r fig.height=3.5}
x <- c(0:50)
p_x <- dbinom(x = x, size = n, prob = pi_0)
observed <- factor(if_else(x == o[1], 1, 0))
data.frame(x, p_x, observed) %>%
  ggplot(aes(x = x, y = p_x, fill = observed)) +
  geom_col() +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Exact Binomial")
```

There were >=5 failures, >=30 observations, and the measured probability of success was within (.2,.80), so you can also use the Wald normal approximation method where $\pi = p \pm z_{\alpha/2} SE$ and $Z = (p - \pi_0) / SE$ where $SE = \sqrt{\pi_0 (1 - \pi_0) / n}$.

```{r}
p <- o[1] / sum(o)
se <- sqrt(pi_0 * (1 - pi_0) / sum(o))
z <- (p - pi_0) / se
pnorm(q = p, mean = pi_0, sd = se, lower.tail = TRUE)
```

Again, there is insufficient evidence (p = 0.2685) to reject $H_0$ that true probability of success is less than 0.7.  The 95% CI around the measured p = 0.66 is

```{r}
z_alpha <- qnorm(0.95, mean = p, sd = se, lower.tail = FALSE)
c(0, p + z_alpha * se)
```


## Two-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/69/).

A two-way frequency table is a frequency table for *two* categorical variables.  You usually construct a two-way table to test whether the frequency counts in one categorical variable differ from the other categorical variable using the chi-square independence test.  If there is a significant difference (i.e., the variables are related), then describe the relationship with an analysis of the residuals, calculations of measures of association (difference in proportions, relative risk, or odds ratio), and partition tests. 

Here are three case studies that illustrate the concepts.  The first is a simple 2x2 table.  The second is a 3x2 table that extends some of the concepts.  The third is a 2x4 table where one factor is *ordinal*.

**Study 1: "Vitamin C" 2x2 Table**.  A double blind study investigated whether vitamin C prevents common colds on a sample of *n* = 279 persons. This study has two categorical variables each with two levels, a 2x2 two way table.

```{r}
vitc_o <- matrix(
  c(31, 17, 109, 122), 
  ncol = 2, 
  dimnames = list(
    treat = c("Placebo", "VitaminC"), 
    resp = c("Cold", "NoCold")
  )
)
vitc_o %>% data.frame() %>% rownames_to_column(var = " ") %>%
janitor::adorn_totals(where = c("row", "col"))
```

**Study 2: "Smoking" 3x2 Table**.  An analysis classifies *n* = 5375 high school students by their smoking behavior and the smoking behavior of their parents.

```{r}
smoke_o <- matrix(
  c(400, 416, 188, 1380, 1823, 1168), 
  ncol = 2,
  dimnames = list(
    parents = c("Both", "One", "Neither"),
    student = c("Smoker", "Non-smoker"))
)
smoke_o %>% data.frame() %>% rownames_to_column(var = " ") %>%
janitor::adorn_totals(where = c("row", "col"))
```

**Study 3: "CHD" Ordinal Table**.  A study of classified *n* = 1329 patients by cholesterol level and whether they had been diagnosed with coronary heart disease (CHD).

```{r}
# tribble() is a little easier.
chd_o <- tribble(
  ~L_0_199, ~L_200_219, ~L_220_259, ~L_260p,
  12, 8, 31, 41,
  307, 246, 439, 245
) %>% as.matrix()
rownames(chd_o) <- c("CHD", "No CHD")
chd_o %>% data.frame() %>% rownames_to_column(var = " ") %>%
janitor::adorn_totals(where = c("row", "col"))
```


### Chi-Square Independence Test

The **chi-square independence test** tests whether observed joint frequency counts $O_{ij}$ differ from expected frequency counts $E_{ij}$ under the *independence model* (the model of independent explanatory variables, $\pi_{ij} = \pi_{i+} \pi_{+j}$. $H_0$ is $O_{ij} = E_{ij}$.

There are two possible test statistics for this test, Pearson $X^2$ (and the continuity adjusted $X^2$), and deviance $G^2$.  As $n \rightarrow \infty$ their sampling distributions approach $\chi_{df}^2$ with degrees of freedom (df) equal to the saturated model df $I \times J - 1$ minus the independence model df $(I - 1) + (J - 1)$, which you can algebraically solve for $df = (I - 1)(J - 1)$.

The Pearson goodness-of-fit statistic is

$$X^2 = \sum \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where $O_{ij}$ is the observed count, and $E_{ij}$ is the product of the row and column marginal probabilities.  For the Vitamin C study, $X^2$ is 

```{r}
vitc_e <- sum(vitc_o) * prop.table(vitc_o, 1) * prop.table(vitc_o, 2)
X2 <- sum((vitc_o - vitc_e)^2 / vitc_e)
print(X2)
```

and the deviance statistic is 

$$G^2 = 2 \sum_{ij} O_{ij} \log \left( \frac{O_{ij}}{E_{ij}} \right)$$

```{r}
G2 <- - 2 * sum(vitc_o * log(vitc_o / vitc_e))
print(G2)
```

$X^2$ and $G^2$ increase with the disagreement between the saturated model proportions $p_{ij}$ and the independence model proportions $\pi_{ij}$.

The degrees of freedom is

```{r}
vitc_dof <- (nrow(vitc_o) - 1) * (ncol(vitc_o) - 1)
print(vitc_dof)
```

The associated p-values are
```{r}
pchisq(q = G2, df = vitc_dof, lower.tail = FALSE)
pchisq(q = X2, df = vitc_dof, lower.tail = FALSE)
```

The `chisq.test()` function applies the Yates continuity correcton by default to correct for situations with small cell counts.  The Yates continuity correction  subtracts 0.5 from the $O_{ij} - E_{ij}$ differences.  Set `correct = FALSE` to suppress Yates.

```{r}
vitc_chisq_test <- chisq.test(vitc_o, correct = FALSE)
print(vitc_chisq_test)
```

The Yates correction yields more conservative p-values.

```{r}
chisq.test(vitc_o)
```

These p-values are evidence for rejecting the independence model.

Here is the chi-square test applied to the CHD data.  Recall this data set is 4x2, so the degrees of freedom are $(4-1)(2-1) = 3$.  The Yates continuity correction does not apply to data other than 2x2, so the `correct = c(TRUE, FALSE)` has no effect in `chisq.test()`.

```{r}
(chd_chisq_test <- chisq.test(chd_o))
```

The p-value is very low, so reject the null hypothesis of independence. This demonstrates that a relationship exists between cholesterol and CHD.  Now you should describe that relationship by evaluating the (i) residuals, (ii) measures of association, and (iii) partitioning chi-square.


### Residuals Analysis

If the chi-squared independence test rejects $H_0$ of identical frequency distributions, the next step is to identify *which* cells may be driving the lack of fit.  The Pearson residuals in the two-way table are 

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}$$

where $X^2 = \sum{r_{ij}}$.  The $r_{ij}$ values have a normal distribution with mean 0, but with *unequal* variances.  The *standardized* Pearson residual for a two-way table is 

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}(1 - p_{i+})(1 - p_{+j})}}$$

and the $r_{ij}$ values *do* have a $\sim N(0, 1)$ distribution.  $r_{ij}^2 > 4$ is a sign of lack of fit. The `chissq.test()` object includes `residuals` that match the manual calculation.

```{r}
(vitc_o - vitc_e) / sqrt(vitc_e)
vitc_chisq_test$residuals
```

It also includes `stdres` that match the manual standardized calculation.  (*well, no it doesn't, but I don't know what my mistake is.*)

```{r}
(vitc_e - vitc_o) / 
  sqrt(vitc_e * (1 - prop.table(vitc_o, margin = 1)) 
              * (1 - prop.table(vitc_o, margin = 2))
  )
vitc_chisq_test$stdres
```

Here are the squared Pearson residuals for the CHD data.  The squared Pearson residuals for CHD 0-199, 200-219, and 260+ are greater than 4, and seem to be driving the lack of independence.

```{r}
chd_chisq_test$residuals^2
```


### Difference in Proportions

The difference in proportions measure is the difference in the probabilities of characteristic $Z$ conditioned on two groups $Y = 1$ and $Y = 2$: $\delta = \pi_{1|1} - \pi_{1|2}$.  In social sciences and epidemiology $\pi_{1|1}$ and $\pi_{1|2}$ are sometimes referred to as "risk" values. The point estimate for $\delta$ is $r = p_{1|1} - p_{1|2}$.  

Under the normal approximation method, the sampling distribution of the difference in population proportions has a normal distribution centered at $d$ with variance $Var(\delta)$. The point estimate for $Var(\delta)$ is $Var(d)$.

$$Var(d) = \frac{p_{1|1} (1 - p_{1|1})}{n_{1+}} + \frac{p_{1|2} (1 - p_{1|2})}{n_{2+}}$$

In the vitamin C acid example, $\delta$ is the difference in the row conditional frequencies.

```{r}
p <- prop.table(vitc_o, margin = 1)
d <- p[2, 1] - p[1, 1]
print(d)
```

The variance is

```{r}
var_d <- (p[2, 1])*(1 - p[2, 1]) / sum(vitc_o[2, ]) +
  (p[1, 1])*(1 - p[1, 1]) / sum(vitc_o[1, ])
print(var_d)
```

The 95% CI is

```{r}
d + c(-1, 1) * qnorm(.975) * sqrt(var_d)
```

This is how `prop.test()` without the continuity correction calculates the confidence interval.

```{r}
(prop.test.result <- prop.test(vitc_o, correct = FALSE))
```

```{r fig.height=3.5}
lcl <- -round(prop.test.result$conf.int[2], 3)
ucl <- -round(prop.test.result$conf.int[1], 3)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = d, sd = sqrt(var_d))) %>%
  mutate(rr = ifelse(d_i < lcl | d_i > ucl, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  theme_mf() +
  labs(title = bquote("Difference in Proportions Confidence Interval"),
       subtitle = paste0(
         "d = ", round(d, 3) 
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none")

```


The normal approximation method applies when the central limit theorem conditions hold:

* the sample is independently drawn (random sampling without replacement from $n < 10\%$ of the population in observational studies, or random assignment in experiments), 
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group, 
* the sample sizes are both $>=30$, and 
* the probability of success for each group is not extreme, $(0.2, 0.8)$.   

Test $H_0: d = \delta_0$ for some hypothesized population $\delta$ (usually 0) with test statistic 

$$Z = \frac{d - \delta_0}{se_{d}}$$ 

where 

$$se_{d} = \sqrt{p (1 - p) \left( \frac{1}{n_{1+}} + \frac{1}{n_{2+}} \right)}$$ 

approximates $se_{\delta_0}$ where $p$ is the pooled proportion 

$$p = \frac{n_{11} + n_{21}}{n_{1+} + n_{2+}}.$$

```{r}
p_pool <- (vitc_o[1, 1] + vitc_o[2, 1]) / sum(vitc_o)
se_d <- sqrt(p_pool * (1 - p_pool) * (1 / sum(vitc_o[1, ]) + 1 / sum(vitc_o[2, ])))
z <- (d - 0) / se_d
pnorm(z) * 2
```

```{r message=FALSE, warning=FALSE, fig.height=3.5}
lrr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = TRUE)
urr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = FALSE)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = 0, sd = se_d)) %>%
  mutate(rr = ifelse(d_i < lrr | d_i > urr, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  geom_vline(aes(xintercept = 0), color = "black") +
  theme_mf() +
  labs(title = "Hypothesis Test of Difference in Proportions",
       subtitle = paste0(
         "d = ", round(d, 3), 
         " (Z = ", round(z, 2), 
         ", p = ", round(pnorm(z) * 2, 4), ")."
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none") 
```

The null hypothesis $H_0: \delta_0 = 0$ is equivalent to saying that two variables are independent, $\pi_{1|1} = \pi_{1|2}$, so you can also use the $\chi^2$ or $G^2$ test for independence in a 2 × 2.  That's what `prop.test()` is doing.  The square of the z-statistic is algebraically equal to $\chi^2$.  The two-sided test comparing $Z$ to a $N(0, 1)$ is identical to comparing $\chi^2$ to a chi-square distribution with df = 1.  Compare the $Z^2$ to the output from `prop.test()`.

```{r}
z^2
prop.test.result$statistic
```

The difference in proportions is easy to interpret, but when $Z = 1$ is a rare event, the individual probabilities $\pi_{1|1}$ and $\pi_{1|2}$ are both small and $\delta$ is nearly zero even when the effect is strong.

In the CHD study, two of the conditional probabilities of CHD within the four cholesterol groups are similar, 0-199 (0.038) and 200-219 (.031).

```{r}
round(prop.table(chd_o, margin = 2), 3)
```

Is the difference in these proportions statistically signficant?  You can test this with the difference in proportions test or a chisq test.

```{r}
prop.test(t(chd_o[, c(1:2)]))
chisq.test(t(chd_o[, c(1:2)]))
```

You could go on to try other pairwise tests to establish which levels differ from the others.


### Relative Risk

The relative risk measure is the ratio of the probabilities of characteristic $Z$ conditioned on two groups $Y = 1$ and $Y = 2$: $\rho = \pi_{1|1} / \pi_{1|2}$.  In social sciences and epidemiology $\rho$ is sometimes referred to as the "relative risk". The point estimate for $\rho$ is $r = p_{1|1} / p_{1|2}$.  

Because $\rho$ is non-negative, a normal approximation for $\log \rho$ has a less skewed distribution than $\rho$. The approximate variance of $\log \rho$ is

$$Var(\log \rho) = \frac{1 - \pi_{11}/\pi_{1+}}{n_{1+}\pi_{11}/\pi_{1+}} + \frac{1 - \pi_{21}/\pi_{2+}}{n_{2+}\pi_{21}/\pi_{2+}}$$

and is estimated by 

$$Var(\log r) = \left( \frac{1}{n_{11}} - \frac{1}{n_{1+}} \right) + \left( \frac{1}{n_{21}} - \frac{1}{n_{2+}} \right)$$

In the vitamin C acid example, $r$ is the ratio of the row conditional frequencies.

```{r}
vitc_prop <- prop.table(vitc_o, margin = 1)
vitc_risk <- vitc_prop[2, 1] / vitc_prop[1, 1]
print(vitc_risk)
```

The variance is

```{r}
vitc_risk_var <- 1 / vitc_o[1, 1] - 1 / sum(vitc_o[1, ]) + 
  1 / vitc_o[2, 1] - 1 / sum(vitc_o[2, ])
print(vitc_risk_var)
```

The 95% CI is

```{r}
exp(log(vitc_risk) + c(-1, 1) * qnorm(.975) * sqrt(vitc_risk_var))
```

Thus, at 0.05 level, you can reject the independence model. People taking vitamin C are half as likely to catch a cold.

In the CHD study, you could summarize the relationship between CHD and cholesterol level by a set of three relative risks using 0-199 as the baseline:

* 200–219 versus 0–199, 
* 220–259 versus 0–199, and 
* 260+ versus 0–199. 

```{r}
(chd_prop <- prop.table(chd_o, margin = 2))
(chd_risk <- chd_prop[1, ] / chd_prop[1, 1])
```


### Odds Ratio

The odds ratio is the most commonly used measure of association.  It is also a natural parameter for many of the log-linear and logistic models. The odds is the ratio of probabilities of "success" and "failure".  When conditioned on a variable, the odds ratio is 

$$\theta = \frac{\pi_{1|1} / \pi_{2|1}} {\pi_{1|2} / \pi_{2|2}}$$

and is estimated by the sample frequencies

 $$\hat{\theta} = \frac{n_{11} n_{22}} {n_{12} n_{21}}$$
 
The log-odds ratio has a better normal approximation than the odds ratio, so define the confidence interval on the log scale.

$$Var(\log \hat{\theta}) = \frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}$$

For the Vitamin C example, the odds of getting a cold after taking a placebo pill are $0.22 / 0.78 = 0.28$ and the odds of getting a cold after taking Vitamin C are $0.12 / 0.88 = 0.14$.

```{r}
vitc_odds <- vitc_prop[, 1] / vitc_prop[, 2]
print(vitc_odds)
```

The odds of getting a cold given vitamin C are $0.14 / 0.28 = 0.49$ times the odds of getting cold given a placebo.  

```{r}
vitc_theta_hat <- vitc_odds[2] / vitc_odds[1]
print(vitc_theta_hat)
```

with variance 

```{r}
(var_vitc_theta_hat <- sum(1 / vitc_o))
```

The 95% CI is

```{r}
z_alpha <- qnorm(p = 0.975)
exp(log(vitc_theta_hat) + c(-1, 1) * z_alpha * sqrt(var_vitc_theta_hat))
```

Keep in mind the following properties of odds ratios.

* You can convert an odds pack to probabilities by solving $\pi / (1 - \pi)$ for $\pi = odds / (1 + odds)$. 

* If two variables are independent, then the conditional probabilities $\pi_{1|1}$ and $\pi_{1|2}$ will be equal and therefore the odds ratio will equal 1.

* If $\pi_{1|1} > \pi_{1|2}$ then the odds ratio will be $1 < \theta < \infty$.

* If $\pi_{1|1} < \pi_{1|2}$ then the odds ratio will be $0 < \theta < 1$.

* the sample odds ratio will equal $0$ or $\infty$ if any $n_{ij} = 0$. If you have any empty cells add 1/2 to each cell count.


### Partitioning Chi-Square

Besides looking at the residuals or the measures of association, another way to describe the effects is to form a sequence of smaller tables by combining or collapsing rows and/or columns in a meaningful way. 

For the smoking study, you might ask whether a student is more likely to smoke if *either* parent smokes.  Collapse the first two rows (1 parent smokes, both parents smoke) and run the chi-squared test.

```{r}
smoke_clps_1 <- rbind(smoke_o[1, ] + smoke_o[2, ], smoke_o[3, ])
smoke_clps_1_theta <- (smoke_clps_1[1, 1] / smoke_clps_1[1, 2]) /
  (smoke_clps_1[2, 1] / smoke_clps_1[2, 2])
(smoke_clps_1_chisq <- chisq.test(smoke_clps_1))
```

The estimated odds a student smokes if at least one parent smokes is `r round(smoke_clps_1_theta, 2)` (X^2 = `r round(smoke_clps_1_chisq$statistic, 1)`, p = `r smoke_clps_1_chisq$p.value`.

Or you may ask, whether among students with at least one smoking parent, there is a difference between those with one smoking parent and those with two smoking parents. Answer this by running a chi-square test on the first two rows of the data table, discarding the row where neither parent smokes.

```{r}
smoke_clps_2_theta <- (smoke_o[1, 1] / smoke_o[1, 2]) /
  (smoke_o[2, 1] / smoke_o[2, 2])
(smoke_clps_2_chisq <- chisq.test(smoke_o[c(1:2), ]))
```

The estimated odds a student smokes if both parents smoke compared to one parent is `r round(smoke_clps_2_theta, 2)` (X^2 = `r round(smoke_clps_2_chisq$statistic, 1)`, p = `r smoke_clps_2_chisq$p.value`.

### Correlation

When classification is ordinal, there may exist a linear trend among the levels of the characteristics.  Measure the linear relationship with Pearson's correlation coefficient, or its nonparametric alternatives, Spearman's correlation coefficient and Kendall's tau.

In the CHD study, the four levels of cholesterol (0-199, 200-219, 220-259, and 260+) may be treated as ordinal data.  You can also treat the presence of heart disease as ordinal. The Pearson correlation, 

$$r = \frac{cov(X, Y)}{s_X s_Y}$$

is 

```{r}
#dim=dim(table) 
rbar <- sum(margin.table(chd_o, 1) * c(1, 2)) / sum(chd_o) 
rdif <- c(1, 2) - rbar 
cbar <- sum(margin.table(chd_o, 2) * c(1, 2, 3, 4)) / sum(chd_o) 
cdif <- c(1, 2, 3, 4) - cbar 
ssr <- sum(margin.table(chd_o, 1) * (rdif^2)) 
ssc <- sum(margin.table(chd_o, 2) * (cdif^2)) 
ssrc <- sum(t(chd_o * rdif) * cdif) 
pcor <- ssrc / (sqrt(ssr * ssc)) 
pcor 
M2 <- (sum(chd_o) - 1) * pcor^2
M2
```

## K-Way Tables

A *k*-way table has *k* independent variables.  Each cell in the *k*-way table has joint probability $\pi_{ij..k}$.  You can collapse a *k*-way table along a dimension to create a *marginal* table and the resulting joint probabilities in the marginal table are referred to as *marginal distributions*.  Alternatively, you can consider a single level of a dimension, creating a conditional distribution.

Here are two case studies to illustrate the concepts.  

**Study 1: DeathP**.  Data is collected for *n = 326* murder cases along three dimensions: defendant's color, victim's color, and whether or not the defendant received the death penalty. 

```{r}
deathp <- array(c(19, 132, 11, 52, 0, 9, 6, 97), dim=c(2,2,2))
dimnames(deathp) <- list(
  DeathPen = c("yes","no"),
  Defendant=c("white","black"),
  Victim=c("white","black")
)
ftable(deathp, row.vars = c("Defendant", "Victim"), col.vars = "DeathPen")
```

**Study 2: BoyS**.  Data is collected for *n = 800* boyscouts along three dimensions: socioeconomic status, scout status, and delinquency status.

```{r}
boys <- expand.grid(
  delinquent = c("no","yes"),
  scout = c("no", "yes"),
  SES = c("low", "med","high")
)
boys <- cbind(boys, count = c(169,42,43,11,132,20,104,14,59,2,196,8))

boys_ct <- xtabs(count ~ ., boys)

ftable(boys_ct, row.vars = c("SES", "scout"), col.vars = "delinquent")
```


### Odds Ratio

In the *DeathP* study, the *marginal odds ratio* is 1.18, meaning the odds of death penalty for a white defendant are 1.18 times as high as they are for a black defendant.

```{r}
deathp_m <- margin.table(deathp, margin = c(1, 2))
deathp_p <- prop.table(deathp_m, margin = 2)
deathp_odds <- deathp_p[1, ] / deathp_p[2, ]
deathp_or <- deathp_odds[1] / deathp_odds[2]
print(deathp_or)
```

The *conditional odds ratio* given the victim is white is 0.68, and 0.79 given the victim is black, meaning the odds of death penalty for a white defendant are 0.68 times as high as they are for a black defendent if the victim is white, and 0.79 times as high as they are for a black defendent if the victim is black.

```{r}
deathp_m <- margin.table(deathp[, , 1], margin = c(2, 1))
deathp_p <- prop.table(deathp_m, margin = 2)
deathp_odds <- deathp_p[1, ] / deathp_p[2, ]
deathp_or <- deathp_odds[1] / deathp_odds[2]
print(deathp_or)

deathp_m <- margin.table(deathp[, , 2], margin = c(2, 1)) + 0.5
deathp_p <- prop.table(deathp_m, margin = 2)
deathp_odds <- deathp_p[1, ] / deathp_p[2, ]
deathp_or <- deathp_odds[1] / deathp_odds[2]
print(deathp_or)
```

Notice above that the second margin table had a zero in one cell.  To get an odds ratio in this case, the convention is to add 0.5 to all cells.

Interesting that the marginal odds of a white defendent receiving the death penalty are >1, but the conditional odds of a white defendent receiving the death penalty given the race of the victim are <1.  *Simpson's paradox* is the phenomenon that a pair of variables can have marginal association  and partial (conditional) associations in opposite direction. Another way to think about this is that the nature and direction of association changes due to presence or absence of a third (possibly confounding) variable.


### Chi-Square Independence Test

There are several ways to think about "independence" when dealing with a *k*-way table.

* **Mutual independence**.  All variables are independent from each other.
* **Joint independence**.  Two variables are jointly independent of the third.
* **Marginal independence**. Two variables are independent if you ignore the third.
* **Conditional independence** Two variables are independent given the third.
* **Homogeneous associations** Conditional (partial) odds-ratios are not related on the value of the third.


Under the assumption that the model of independence is true, once you know the marginal probability values, we have enough information to estimate all unknown cell probabilities. Because each of the marginal probability vectors must add up to one, the number of free parameters in the model is (I − 1) + (J − 1) + (K −I ). This is exactly like the two-way table


#### Mutual Independence

The simplest model is that ALL variables are independent of one another (A, B, C). The joint probabilities are equal to the product of the marginal probabilities, $\pi_{ijk} = \pi_i \pi_j \pi_k$. In terms of odds ratios, the model (A, B, C) implies that the odds ratios in the marginal tables A × B, B × C, and A × C are equal to 1.

The **chi-square independence test** tests whether observed joint frequency counts $O_{ijk}$ differ from expected frequency counts $E_{ijk}$ under the *independence model* $\pi_{ijk} = \pi_{i+} \pi_{+j} \pi_{k+}$. $H_0$ is $O_{ijk} = E_{ijk}$.  This is completely analagous to the two-way table chi-square independence test.

For the *BoyS* study, $X^2$ and its associated p-value is 

```{r}
boys_e <- array(NA, dim(boys_ct))
for (i in 1:dim(boys_ct)[1]) {
  for (j in 1:dim(boys_ct)[2]) {
    for (k in 1:dim(boys_ct)[3]) {
      boys_e[i,j,k] <- (
        margin.table(boys_ct, 3)[k] *
        margin.table(boys_ct, 2)[j] * 
        margin.table(boys_ct, 1)[i]) /
        (sum(boys_ct))^2
    }
  }
}
boys_df <- (prod(dim(boys_ct)) - 1) - sum(dim(boys_ct) - 1)
boys_x2 <- sum((boys_ct - boys_e)^2 / boys_e)
print(boys_x2)
pchisq(boys_x2, df = boys_df, lower.tail = FALSE)
```

and the deviance statistic is 

```{r}
boys_g2 <- 2 * sum(boys_ct * log(boys_ct / boys_e))
print(boys_g2)
pchisq(boys_g2, df = boys_df, lower.tail = FALSE)
```

`chisq.test()` does not provide a test of the complete independence model for *k*-way tables.  Instead, you can conduct tests on all the embedded 2-way tables.  For the *BoyS* study, you can conduct a series of 2-way chi-sq tests.

SES*scout:

```{r}
boys_ses_scout <- margin.table(boys_ct, c(3, 2))
chisq.test(boys_ses_scout)
```

SES*delinquent

```{r}
boys_ses_delinquent <- margin.table(boys_ct, c(3, 1))
chisq.test(boys_ses_delinquent)
```

delinquent*scout

```{r}
boys_delinquent_scout <- margin.table(boys_ct, c(1, 2))
chisq.test(boys_delinquent_scout)
```

The odds ratio for scout delinquency vs non-scout delinquency is 0.54 with 95% CI (0.347, 0.845), so reject the null hypothesis that boy scout status and delinquent status are independent of one another (OR = 1), and thus that boy scout status, delinquent status, and socioeconomic status are not mutually independent.

```{r}
(boys_or <- vcd::oddsratio(boys_delinquent_scout, log = FALSE))
confint(boys_or)
```

#### Joint Independence

The joint independence model is that two variables are jointly independent of a third (AB, C). The joint probabilities are equal to the product of the AB marginal probability and the C marginal probability, $\pi_{ijk} = \pi_{ij} \pi_k$. 

The degrees of freedom in the associated chi-sq test are the number of free parameters in the saturated model minus the number of free parameters in the joint independence model. $df = (IJK-1) - ((IJ-1) + (K-1)) = (IJ-1)(K-1)$.

From the *BoyS* study, suppose juvenile delinquency D as a response variable, and the other two —  boy scout status B and socioeconomic status S—as explanatory variables or predictors. Therefore, it would be useful to test the null hypothesis that D is independent of B and S, or (D, BS).

```{r}
#### Table of deliquent*scout at different level of SES

temp

#### Test for Joint Independence of (D,BS)
#### creating 6x2 table, BS x D

SESscout_deliquent=ftable(temp, row.vars=c(3,2))
result=chisq.test(SESscout_deliquent)
result

#### Test for Marginal Independence (see above analysis of two-way tables)

#### Test for conditional independence
#### To get partial tables of DB for each level of S

temp[,,1]
chisq.test(temp[,,1], correct=FALSE)
temp[,,2]
chisq.test(temp[,,2], correct=FALSE)
temp[,,3]
chisq.test(temp[,,3], correct=FALSE)
X2=sum(chisq.test(temp[,,1], correct=FALSE)$statistic+chisq.test(temp[,,2], correct=FALSE)$statistic+chisq.test(temp[,,3], correct=FALSE)$statistic)
1-pchisq(X2,df=3)

#### Cochran-Mantel-Haenszel test

mantelhaen.test(temp)
mantelhaen.test(temp,correct=FALSE) 

#### Breslow-Day test
#### make sure to first source/run breslowday.test_.R

breslowday.test(temp)

################################################################
#### Log-linear Models
#### Let's see how we would do this via Log_linear models 
#### We will look at the details later in the course
#### test of conditional independence
#### via loglinear model, but the object has to be a table!

is.table(temp)  #### to check if table
temp<-as.table(temp) #### to save as table
temp.condind<-loglin(temp, list(c(1,3), c(2,3)), fit=TRUE, param=TRUE) #### fit the cond.indep. model
temp.condind
1-pchisq(temp.condind$lrt, temp.condind$df)

#### There is no way to do Breslow-Day stats in R; 
#### you need to write your own function or fit the homogeneous association model
#### to test for identity of odds-ratios

#### test of homogenous association model

temp.hom<-loglin(temp, list(c(1,3), c(2,3), c(1,2)), fit=TRUE, param=TRUE)
temp.hom
1-pchisq(temp.hom$lrt, temp.hom$df)


#### Here is how to do the same but with the glm() function in R
#### test of conditional independence
#### via loglinear mode, but using glm() funcion
#### the object now needs to be dataframe

temp.data<-as.data.frame(temp)
temp.data
temp.condind<-glm(temp.data$Freq~temp.data$scout*temp.data$SES+temp.data$deliquent*temp.data$SES, family=poisson())
summary(temp.condind)
fitted(temp.condind)

#### test of homogenous association model

temp.hom<-glm(temp.data$Freq~temp.data$scout*temp.data$SES+temp.data$deliquent*temp.data$SES+temp.data$scout*temp.data$deliquent, family=poisson())
summary(temp.hom)
fitted(temp.hom)
```



Example of Chi-Square Test of Homogeneity
A project studied whether attending physicians order more unnecessary blood transfusions than residents.  The categorical variable frequency of orders has 4 levels: frequently, occasionally, rarely, and never. 
```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(stats)

pop <- NULL
pop[1:49] <- 1
pop[50:120] <- 2
lev <- NULL
lev[c(1:2, 50:64)] <- 1
lev[c(3:5, 65:92)] <- 2
lev[c(6:36, 93:115)] <- 3
lev[c(37:49, 116:120)] <- 4
dat <- data.frame(pop, lev)
dat$pop <- factor(dat$pop, levels = c(1, 2), 
                  labels = c("attending", "resident"))
dat$lev <- factor(dat$lev, levels = c(1, 2, 3, 4), 
                  labels = c("frequently", "occasionally", "rarely", "never"))

df <- (2-1)*(4-1)
alpha <- 0.05


(test <- chisq.test(dat$lev, dat$pop))

# Graph of hypothesis test
lrr = -Inf
urr = qchisq(p = alpha, df = df, lower.tail = FALSE)
data.frame(xi = 0:400 / 10) %>%
  mutate(density = dchisq(x = xi, df = df)) %>%
  mutate(rr = ifelse(xi < lrr | xi > urr, density, 0)) %>%
ggplot() +
  geom_line(aes(x = xi, y = density), color = "black") +
  geom_area(aes(x = xi, y = rr), fill = "red", alpha = 0.3) +
  geom_vline(aes(xintercept = test$statistic), color = "blue") +
  labs(title = bquote("Chi-Square Test for Homogeneity"),
       subtitle = bquote("P-value ="~.(test$p.value)),
       x = "xi^2",
       y = "Density") +
  theme(legend.position="none")

```



