```{r include=FALSE}
library(tidyverse)
library(mfstylr)
```

# Discrete Variables {#discrete_variables}


## Chi-Square Test

These notes rely on [PSU STAT 500](https://newonlinecourses.science.psu.edu/stat500/node/56/),  [Wikipedia](https://en.wikipedia.org/wiki/Chi-squared_test), and [Disha M](http://www.yourarticlelibrary.com/project-reports/chi-square-test/chi-square-test-meaning-applications-and-uses-statistics/92394)].

The chi-square test compares observed categorical variable frequency counts $O$ with their expected values $E$.  The test statistic $X^2 = \sum (O - E)^2 / E$ is distributed $\chi^2$.  $H_0: O = E$ and $H_a$ is at least one pair of frequency counts differ. The chi-square test relies on the central limit theorem, so it is valid for independent, normally distributed samples, typically affirmed with at least 5 successes and failures in each cell.  There a small variations in the chi-square for its various applications.

* The **chi-square goodness-of-fit test** tests whether observed frequency counts $O_j$ of the $j \in (0, 1, \cdots k)$ levels of a single categorical variable differ from expected frequency counts $E_j$. $H_0$ is $O_j = E_j$.

* The **chi-square independence test** tests whether observed joint frequency counts $O_{ij}$ of the $i \in (0, 1, \cdots I)$ levels of categorical variable $Y$ and the $j \in (0, 1, \cdots J)$ levels of categorical variable $Z$ differ from expected frequency counts $E_{ij}$ under the *independence model* where $\pi_{ij} = \pi_{i+} \pi_{+j}$, the joint densities. $H_0$ is $O_{ij} = E_{ij}$.

* The **chi-square homogeneity test** tests whether frequency counts of the $R$ levels of a categorical variable are distributed identically across $C$ different populations.

## One-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/60/).

A one-way table is a frequency table for a single categorical variable.  You usually construct a one-way table to test whether the frequency counts differ from a hypothesized distribution using the chi-square goodness-of-fit test.  You may also simply want to construct a confidence interval around a proportion.

Here is an example.  A researcher crosses tall cut-leaf tomatoes with dwarf potato-leaf tomatoes, then classifies the ($n = 1,611$) offspring phenotypes.

```{r}
o <- c(926, 288, 293, 104)
cell_names <- c("tall cut-leaf", "tall potato-leaf", "dwarf cut-leaf", "dwarf potato-leaf")
names(o) <- cell_names
print(o)
```

The four phenotypes are expected to occur with relative frequencies 9:3:3:1.

```{r}
pi <- c(9, 3, 3, 1) / (9 + 3 + 3 + 1)
print(pi)
```

```{r}
e <- sum(o) * pi
names(e) <- cell_names
print(e)
```


```{r fig.height=4}
data.frame(O = o, E = e) %>%
  rownames_to_column(var = "i") %>%
  pivot_longer(cols = -i, values_to = "freq") %>%
  group_by(name) %>%
  mutate(pct = freq / sum(freq)) %>%
  ungroup() %>%
  ggplot(aes(x = i, y = freq, fill = name, 
             label = paste0(round(freq, 0), "\n", 
                            scales::percent(pct, accuracy = 0.1)))
         ) +
  geom_col(position = position_dodge()) +
  geom_text(position = position_dodge(width = 0.9), size = 2.8) +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Observed vs Expected", fill = "")
```

Do the observed phenotype counts conform to the expected proportions?  This is a goodness-of-fit question because you are comparing frequencies from a single categorical variable to a set of hypothesized frequencies.


### Chi-Square Goodness-of-Fit Test

The **chi-square goodness-of-fit test** tests whether observed frequency counts $O_j$ of the $J$ levels of a categorical variable differ from expected frequency counts $E_j$ in a sample. $H_0$ is $O_j = E_j$.

There are two possible test statistics for this test, Pearson $X^2$ and deviance $G^2$.  The sampling distributions of $X^2$ and $G^2$ approach the $\chi_{J-1}^2$ as the sample size $n \rightarrow \infty$.  It's a good idea to calculate both test statistics.

The Pearson goodness-of-fit statistic is

$$X^2 = \sum \frac{(O_j - E_j)^2}{E_j}$$

where $O_j = p_j n$ and $E_j = \pi_j n$.  The deviance statistic is 

$$G^2 = 2 \sum O_j \log \left[ \frac{O_j}{E_j} \right]$$

If the saturated model (the observed data represent the fit of the saturated model, the most complex model possible with the data) proportions $p_j$ equal equal the expected proportions $\pi_j$, $X^2$ and $G^2$ will equal zero.  Large values indicate the data do not agree well with the proposed model.

You can perform a chi-square test of significance with the $G^2$ and $X^2$ test statistics with $dof$ degrees of freedom (d.f.).  The chi-square test is reliable when at least 80% of $E_j >= 5$.  

```{r echo=FALSE}
x2 <- sum((o - e)^2 / e)
g2 <- 2 * sum(o * log(o / e))
dof <- length(o) - 1
```

Calculate $X^2$ as `x2 <- sum((o - e)^2 / e) = ` `r  x2` and the $G^2$ as `g2 <- 2 * sum(o * log(o / e)) = ` `r g2`. The degrees of freedom are `length(o) - 1 = ` `r dof`.  The chi-sq test p-values are nearly identical.

```{r}
pchisq(q = x2, df = dof, lower.tail = FALSE)
pchisq(q = g2, df = dof, lower.tail = FALSE)
```

`chisq.test()` performs the chi-square test of the Pearson test statistic.

```{r}
chisq.test(o, p = pi)
```

The p-values based on the $\chi^2$ distribution with 3 d.f. are about 0.69, so the test fails to reject the null hypothesis that the observed frequencies are consistent with the theory. The plot of the chi-squared distribution shows $X^2$  well outside the $\alpha = 0.05$ range of rejection.

```{r warning=FALSE, message=FALSE, fig.height=4}
alpha <- 0.05
dof <- length(e) - 1
lrr = -Inf
p_val <- pchisq(q = x2, df = length(o) - 1, lower.tail = FALSE)
urr = qchisq(p = alpha, df = dof, lower.tail = FALSE)
 data.frame(chi2 = seq(from = 0, to = 20, by = .1)) %>%
   mutate(density = dchisq(x = chi2, df = dof)) %>%
   mutate(rr = ifelse(chi2 < lrr | chi2 > urr, density, 0)) %>%
 ggplot() +
   geom_line(aes(x = chi2, y = density), color = mf_pal(12)(12)[12], size = 0.8) +
   geom_area(aes(x = chi2, y = rr), fill = mf_pal(12)(12)[2], alpha = 0.8) +
   geom_vline(aes(xintercept = x2), color = mf_pal(12)(12)[11], size = 0.8) +
   labs(title = bquote("Chi-Square Goodness-of-Fit Test"),
        subtitle = paste0("X^2=", round(x2,2), ", ",
                          "Critical value=", round(urr,2), ", ",
                          "p-value=", round(p_val,3), "."
                          ),
        x = "chisq",
        y = "Density") +
   theme(legend.position="none") +
   theme_mf()
```

If you reject $H_0$, you can inspect the residuals to learn which differences may have lead to rejecting the rejection.  $X^2$ and $G^2$ are sums of squared cell comparisons, or "residuals".  The expected value of a $\chi^2$ random variable is its d.f., $k - 1$, so the average residual size is $(k - 1) / k$. The typical residual should be within 2 $\sqrt{(k - 1) / k}$.

```{r}
e2_res <- sqrt((o - e)^2 / e)
g2_res <- sign(o - e) * sqrt(abs(2 * o * log(o / e)))
```

```{r fig.height=4}
data.frame(e2_res) %>%
  rownames_to_column() %>%
  # pivot_longer(cols = e2_res:g2_res) %>%
  ggplot(aes(x = rowname, y = e2_res)) +
  geom_point(size = 3, color = mf_pal(12)(12)[2], alpha = 0.8) +
  theme_mf() +
  labs(title = "X^2 Residuals by Cell", color = "", x = "", y = "")
```

If you want to test whether the data conform to a particular distribution instead of some set of theoretical values, the test is nearly the same except for an adjustment to the d.f.  Your first step is the estimate the distribution's parameter(s).  Then you perform the goodness of fit test, but with degrees of freedom reduced for each estimated parameter.

For example, suppose you sample $n = 100$ families and count the number of children.  The count of children should be a Poisson random variable, $J \sim  Pois(\lambda)$.  

```{r}
dat <- data.frame(j = 0:5, o = c(19, 26, 29, 13, 10, 3))
```

The ML estimate for $\lambda$ is

$$\hat{\lambda} = \frac{j_0 O_0 + j_1 O_1, + \cdots j_k O_k}{O}$$

```{r}
lambda_hat <- sum(dat$j * dat$o) / sum(dat$o)
print(lambda_hat)
```

The probabilities for each possible count are 

$$f(j; \lambda) = \frac{e^{-\hat{\lambda}} \hat{\lambda}^j}{j!}.$$

```{r}
f <- exp(-lambda_hat) * lambda_hat^dat$j / factorial(dat$j)
E <- f * sum(dat$o)
dat <- cbind(dat, e = E)
```

```{r fig.height=4}
dat %>%
  rename(pois = e) %>%
  pivot_longer(cols = -j, values_to = "freq") %>%
  group_by(name) %>%
  mutate(pct = freq / sum(freq)) %>%
  ungroup() %>%
  ggplot(aes(x = fct_inseq(as.factor(j)), y = freq, fill = name, 
             label = paste0(round(freq, 0), "\n", 
                            scales::percent(pct, accuracy = 0.1)))
         ) +
  geom_col(position = position_dodge()) +
  geom_text(position = position_dodge(width = 0.9), size = 2.8) +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Observed vs Expected", fill = "", x = "children in family")
```

Compare the expected values to the observed values with the $\chi^2$ goodness of fit test.  In this case, $df = 6 - 1 - 1$ because the estimated paramater $\lambda$ reduces d.f. by 1.

```{r}
(X2 <- sum((dat$o - dat$e)^2 / dat$e))
(dof <- nrow(dat) - 1 - 1)
pchisq(q = X2, df = dof)
```

Be careful of this adjustment to the d.f. because `chisq.test()` does not take this into account, and you cannot override the d.f..

```{r}
chisq.test(dat$o, p = dat$e / sum(dat$e))
```


### Proportion Test

A special case of the one-way table is the $2 \times 1$ table for a binomial random variable.  When you calculate a single proportion $p$, you can compare it to a hypothesized $\pi_0$, or create a confidence interval around the estimate.

Suppose a company claims to resolve at least 70% of maintenance requests within 24 hours. In a random sample of $n = 50$ repair requests, the company resolves $O_1 = 33$ ($p_1 = 66\%)$ within 24 hours. At a 5% level of significance, is the maintenance company’s claim valid?

```{r}
o <- c(33, 17)
n <- sum(o)
cell_names <- c("resolved", "not resolved")
names(o) <- cell_names
print(o)
```

The null hypothesis is that the maintenance company resolves $\pi_0 = 0.70$ of requests within 24 hours, $H_0: \pi = \pi_0$ with alternative hypothesis $H_a: \pi < \pi_0$. This is a left-tailed test with an $\alpha = 0.05$ level of significance.

```{r}
pi_0 <- 0.70
alpha <- 0.05
```

The sample is independently drawn without replacement from <10% of the population (by assumption) and there were >=5 successes, so you can use the Clopper-Pearson exact binomial test. *Clopper-Pearson inverts two single-tailed binomial tests at the desired alpha*.

```{r}
binom.test(x = o, p = pi_0, alternative = "less", conf.level = 1 - alpha)
```

There is insufficient evidence (p = 0.3161) to reject $H_0$ that true probability of success is less than 0.7.

```{r fig.height=3.5}
x <- c(0:50)
p_x <- dbinom(x = x, size = n, prob = pi_0)
observed <- factor(if_else(x == o[1], 1, 0))
data.frame(x, p_x, observed) %>%
  ggplot(aes(x = x, y = p_x, fill = observed)) +
  geom_col() +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Exact Binomial")
```

There were >=5 failures, >=30 observations, and the measured probability of success was within (.2,.80), so you can also use the Wald normal approximation method where $\pi = p \pm z_{\alpha/2} SE$ and $Z = (p - \pi_0) / SE$ where $SE = \sqrt{\pi_0 (1 - \pi_0) / n}$.

```{r}
p <- o[1] / sum(o)
se <- sqrt(pi_0 * (1 - pi_0) / sum(o))
z <- (p - pi_0) / se
pnorm(q = p, mean = pi_0, sd = se, lower.tail = TRUE)
```

Again, there is insufficient evidence (p = 0.2685) to reject $H_0$ that true probability of success is less than 0.7.  The 95% CI around the measured p = 0.66 is

```{r}
z_alpha <- qnorm(0.95, mean = p, sd = se, lower.tail = FALSE)
c(0, p + z_alpha * se)
```


## Two-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/69/).

A two-way frequency table is a frequency table for *two* categorical variables.  You usually construct a two-way table to test whether the frequency counts in one categorical variable differ from the other categorical variable using the chi-square independence test.  If there is a significant difference (i.e., the variables are related), then describe the relationship with an analysis of the residuals, calculations of measures of association (difference in proportions, relative risk, or odds ratio), and partition tests. 

Here is an example.  A double blind study investigated the therapeutic value of vitamin C $(trt \in [Placebo, VitaminC])$ for treating common colds $(result \in [Cold, No Cold])$ on a sample of $n = 279$ persons. There are two discrete variables each with two levels, hence a two way table.

```{r}
vitc_dat <- matrix(
  c(31, 17, 109, 122), 
  ncol = 2, 
  dimnames = list(
    treat = c("Placebo", "VitaminC"), 
    resp = c("Cold", "NoCold")
  )
)
print(vitc_dat)
prop.table(vitc_dat)
prop.table(vitc_dat, margin = 1)  # row pct
prop.table(vitc_dat, margin = 2)  # col pct
```

Do the observed frequency counts differ from the expected frequency counts under the independence model?  (The *independence model* is a subset of the *saturated model* where the two explanatory variables are independent.)


### Chi-Square Independence Test

The **chi-square independence test** tests whether observed joint frequency counts $O_{ij}$ of the $i \in (0, 1, \cdots I)$ levels of categorical variable $Y$ and the $j \in (0, 1, \cdots J)$ levels of categorical variable $Z$ differ from expected frequency counts $E_{ij}$ under the *independence model* where $\pi_{ij} = \pi_{i+} \pi_{+j}$. $H_0$ is $O_{ij} = E_{ij}$.

There are two test statistics for this test, Pearson $\chi^2$ and deviance $G^2$.  The sampling distribution of each test statistic approach the chi-square distribution as $n$ becomes large.  It's a good idea to calculate both test statistics.

The Pearson goodness-of-fit statistic is

$$\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where $O_{ij}$ is the observed count, and $E_{ij}$ equals the marginal probability of row $i$ multiplied by the marginal probability of column $j$.  

```{r}
E <- sum(vitc_dat) * prop.table(vitc_dat, 1) * prop.table(vitc_dat, 2)
X2 <- sum((vitc_dat - E)^2 / E)
print(X2)
```

The deviance statistic is 

$$G^2 = 2 \sum_{ij} O_{ij} \log \left( \frac{O_{ij}}{E_{ij}} \right)$$

```{r}
G2 <- - 2 * sum(vitc_dat * log(vitc_dat / E))
print(G2)
```

If the saturated model proportions $p_{ij}$ equal equal the independence model proportions $\pi_{ij}$, $\chi^2$ and $G^2$ will equal zero.  Large values indicate the data do not agree well with the proposed model.

As $n \rightarrow \infty$, $G^2$ and $\chi^2$ approach $\chi_{k-1}^2$ where $k$ is the number of n-way table cells.  You can therefore run a chi-square test of significance using the $G^2$ and $\chi^2$ test statistics.  The degrees of freedom for $E^2$ and $G^2$ equals the degrees of freedom for the saturated model, $I \times J - 1$, minus the degrees of freedom from the independence model, $(I - 1) + (J - 1)$, which you can algebraically solve for $(I - 1)(J - 1)$.

```{r}
dof <- (nrow(vitc_dat) - 1) * (ncol(vitc_dat) - 1)
print(dof)
```

The associated p-values are
```{r}
pchisq(q = G2, df = dof, lower.tail = FALSE)
pchisq(q = X2, df = dof, lower.tail = FALSE)
```

The `chisq.test()` function applies the Yates's continuity correcton by default to correct for situations with small cell counts by subtracting 0.5 from the observed - expected differences.

```{r}
chisq.test(vitc_dat, correct = FALSE)
```

The correction yields more conservative p-values.

```{r}
vitc_chisq_test <- chisq.test(vitc_dat)
print(vitc_chisq_test)
```

The p-values indicate strong evidence for rejecting the independence model.


### Residuals Analysis

The residuals in a Chi-squared test indicate *which* cells drive the lack of fit.  The Pearson residuals in a two-way table are 

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}$$

and the $r_{ij}$ values have a normal distribution with mean 0, but with *unequal* variances.  The *standardized* Pearson residual for a two-way table is 

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}(1 - p_{i+})(1 - p_{+j})}}$$

and the $r_{ij}$ values *do* have a $\sim N(0, 1)$ distribution.  The `chissq.test()` object includes `residuals` that match the manual calculation.

```{r}
(r <- (vitc_dat - E) / sqrt(E))
vitc_chisq_test$residuals
```

It alco includes `stdres` that match the manual standardized calculation.

```{r}
p <- prop.table(vitc_dat, margin = 1)
(r_std <- (vitc_dat - E) / 
    sqrt(E * 
           (1 - prop.table(vitc_dat, margin = 1)[, 1]) * 
           (1 - prop.table(vitc_dat, margin = 2)[1, ])))
vitc_chisq_test$stdres
```



### Difference in Proportions

The difference in proportions measure is the difference in the relative frequency of a characteristic $Z$ between two groups $Y = 1$ and $Y = 2$: $\delta = \pi_{1|1} - \pi_{1|2}$.  In social sciences and epidemiology $\pi_{1|1}$ and $\pi_{1|2}$ are sometimes referred to as "risk" values. The point estimate for $\delta$ is $d = p_{1|1} - p_{1|2}$.  

Under the normal approximation method, the sampling distribution of the difference in population proportions has a normal distribution centered at $d$ with variance $Var(\delta)$. The point estimate for $Var(\delta)$ is $Var(d)$.

$$Var(d) = \frac{p_{1|1} (1 - p_{1|1})}{n_{1+}} + \frac{p_{1|2} (1 - p_{1|2})}{n_{2+}}$$

In the vitamin C acid example, $\delta$ is the difference in the row conditional frequencies.

```{r}
p <- prop.table(vitc_dat, margin = 1)
d <- p[2, 1] - p[1, 1]
print(d)
```

The variance is

```{r}
var_d <- (p[2, 1])*(1 - p[2, 1]) / sum(vitc_dat[2, ]) +
  (p[1, 1])*(1 - p[1, 1]) / sum(vitc_dat[1, ])
```

The 95% CI is

```{r}
d + c(-1, 1) * qnorm(.975) * sqrt(var_d)
```

This is how `prop.test()` without the continuity correction calculates the confidence interval.

```{r}
(prop.test.result <- prop.test(vitc_dat, correct = FALSE))
```

```{r fig.height=3.5}
lcl <- -round(prop.test.result$conf.int[2], 3)
ucl <- -round(prop.test.result$conf.int[1], 3)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = d, sd = sqrt(var_d))) %>%
  mutate(rr = ifelse(d_i < lcl | d_i > ucl, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  theme_mf() +
  labs(title = bquote("Difference in Proportions Confidence Interval"),
       subtitle = paste0(
         "d = ", round(d, 3) 
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none")

```


The normal approximation method applies when the central limit theorem conditions hold:

* the sample is independently drawn (random sampling without replacement from $n < 10\%$ of the population in observational studies, or random assignment in experiments), 
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group, 
* the sample sizes are both $>=30$, and 
* the probability of success for each group is not extreme, $(0.2, 0.8)$.   

Test $H_0: d = \delta_0$ for some hypothesized population $\delta$ (usually 0) with test statistic 

$$Z = \frac{d - \delta_0}{se_{d}}$$ 

where 

$$se_{d} = \sqrt{p (1 - p) \left( \frac{1}{n_{1+}} + \frac{1}{n_{2+}} \right)}$$ 

approximates $se_{\delta_0}$ where $p$ is the pooled proportion 

$$p = \frac{n_{11} + n_{21}}{n_{1+} + n_{2+}}.$$

```{r}
p_pool <- (vitc_dat[1, 1] + vitc_dat[2, 1]) / sum(vitc_dat)
se_d <- sqrt(p_pool * (1 - p_pool) * (1 / sum(vitc_dat[1, ]) + 1 / sum(vitc_dat[2, ])))
z <- (d - 0) / se_d
pnorm(z) * 2
```

```{r message=FALSE, warning=FALSE, fig.height=3.5}
lrr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = TRUE)
urr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = FALSE)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = 0, sd = se_d)) %>%
  mutate(rr = ifelse(d_i < lrr | d_i > urr, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  geom_vline(aes(xintercept = 0), color = "black") +
  theme_mf() +
  labs(title = "Hypothesis Test of Difference in Proportions",
       subtitle = paste0(
         "d = ", round(d, 3), 
         " (Z = ", round(z, 2), 
         ", p = ", round(pnorm(z) * 2, 4), ")."
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none") 
```


The null hypothesis $H_0: \delta_0 = 0$ is equivalent to saying that two variables are independent, $\pi_{1|1} = \pi_{1|2}$, so you can also use the $\chi^2$ or $G^2$ test for independence in a 2 × 2.  That's what `prop.test()` is doing.  The square of the z-statistic is algebraically equal to $\chi^2$.  The two-sided test comparing $Z$ to a $N(0, 1)$ is identical to comparing $\chi^2$ to a chi-square distribution with df = 1.  Compare the $Z^2$ to the output from `prop.test()`.

```{r}
z^2
```

Even though difference of two proportions is very easy to interpret, one problem with using δ is that when Z = 1 is a rare event, the individual probabilities P(Z = 1 | Y = 1) and P(Z = 1 | Y = 2) are both small, i.e., close to zero. Absolute value of δ will be close to zero even when the effect is strong. In the following sections we study two other  common measures of association which compare the relative value of the proportions, rather than the absolute values.

One problem with evaluating $\delta$ is that when $Z = 1$ is a rare event, the individual probabilities $\pi_{1|1}$ and $\pi_{1|2}$ are both close to zero, and so $\delta$ is also close to zero even when the effect is strong.

More notes online at [PSU STAT 500](https://newonlinecourses.science.psu.edu/stat500/node/55/), [PSU STAT 504](https://online.stat.psu.edu/stat504/node/77/), and [Stat Trek](https://www.stattrek.com/estimation/difference-in-proportions.aspx?Tutorial=AP).]


### Relative Risk

The relative risk measure is the ratio of the probabilities of characteristic $Z$ conditioned on two groups $Y = 1$ and $Y = 2$: $\rho = \pi_{1|1} / \pi_{1|2}$.  In social sciences and epidemiology $\rho$ is sometimes referred to as the "relative risk". The point estimate for $\rho$ is $r = p_{1|1} / p_{1|2}$.  

Because $\rho$ is non-negative, a normal approximation for $\log \rho$ has a less skewed distribution than $\rho$. The approximate variance of $\log \rho$ is

$$Var(\log \rho) = \frac{1 - \pi_{11}/\pi_{1+}}{n_{1+}\pi_{11}/\pi_{1+}} + \frac{1 - \pi_{21}/\pi_{2+}}{n_{2+}\pi_{21}/\pi_{2+}}$$

which is estimated by 

$$Var(\log r) = \left( \frac{1}{n_{11}} - \frac{1}{n_{1+}} \right) + \left( \frac{1}{n_{21}} - \frac{1}{n_{2+}} \right)$$

In the vitamin C acid example, $r$ is the ratio of the row conditional frequencies.

```{r}
p <- prop.table(vitc_dat, margin = 1)
r <- p[2, 1] / p[1, 1]
print(r)
```

The variance is

```{r}
var_r <- 1 / vitc_dat[1, 1] - 1 / sum(vitc_dat[1, ]) + 
  1 / vitc_dat[2, 1] - 1 / sum(vitc_dat[2, ])
```

The 95% CI is

```{r}
exp(log(r) + c(-1, 1) * qnorm(.975) * sqrt(var_r))
```

Thus, at 0.05 level, you can reject the independence model. People taking vitamin C are half as likely to catch a cold.


### Odds Ratio

The odds ratio is the most commonly used measure of association.  It is also a natural parameter for many of the log-linear and logistic models. The odds is the ratio of probabilities of "success" and "failure".  When conditioned on a variable, the odds ratio is 

$$\theta = \frac{\pi_{1|1} / \pi_{2|1}} {\pi_{1|2} / \pi_{2|2}}$$

and is estimated by the sample frequencies

 $$\hat{\theta} = \frac{n_{11} n_{22}} {n_{12} n_{21}}$$
 
The log-odds ratio has a better normal approximation than the odds ratio, so define a confidence interval on the log scale.

$$Var(\log \hat{\theta}) = \frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}$$

For the Vitamin C example, the odds of getting a cold given a skier took vitamin C, are 

The odds of getting a cold after taking a placebo pill are $0.22 / 0.78 = 0.28$ and the odds of getting a cold after taking Vitamin C are $0.12 / 0.88 = 0.14$.

```{r}
odds <- p[, 1] / p[, 2]
```

The odds of getting a cold given vitamin C are $0.14 / 0.28 = 0.49$ times the odds of getting cold given a placebo.  

```{r}
theta_hat <- odds[2] / odds[1]
print(theta_hat)
```

with variance 

```{r}
var_theta_hat <- sum(1 / vitc_dat)
```

The 95% CI is

```{r}
z_alpha <- qnorm(p = 0.975)
exp(log(theta_hat) + c(-1, 1) * z_alpha * sqrt(var_theta_hat))
```

Keep in mind the following properties of odds ratios.

* You can convert an odds pack to probabilities by solving $\pi / (1 - \pi)$ for $\pi = odds / (1 + odds)$. 

* If two variables are independent, then the conditional probabilities $\pi_{1|1}$ and $\pi_{1|2}$ will be equal and therefore the odds ratio will equal 1.

* If $\pi_{1|1} > \pi_{1|2}$ then the odds ratio will be $1 < \theta < \infty$.

* If $\pi_{1|1} < \pi_{1|2}$ then the odds ratio will be $0 < \theta < 1$.

* the sample odds ratio will equal $0$ or $\infty$ if any $n_{ij} = 0$. If you have any empty cells add 1/2 to each cell count.


## Example of Chi-Square Test of Homogeneity
A project studied whether attending physicians order more unnecessary blood transfusions than residents.  The categorical variable frequency of orders has 4 levels: frequently, occasionally, rarely, and never. 
```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(stats)

pop <- NULL
pop[1:49] <- 1
pop[50:120] <- 2
lev <- NULL
lev[c(1:2, 50:64)] <- 1
lev[c(3:5, 65:92)] <- 2
lev[c(6:36, 93:115)] <- 3
lev[c(37:49, 116:120)] <- 4
dat <- data.frame(pop, lev)
dat$pop <- factor(dat$pop, levels = c(1, 2), 
                  labels = c("attending", "resident"))
dat$lev <- factor(dat$lev, levels = c(1, 2, 3, 4), 
                  labels = c("frequently", "occasionally", "rarely", "never"))

df <- (2-1)*(4-1)
alpha <- 0.05


(test <- chisq.test(dat$lev, dat$pop))

# Graph of hypothesis test
lrr = -Inf
urr = qchisq(p = alpha, df = df, lower.tail = FALSE)
data.frame(xi = 0:400 / 10) %>%
  mutate(density = dchisq(x = xi, df = df)) %>%
  mutate(rr = ifelse(xi < lrr | xi > urr, density, 0)) %>%
ggplot() +
  geom_line(aes(x = xi, y = density), color = "black") +
  geom_area(aes(x = xi, y = rr), fill = "red", alpha = 0.3) +
  geom_vline(aes(xintercept = test$statistic), color = "blue") +
  labs(title = bquote("Chi-Square Test for Homogeneity"),
       subtitle = bquote("P-value ="~.(test$p.value)),
       x = "xi^2",
       y = "Density") +
  theme(legend.position="none")

```



