```{r include=FALSE}
library(tidyverse)
library(mfstylr)
```

# Discrete Variables {#discrete_variables}


## One-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/60/).

A one-way frequency table is a frequency table for a single categorical variable.  You usually construct a one-way table to test whether the frequency counts differ from a hypothesized distribution using the chi-squared goodness-of-fit test.

Here is an example.  A researcher crosses tall cut-leaf tomatoes with dwarf potato-leaf tomatoes, then classifies the ($n = 1,611$) offspring phenotypes.

```{r}
o <- c(926, 288, 293, 104)
cell_names <- c("tall cut-leaf", "tall potato-leaf", "dwarf cut-leaf", "dwarf potato-leaf")
names(o) <- cell_names
print(o)
```

The four phenotypes are expected to occur with relative frequencies 9:3:3:1.

```{r}
pi <- c(9, 3, 3, 1) / (9 + 3 + 3 + 1)
print(pi)
```

Do the observed phonotype counts conform to the expected proportions?


### Chi-Squared Goodness-of-Fit Test

The **chi-squared goodness-of-fit test**, tests whether observed frequency counts $O_j$ of the $j \in (0, 1, \cdots k)$ levels of a categorical variable differ from expected frequency counts $E_j$. $H_0$ is $O_j = E_j$.

There are two test statistics for this test, Pearson $X^2$ and deviance $G^2$.  The sampling distribution of each test statistic approach the chi-squared distribution as $n$ becomes large.  It's a good idea to calculate both test statistics.

The Pearson goodness-of-fit statistic is

$$X^2 = \sum \frac{(O_j - E_j)^2}{E_j}$$

where $O_j = p_j n$ and $E_j = \pi_j n$.  The deviance statistic is 

$$G^2 = 2 \sum O_j \log \left( \frac{O_j}{E_j} \right)$$

If the saturated model (the observed data represent the fit of the saturated model, the most complex model possible with the data) proportions $p_j$ equal equal the expected proportions $\pi_j$, $X^2$ and $G^2$ will equal zero.  Large values indicate the data do not agree well with the proposed model.

As $n \rightarrow \infty$, $G^2$ and $X^2$ approach $\chi_{k-1}^2$ where $k$ is the number of n-way table cells.  You can therefore run a chi-square test of significance using the $G^2$ and $X^2$ test statistics.  The chi-square test provides reliable results when at least 80% of $E_j >= 5$.  


```{r}
e <- sum(o) * pi
names(e) <- cell_names
print(e)
```

```{r fig.width=6, fig.height=3.5}
data.frame(o, e) %>%
  rownames_to_column() %>%
  pivot_longer(cols = -rowname) %>%
  ggplot(aes(x = as.factor(rowname), y = value, fill = name)) +
  geom_col(position = position_dodge()) +
  geom_text(aes(label = round(value, 0)), position = position_dodge(width = 0.9)) +
  theme_mf() +
  scale_fill_mf() +
  labs(title = "Observed vs Expected", fill = "", x = "", y = "")
```

```{r echo=FALSE}
x2 <- sum((o - e)^2 / e)
g2 <- 2 * sum(o * log(o / e))
```

The $X^2$ statistic is `x2 <- sum((o - e)^2 / e) = ` `r  x2` and the $G^2$ statistic is `g2 <- 2 * sum(o * log(o / e)) = ` `r g2`, so nearly identical. The chi-sq test p-values are also nearly identical.

```{r}
pchisq(q = x2, df = length(o) - 1, lower.tail = FALSE)
pchisq(q = g2, df = length(o) - 1, lower.tail = FALSE)
```

You can also perform the chi-square test of the Pearson test statistic with `chisq.test()`.

```{r}
chisq.test(o, p = pi)
```

The p-values based on the $\chi^2$ distribution with 3 d.f. are about 0.69, so we fail to reject the null hypothesis - the data are consistent with the genetic theory. The $\chi^2$ value is well outside the $\alpha = 0.05$ level range of regression.

```{r warning=FALSE, message=FALSE, fig.height=4}
alpha <- 0.05
dof <- length(e) - 1
lrr = -Inf
p_val <- pchisq(q = x2, df = length(o) - 1, lower.tail = FALSE)
urr = qchisq(p = alpha, df = dof, lower.tail = FALSE)
 data.frame(chi2 = seq(from = 0, to = 20, by = .1)) %>%
   mutate(density = dchisq(x = chi2, df = dof)) %>%
   mutate(rr = ifelse(chi2 < lrr | chi2 > urr, density, 0)) %>%
 ggplot() +
   geom_line(aes(x = chi2, y = density), color = mf_pal(12)(12)[12], size = 0.8) +
   geom_area(aes(x = chi2, y = rr), fill = mf_pal(12)(12)[2], alpha = 0.8) +
   geom_vline(aes(xintercept = x2), color = mf_pal(12)(12)[11], size = 0.8) +
   labs(title = bquote("Chi-Squared Goodness-of-Fit Test"),
        subtitle = paste0("Chisq=", round(x2,2), ", ",
                          "Critical value=", round(urr,2), ", ",
                          "p-value=", round(p_val,3), "."
                          ),
        x = "chisq",
        y = "Density") +
   theme(legend.position="none") +
   theme_mf()
```

Inspect the residuals to learn which differences may lead to rejecting the null hypothesis.  The Pearson and deviance statistics are sums of individual cell comparisons which can be termed the squared "residuals", $X^2 = \sum r_j^2$ and $G^2 = \sum G_j^2$.  The expected value of a $\chi^2$ random variable is its d.f., $k-1$, and the typical size of any $\chi^2$ value is be $(k - 1) / k$, so for the typical residual should be within 2 $\sqrt{(k - 1) / k}$.

```{r}
e2_res <- sqrt((o - e)^2 / e)
g2_res <- sign(o - e) * sqrt(abs(2 * o * log(o / e)))
```

```{r}
data.frame(e2_res, g2_res, obs = 1:length(e2_res)) %>%
  rownames_to_column() %>%
  pivot_longer(cols = e2_res:g2_res) %>%
  ggplot(aes(x = obs, y = value, color = name)) +
  geom_point() +
  scale_color_mf()
```


Here is a summary of the analysis.

```{r}
data.frame(o, e, e2_res, g2_res)
```

If you do not have a theoretical value, but instead want to test whether the data conform to a distribution, the test is nearly the same.  Your first step is the estimate the distribution parameter(s).  Then you can perform the goodness of fit test, but with degrees of freedom reduced for each estimated parameter.

For example, suppose sample $n = 100$ families and count the number of children $J$, and you want to test whether the counts $O$ follow a Poisson distribution, $X \sim  Pois(\lambda)$.  

```{r}
dat <- data.frame(j = 0:5, o = c(19, 26, 29, 13, 10, 3))
print(dat)
```

The ML estimate for $\lambda$ is

$$\hat{\lambda} = \frac{j_0 O_0 + j_1 O_1, + \cdots j_k O_k}{O}$$

```{r}
lambda_hat <- sum(dat$j * dat$o) / sum(dat$o)
print(lambda_hat)
```


Then the expected values are 

$$f(j; \lambda) = \frac{e^{-\hat{\lambda}} \hat{\lambda}^j}{j!}.$$

```{r}
E <- exp(-lambda_hat) * lambda_hat^dat$j / factorial(dat$j) * sum(dat$o)
dat <- cbind(dat, e = E)
print(dat)
```

Compare the expected values to the observed values with the $\chi^2$ goodness of fit test.

```{r}
X2 <- sum((dat$o - dat$e)^2 / dat$e)
pchisq(q = X2, df = nrow(dat) - 1 - 1)
chisq.test(dat$o, p = dat$e / sum(dat$e))
```


## Two-Way Tables

These notes rely on PSU STATS 504 [course notes](https://online.stat.psu.edu/stat504/node/69/).

A two-way frequency table is a frequency table for *two* categorical variables.  You usually construct a two-way table to test whether the frequency counts in one categorical variable differ from the other categorical variable using the chi-squared test of independence.  If there is a significant difference, then you characterize the effect sizes with measures of association (difference in proportions, relative risk, or odds ratio). 

Here is an example.  A double blind study investigated the therapeutic value of vitamin C $(trt \in [Placebo, VitaminC])$ for treating common colds $(result \in [Cold, No Cold])$ on a sample of $n = 279$ persons. There are two discrete variables each with two levels, hence a two way table.

```{r}
vitc_dat <- matrix(
  c(31, 17, 109, 122), 
  ncol = 2, 
  dimnames = list(
    treat = c("Placebo", "VitaminC"), 
    resp = c("Cold", "NoCold")
  )
)
print(vitc_dat)
prop.table(vitc_dat)
prop.table(vitc_dat, margin = 1)  # row pct
prop.table(vitc_dat, margin = 2)  # col pct
```

Do the observed frequency counts differ from the expected frequency counts under the independence model?  (The *independence model* is a subset of the *saturated model* where the two explanatory variables are independent.)


### Chi-Squared Test of Independence

The chi-squared test for independence tests whether there is a significant relationship between two categorical variables.  The null hypothesis is no relationship.

The Pearson goodness-of-fit statistic is

$$X^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where for independent variables $E_{ij}$ equals the marginal probability of row $i$ multiplied by the marginal probability of column $j$.
```{r}
E <- sum(vitc_dat) * prop.table(vitc_dat, 1) * prop.table(vitc_dat, 2)
X2 <- sum((vitc_dat - E)^2 / E)
print(X2)
```

The deviance statistic is 

$$G^2 = 2 \sum_{ij} O_{ij} \log \left( \frac{O_{ij}}{E_{ij}} \right)$$

```{r}
G2 <- - 2 * sum(vitc_dat * log(vitc_dat / E))
print(G2)
```

The degrees of freedom for $E^2$ and $G^2$ equals the degrees of freedom for the saturated model, $I \times J - 1$, minus the degrees of freedom from the independence model, $(I - 1) + (J - 1)$, which you can solve for $(I - 1)(J - 1)$.

```{r}
dof <- (nrow(vitc_dat) - 1) * (ncol(vitc_dat) - 1)
print(dof)
```

The associated p-values are
```{r}
pchisq(q = G2, df = dof, lower.tail = FALSE)
pchisq(q = X2, df = dof, lower.tail = FALSE)
```

The `chisq.test()` function applies the Yates's continuity correcton by default to correct for situations with small cell counts by subtracting 0.5 from the observed - expected differences.

```{r}
chisq.test(vitc_dat, correct = FALSE)
```

The correction yields more conservative p-values.

```{r}
chisq.test(vitc_dat)
```

The p-values indicate strong evidence for rejecting the independence model.


### Difference in Proportions

The difference in proportions measure is the difference in the relative frequency of a characteristic $Z$ between two groups $Y = 1$ and $Y = 2$: $\delta = \pi_{1|1} - \pi_{1|2}$.  In social sciences and epidemiology $\pi_{1|1}$ and $\pi_{1|2}$ are sometimes referred to as "risk" values. The point estimate for $\delta$ is $d = p_{1|1} - p_{1|2}$.  

Under the normal approximation method, the sampling distribution of the difference in population proportions has a normal distribution centered at $d$ with variance $Var(\delta)$. The point estimate for $Var(\delta)$ is $Var(d)$.

$$Var(d) = \frac{p_{1|1} (1 - p_{1|1})}{n_{1+}} + \frac{p_{1|2} (1 - p_{1|2})}{n_{2+}}$$

In the vitamin C acid example, $\delta$ is the difference in the row conditional frequencies.

```{r}
p <- prop.table(vitc_dat, margin = 1)
d <- p[2, 1] - p[1, 1]
print(d)
```

The variance is

```{r}
var_d <- (p[2, 1])*(1 - p[2, 1]) / sum(vitc_dat[2, ]) +
  (p[1, 1])*(1 - p[1, 1]) / sum(vitc_dat[1, ])
```

The 95% CI is

```{r}
d + c(-1, 1) * qnorm(.975) * sqrt(var_d)
```

This is how `prop.test()` without the continuity correction calculates the confidence interval.

```{r}
(prop.test.result <- prop.test(vitc_dat, correct = FALSE))
```

```{r fig.height=3.5}
lcl <- -round(prop.test.result$conf.int[2], 3)
ucl <- -round(prop.test.result$conf.int[1], 3)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = d, sd = sqrt(var_d))) %>%
  mutate(rr = ifelse(d_i < lcl | d_i > ucl, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  theme_mf() +
  labs(title = bquote("Difference in Proportions Confidence Interval"),
       subtitle = paste0(
         "d = ", round(d, 3) 
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none")

```


The normal approximation method applies when the central limit theorem conditions hold:

* the sample is independently drawn (random sampling without replacement from $n < 10\%$ of the population in observational studies, or random assignment in experiments), 
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group, 
* the sample sizes are both $>=30$, and 
* the probability of success for each group is not extreme, $(0.2, 0.8)$.   

Test $H_0: d = \delta_0$ for some hypothesized population $\delta$ (usually 0) with test statistic 

$$Z = \frac{d - \delta_0}{se_{d}}$$ 

where 

$$se_{d} = \sqrt{p (1 - p) \left( \frac{1}{n_{1+}} + \frac{1}{n_{2+}} \right)}$$ 

approximates $se_{\delta_0}$ where $p$ is the pooled proportion 

$$p = \frac{n_{11} + n_{21}}{n_{1+} + n_{2+}}.$$

```{r}
p_pool <- (vitc_dat[1, 1] + vitc_dat[2, 1]) / sum(vitc_dat)
se_d <- sqrt(p_pool * (1 - p_pool) * (1 / sum(vitc_dat[1, ]) + 1 / sum(vitc_dat[2, ])))
z <- (d - 0) / se_d
pnorm(z) * 2
```

```{r message=FALSE, warning=FALSE, fig.height=3.5}
lrr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = TRUE)
urr = qnorm(p = .05/2, mean = 0, sd = se_d, lower.tail = FALSE)
data.frame(d_i = -300:300 / 1000) %>%
  mutate(density = dnorm(x = d_i, mean = 0, sd = se_d)) %>%
  mutate(rr = ifelse(d_i < lrr | d_i > urr, density, 0)) %>%
ggplot() +
  geom_line(aes(x = d_i, y = density)) +
  geom_area(aes(x = d_i, y = rr, fill = mf_pal(12)(12)[2]), alpha = 0.8) +
  geom_vline(aes(xintercept = d), color = "blue") +
  geom_vline(aes(xintercept = 0), color = "black") +
  theme_mf() +
  labs(title = "Hypothesis Test of Difference in Proportions",
       subtitle = paste0(
         "d = ", round(d, 3), 
         " (Z = ", round(z, 2), 
         ", p = ", round(pnorm(z) * 2, 4), ")."
        ),
       x = "d",
       y = "Density") +
  theme(legend.position="none") 
```


The null hypothesis $H_0: \delta_0 = 0$ is equivalent to saying that two variables are independent, $\pi_{1|1} = \pi_{1|2}$, so you can also use the $\chi^2$ or $G^2$ test for independence in a 2 × 2.  That's what `prop.test()` is doing.  The square of the z-statistic is algebraically equal to $\chi^2$.  The two-sided test comparing $Z$ to a $N(0, 1)$ is identical to comparing $\chi^2$ to a chi-squared distribution with df =1.  Compare the $Z^2$ to the output from `prop.test()`.

```{r}
z^2
```

Even though difference of two proportions is very easy to interpret, one problem with using δ is that when Z = 1 is a rare event, the individual probabilities P(Z = 1 | Y = 1) and P(Z = 1 | Y = 2) are both small, i.e., close to zero. Absolute value of δ will be close to zero even when the effect is strong. In the following sections we study two other  common measures of association which compare the relative value of the proportions, rather than the absolute values.


One problem with evaluating $\delta$ is that when $Z = 1$ is a rare event, the individual probabilities $\pi_{1|1}$ and $\pi_{1|2}$ are both close to zero, and so $\delta$ is also close to zero even when the effect is strong.

More notes online at [PSU STAT 500](https://newonlinecourses.science.psu.edu/stat500/node/55/), [PSU STAT 504](https://online.stat.psu.edu/stat504/node/77/), and [Stat Trek](https://www.stattrek.com/estimation/difference-in-proportions.aspx?Tutorial=AP).]


### Relative Risk


### Odds Ratio