# Generalized Linear Models

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)  # for augment()
library(WVPlots)  # for GainCurvePlot()
library(caret)  # for RMSE()
library(readr)
library(mfstylr)
```


These notes are primarily from [PSU STAT 504](https://online.stat.psu.edu/stat504) which uses Alan Agresti's **Categorical Data Analysis** [@Agresti2013].  I also reviewed [PSU STAT 501](https://newonlinecourses.science.psu.edu/stat501/lesson/15), [DataCamp's Generalized Linar Models in R](https://www.datacamp.com/courses/generalized-linear-models-in-r), [DataCamp's Multiple and Logistic Regression](https://www.datacamp.com/courses/multiple-and-logistic-regression), and **Interpretable machine learning*"** [@Molner2020].

The linear regression model, $E(Y|X) = X \beta$, structured as $y_i = X_i \beta + \epsilon_i$ where $X_i \beta = \mu_i$, assumes the response is a linear function of the predictors and the residuals are independent random variables normally distributed with mean zero and constant variance, $\epsilon \sim N \left(0, \sigma^2 \right)$.  This implies that given some set of predictors, the response is normally distributed about its expected value, $y_i \sim N \left(\mu_i, \sigma^2 \right)$.  However, there are many situations where this assumption of normality fails.  Generalized linear models (GLMs) are a generalization of the linear regression model that addresses non-normal response distributions.

The response given a set of predictors will not have a normal distribution if its underlying data-generating process is binomial or multinomial (proportions), Poisson (counts), or exponential (time-to-event).  In these situations a regular linear regression can predict proportions outside [0, 100] or counts or times that are negative.  GLMs solve this problem by modeling *a function* of the expected value of $y$, $f(E(Y|X)) = X \beta$.  There are three components to a GLM: the *random component* is the probability distribution of the response variable (normal, binomial, Poisson, [etc](https://en.wikipedia.org/wiki/Exponential_family#Table_of_distributions).); the *systematic component* is the explanatory variables $X\beta$; and the *link function* $\eta$ specifies the link between random and systematic components, converting the response range to $[-\infty, +\infty]$.  

Linear regression is thus a special case of GLM where link function is the identity function, $f(E(Y|X)) = E(Y|X)$.  For a logistic regression, where the data generating process is binomial, the link function is 

$$f(E(Y|X)) = \ln \left( \frac{E(Y|X)}{1 - E(Y|X)} \right) = \ln \left( \frac{\pi}{1 - \pi} \right) = logit(\pi)$$

where $\pi$ is the event probability.  (As an aside, you have probably heard of the related "probit" regression. The probit regression link function is $f(E(Y|X)) = \Phi^{-1}(E(Y|X)) = \Phi^{-1}(\pi)$. The difference between the logistic and probit link function is theoretical, and [the practical significance is slight](https://www.theanalysisfactor.com/the-difference-between-logistic-and-probit-regression/).  You can probably safely ignore probit).  

For a Poisson regression, the link function is

$$f(E(Y|X)) = \ln (E(Y|X)) = \ln(\lambda)$$

where $\lambda$ is the expected event rate.  

For an exponential regression, the link function is 

$$f(E(Y|X) = -E(Y|X) = -\lambda$$

where $\lambda$ is the expected time to event.

GLM uses maximum likelihood estimation (MLE) rather than ordinary least squares (OLS) to estimate the parameters, and thus relies on large-sample approximations. 

In R, specify a GLM just like an linear model, but with the `glm()` function, specifying the distribution with the `family` parameter.

* `family = "gaussian"`: linear regression
* `family = "binomial"`: logistic regression
* `family = binomial(link = "probit")`: probit regression
* `family = "poisson"`: Poisson regression


## Logistic Regression

Logistic regression estimates the probability of a particular level of a categorical response variable given a set of predictors. The response levels can be binary, nominal (multiple categories), or ordinal (multiple levels).  

The **binary** logistic regression model is

$$y = logit(\pi) = \ln \left( \frac{\pi}{1 - \pi} \right) = X \beta$$

where $\pi$ is the event probability. The model predicts the *log odds* of the response variable.  The maximum likelihood estimator maximizes the likelihood function

$$L(\beta; y, X) = \prod_{i=1}^n \pi_i^{y_i}(1 - \pi_i)^{(1-y_i)} = \prod_{i=1}^n\frac{\exp(y_i X_i \beta)}{1 + \exp(X_i \beta)}.$$

There is no closed-form solution, so GLM estimates coefficients with interatively reweighted least squares.  

Here is a case study to illustrate the points.  Dataset `leuk` contains response variable `REMISS` indicating whether leukemia remission occurred (1|0) and several explanatory variables.

```{r include=FALSE}
leuk <- tribble(
  ~REMISS, ~CELL, ~SMEAR, ~INFIL, ~LI, ~BLAST, ~TEMP,
	1,	0.80,	0.83,	0.66,	1.9,	1.10,	1.00,
	1,	0.90,	0.36,	0.32,	1.4,	0.74,	0.99,
	0,	0.80,	0.88,	0.70,	0.8,	0.18,	0.98,
	0,	1.00,	0.87,	0.87,	0.7,	1.05,	0.99,
	1,	0.90,	0.75,	0.68,	1.3,	0.52,	0.98,
	0,	1.00,	0.65,	0.65,	0.6,	0.52,	0.98,
	1,	0.95,	0.97,	0.92,	1.0,	1.23,	0.99,
	0,	0.95,	0.87,	0.83,	1.9,	1.35,	1.02,
	0,	1.00,	0.45,	0.45,	0.8,	0.32,	1.00,
	0,	0.95,	0.36,	0.34,	0.5,	0.00,	1.04,
	0,	0.85,	0.39,	0.33,	0.7,	0.28,	0.99,
	0,	0.70,	0.76,	0.53,	1.2,	0.15,	0.98,
	0,	0.80,	0.46,	0.37,	0.4,	0.38,	1.01,
	0,	0.20,	0.39,	0.08,	0.8,	0.11,	0.99,
	0,	1.00,	0.90,	0.90,	1.1,	1.04,	0.99,
	1,	1.00,	0.84,	0.84,	1.9,	2.06,	1.02,
	0,	0.65,	0.42,	0.27,	0.5,	0.11,	1.01,
	0,	1.00,	0.75,	0.75,	1.0,	1.32,	1.00,
	0,	0.50,	0.44,	0.22,	0.6,	0.11,	0.99,
	1,	1.00,	0.63,	0.63,	1.1,	1.07,	0.99,
	0,	1.00,	0.33,	0.33,	0.4,	0.18,	1.01,
	0,	0.90,	0.93,	0.84,	0.6,	1.59,	1.02,
	1,	1.00,	0.58,	0.58,	1.0,	0.53,	1.00,
	0,	0.95,	0.32,	0.30,	1.6,	0.89,	0.99,
	1,	1.00,	0.60,	0.60,	1.7,	0.96,	0.99,
	1,	1.00,	0.69,	0.69,	0.9,	0.40,	0.99,
	0,	1.00,	0.73,	0.73,	0.7,	0.40,	0.99
)
```

```{r}
glimpse(leuk)
```

Fit a logistic regression.

```{r}
m1 <- glm(REMISS ~ ., family = binomial(link = logit), data = leuk)
summary(m1)
```

"z value" is the Wald z statistic,  $z = \hat{\beta} / SE(\hat{\beta})$, which if squared is the Wald chi-squared statistic, $z^2 = \left(\hat{\beta} / SE(\hat{\beta}) \right)^2$.  The p.value is the area to the right of $z^2$ in the $\chi_1^2$ density curve.

```{r}
m1 %>% tidy() %>% 
  mutate(
    z = estimate / std.error, 
    p_z2 = pchisq(z^2, df = 1, lower.tail = FALSE)
  ) %>%
  select(term, p.value, p_z2)
```

The "dispersion parameter" refers to overdispersion, a common issues with GLM.  For a logistic regression, the response variable should be distributed $y_i \sim Bin(n_i, \pi_i)$ with $\mu_i = n_i \pi_i$ and $\sigma^2 = \pi (1 - \pi)$.  Overdispersion means the data shows evidence of variance greater than $\sigma^2$.

"Fisher scoring" is a method for ML estimation. Logistic regression uses an iterative procedure to fit the model, so this section indicates whether the algorithm converged.

The log odds of having achieved remission when each predictor equals its mean value is $\hat{y} = -2.68$.  

```{r}
(pred <- predict(m1, newdata = data.frame(t(colMeans(leuk)))))
```

Log odds are not easy to interpet, but are convenient for [updating prior probabilities in Bayesian analyses](https://www.statisticshowto.datasciencecentral.com/log-odds/).  Exponentiate the log odds to get the more intuitive **odds**.

$$\exp (\hat{y}) = \exp (X \hat{\beta}) = \frac{\pi}{1 - \pi}.$$

The odds of having achieved remission when each predictor equals its mean value is $\exp(\hat{y}) = 0.068$.  

```{r}
exp(pred)
```

A common way to express this is with the inverse, 1 / 0.068 = 15:1.  The odds of having achieved remission when each predictor equals its mean value is "15 to 1".

```{r}
round(1/exp(pred), 0)
```

Or, solve for $\pi$ to get the **probability**.  

$$\pi = \frac{\exp (X \beta)}{1 + \exp (X \beta)}$$

The probability of having achieved remission when each predictor equals its mean value is $\pi = 0.064$.  The `predict()` function for a logistic model returns log-odds, but can also return $\pi$ by specifying parameter `type = "response"`.

```{r}
predict(m1, newdata = data.frame(t(colMeans(leuk))), type = "response")
```

Interpret the coefficient estimates using the **odds ratio**, the ratio of the odds before and after an increment to the predictors.  The odds ratio is how much the odds would be multiplied after a $X_1 - X_0$ unit increase in $X$.

$$\theta = \frac{\pi / (1 - \pi) |_{X = X_1}}{\pi / (1 - \pi) |_{X = X_0}} = \frac{\exp (X_1 \hat{\beta})}{\exp (X_0 \hat{\beta})} = \exp ((X_1-X_0) \hat{\beta}) = \exp (\delta \hat{\beta})$$

Increasing LI by .01 increases the odds of remission by a factor of $\exp(0.1 \cdot 4.36) = 1.547$ (from 15:1 to 23:1). 

```{r}
exp(.1 * m1$coefficients) %>% round(3)
```

You can calculate an odds ratio with specified increments using `oddsratio::or_glm()`.

```{r warning=FALSE}
oddsratio::or_glm(data = leuk, model = m1, 
  incr = list(
    CELL = 0.01, SMEAR = 0.01, INFIL = 0.05, LI = 0.1, BLAST = 1.0, TEMP = 0.03
  )
)
```

The predicted values can also be expressed as the probabilities $\pi$.  This produces the familiar signmoidal shape of the binary relationship.

```{r fig.height=3.5}
augment(m1, type.predict = "response") %>%
  ggplot(aes(x = LI)) +
  geom_point(aes(y = REMISS)) +
  geom_line(aes(y = .fitted)) +
  theme_mf() +
  labs(x = "LI",
       y = "Probability of REMISS",
       title = "Binary Fitted Line Plot")
```

Evaluate a logistic model fit with an analysis of deviance.  Deviance is defined as -2 times the log-likelihood $-2l(\beta)$.  The null deviance is the deviance of the null model and is analagous to SST in ANOVA.  The residual deviance is the deviance of the full model and is analagous to SSE in ANOVA.

```{r}
logLik(glm(REMISS ~ ., data = leuk, family = "binomial")) * (-2)
```

`anova()` computes the analysis of deviance table for the model.

```{r}
anova(m1)
```

The deviance of the null model (no regressors) is 34.372.  The deviance of the full model is 21.594. `glance()` also returns the null and residual deviances.  

```{r}
glance(m1)
```

Use the `GainCurvePlot()` function to plot the gain curve (background on gain curve at [Data Science Central](https://www.datasciencecentral.com/profiles/blogs/understanding-and-interpreting-gain-and-lift-charts) from the model predictions. The x-axis is the fraction of items seen when sorted by the predicted value, and the y-axis is the cumulative summed true outcome. The "wizard" curve is the gain curve when the data is sorted by the true outcome.  If the model's gain curve is close to the wizard gain curve, then the model sorted the response variable well. The grey area is the gain over a random sorting.

```{r}
augment(m1) %>% data.frame() %>% 
  GainCurvePlot(xvar = ".fitted", truthVar = "REMISS", title = "Logistic Model")
```

`REMISS` equals 1 in 9 of the 27 responses. 

* The wizard curve shows that after sorting the responses it encounters all 9 1s (100%) after looking at 9 of the 27 response (33%).  
* The bottom of the grey diagonal shows that after making random predictions and sorting the predictions, it encounters only 3 1s (33%) after looking at 9 of the 27 responses (33%).  It has to look at all 27 responses (100%) to encounter all 9 1s (100%).  
* The gain curve encounters 5 1s (55%) after looking at 9 of the 27 responses (33%).  It has to look at 14 responses to encounter all 9 1s (100%).

Another way to evaluate the predictive model is the ROC curve.  It evaluates all possible thresholds for splitting predicted probabilities into predicted classes.  This is often a much more useful metric than simply ranking models by their accuracy at a set threshold, as different models might require different calibration steps (looking at a confusion matrix at each step) to find the optimal classification threshold for that model.

```{r}
library(caTools)
colAUC(m1$fitted.values, m1$data$REMISS, plotROC = TRUE)
```


### Case Study

This case study is from [PSU STAT 504](https://online.stat.psu.edu/stat504/node/197/).  This study investigated the gender differences in [Piaget's water level test](https://online.stat.psu.edu/stat504/node/197/). 166 subjects performed the test and other tests of knowledge and spatial ability.

```{r h2o_data, include=FALSE}
h2o <- tribble(
  ~obs, ~y, ~sex, ~gravity, ~totphys, ~bryant, ~vander, ~triange, ~trailer, ~treee, ~compphys, ~moving, ~total,
  1,  0,  1,  4,  5,  3,  10,  0,  6,  1,  1,  1,  25,
  2,  1,  2,  5,  9,  6,  12,  0,  6,  4,  4,  1,  37,
  3,  0,  1,  3,  3,  5,   4,  2,  5,  0,  0,  1,  19,
  4,  0,  1,  4,  7,  0,   6,  3,  6,  2,  3,  2,  24,
  5,  0,  2,  4,  6,  2,  10,  3,  5,  4,  2,  1,  30,
  6,  1,  2,  4,  8,  6,  12,  1,  5,  2,  4,  1,  34,
  7,  1,  2,  5,  8,  6,  12,  1,  6,  4,  3,  1,  37,
  8,  1,  2,  5,  7,  6,   9,  0,  3,  4,  2,  1,  29,
  9,  0,  1,  5,  6,  3,   8,  0,  6,  4,  1,  1,  27,
   10,  0,  2,  3,  5,  3,   2,  4,  1,  2,  2,  1,  17,
   11,  1,  2,  2,  4,  5,  10,  0,  6,  3,  2,  1,  28,
   12,  0,  1,  1,  3,  2,   9,  3,  0,  3,  2,  1,  20,
   13,  1,  1,  5,  7,  6,  10,  0,  6,  4,  2,  1,  33,
   14,  1,  1,  4,  6,  6,  12,  3,  6,  4,  2,  1,  37,
   15,  0,  1,  2,  5,  4,   5,  0,  5,  0,  3,  1,  19,
   16,  1,  1,  2,  3,  4,  10,  0,  3,  3,  1,  1,  23,
   17,  0,  1,  1,  2,  6,   8,  1,  6,  0,  1,  1,  23,
   18,  1,  2,  4,  6,  5,  12,  0,  6,  4,  2,  1,  33,
   19,  1,  2,  5,  7,  4,  12,  0,  3,  4,  2,  1,  30,
   20,  1,  1,  5,  7,  6,  12,  1,  6,  4,  2,  1,  36,
   21,  0,  1,  3,  4,  4,   0,  0,  6,  4,  1,  2,  18,
   22,  0,  1,  4,  6,  2,   6,  0,  3,  2,  2,  1,  19,
   23,  1,  1,  3,  7,  3,  12,  0,  6,  3,  4,  1,  31,
   24,  1,  2,  2,  6,  6,  11,  2,  6,  4,  4,  1,  35,
   25,  0,  1,  0,  2,  4,   8,  2,  6,  2,  2,  2,  24,
   26,  1,  1,  5,  6,  4,  12,  0,  6,  4,  1,  1,  32,
   27,  0,  1,  3,  5,  4,   2,  0,  6,  3,  2,  1,  20,
   28,  1,  2,  4,  6,  6,  12,  2,  6,  2,  2,  1,  34,
   29,  0,  1,  3,  5,  6,   8,  1,  6,  2,  2,  2,  28,
   30,  1,  2,  5,  6,  3,   6,  2,  6,  4,  1,  1,  27,
   31,  1,  2,  5,  9,  2,  10,  2,  6,  4,  4,  1,  33,
   32,  0,  2,  1,  2,  1,   3,  0,  3,  2,  1,  1,  11,
   33,  0,  1,  3,  5,  2,   4,  0,  4,  0,  2,  2,  15,
   34,  1,  1,  3,  5,  3,  12,  0,  6,  4,  2,  1,  30,
   35,  1,  1,  4,  7,  6,  12,  0,  6,  4,  3,  1,  35,
   36,  1,  2,  5,  8,  5,  10,  2,  6,  0,  3,  1,  31,
   37,  0,  1,  4,  5,  4,   9,  0,  6,  2,  1,  1,  26,
   38,  1,  1,  4,  7,  6,   4,  1,  6,  4,  3,  1,  28,
   39,  0,  1,  3,  4,  4,  12,  3,  4,  1,  1,  1,  28,
   40,  0,  2,  3,  6,  3,  12,  0,  5,  4,  3,  1,  30,
   41,  1,  2,  5,  8,  5,   8,  0,  6,  2,  3,  1,  29,
   42,  0,  1,  2,  4,  6,   2,  0,  0,  1,  2,  2,  13,
   43,  0,  1,  2,  4,  5,   4,  0,  4,  1,  2,  1,  18,
   44,  0,  1,  2,  4,  4,   2,  0,  6,  2,  2,  2,  18,
   45,  0,  1,  2,  3,  5,  10,  1,  3,  0,  1,  1,  22,
   46,  0,  1,  1,  2,  5,   6,  2,  2,  1,  1,  2,  18,
   47,  0,  1,  0,  0,  2,  10,  0,  6,  4,  0,  1,  22,
   48,  0,  1,  2,  4,  6,   8,  0,  6,  4,  2,  1,  28,
   49,  1,  1,  1,  2,  1,   4,  1,  6,  1,  1,  1,  15,
   50,  0,  2,  4,  5,  6,   8,  0,  6,  4,  1,  1,  29,
   51,  0,  2,  4,  7,  5,   6,  1,  5,  4,  3,  1,  28,
   52,  1,  2,  4,  6,  4,   6,  1,  6,  2,  2,  1,  25,
   53,  0,  1,  1,  3,  3,   2,  0,  6,  0,  2,  1,  14,
   54,  1,  2,  4,  7,  6,  12,  1,  6,  4,  3,  1,  36,
   55,  0,  2,  1,  3,  3,   5,  5,  6,  2,  2,  2,  24,
   56,  1,  2,  2,  5,  6,   4,  2,  6,  4,  3,  1,  27,
   57,  1,  2,  3,  3,  3,  10,  0,  6,  0,  0,  1,  22,
   58,  0,  1,  3,  6,  2,  10,  9,  3,  4,  3,  1,  34,
   59,  1,  2,  4,  8,  3,  12,  0,  6,  4,  4,  1,  33,
   60,  0,  1,  4,  5,  5,  10,  0,  6,  4,  1,  2,  30,
   61,  0,  1,  3,  5,  3,   2,  5,  5,  2,  2,  1,  22,
   62,  1,  1,  2,  5,  6,   4,  0,  6,  4,  3,  1,  25,
   63,  0,  1,  0,  2,  5,   4,  0,  6,  3,  2,  2,  20,
   64,  0,  2,  3,  6,  5,  10,  2,  4,  4,  3,  1,  31,
   65,  0,  2,  5,  7,  2,  10,  1,  5,  0,  2,  1,  25,
   66,  0,  1,  1,  4,  2,   8,  2,  4,  4,  3,  1,  24,
   67,  0,  2,  3,  6,  3,  12,  0,  6,  2,  3,  1,  29,
   68,  0,  1,  0,  0,  6,   4,  2,  6,  0,  0,  2,  18,
   69,  1,  1,  5,  7,  4,  12,  0,  6,  2,  2,  1,  31,
   70,  1,  2,  5,  9,  3,   4,  1,  6,  4,  4,  1,  27,
   71,  0,  1,  3,  3,  2,   6,  4,  1,  0,  0,  2,  16,
   72,  1,  1,  3,  7,  6,  10,  0,  5,  4,  4,  1,  32,
   73,  0,  1,  1,  3,  4,   8,  0,  6,  2,  2,  2,  23,
   74,  0,  2,  4,  7,  3,   8,  2,  6,  2,  3,  1,  28,
   75,  1,  2,  4,  6,  4,  10,  2,  6,  0,  4,  1,  28,
   76,  0,  2,  1,  3,  4,   0,  2,  4,  0,  2,  1,  13,
   77,  0,  1,  1,  3,  3,   3,  0,  2,  0,  2,  2,  11,
   78,  0,  1,  4,  4,  4,   6,  0,  5,  2,  0,  1,  21,
   79,  0,  1,  4,  5,  6,   4,  1,  6,  4,  1,  2,  26,
   80,  1,  1,  4,  5,  4,   8,  1,  6,  4,  1,  1,  28,
   81,  0,  1,  0,  3,  3,   2,  1,  6,  1,  3,  1,  16,
   82,  0,  1,  2,  3,  2,   0,  1,  1,  1,  1,  2,   8,
   83,  0,  1,  0,  2,  3,   4,  0,  4,  0,  2,  2,  13,
   84,  0,  2,  1,  4,  4,   2,  4,  5,  0,  3,  1,  19,
   85,  0,  1,  2,  2,  3,   2,  0,  0,  1,  0,  2,   8,
   86,  1,  2,  4,  6,  6,  10,  1,  6,  3,  2,  1,  32,
   87,  1,  2,  3,  5,  6,  10,  0,  6,  4,  2,  1,  31,
   88,  0,  1,  4,  7,  4,   8,  1,  6,  4,  3,  2,  30,
   89,  0,  1,  1,  1,  4,   4,  0,  4,  1,  0,  1,  14,
   90,  0,  1,  2,  4,  3,   4,  3,  6,  2,  2,  1,  22,
   91,  1,  1,  3,  4,  3,  10,  3,  6,  4,  1,  1,  30,
   92,  0,  1,  4,  5,  3,   6,  0,  6,  4,  1,  1,  24,
   93,  1,  2,  4,  7,  6,   8,  0,  6,  4,  3,  1,  31,
   94,  0,  2,  4,  6,  3,  12,  1,  6,  1,  2,  1,  29,
   95,  1,  1,  3,  7,  6,   8,  1,  6,  4,  4,  1,  32,
   96,  1,  2,  4,  8,  3,  12,  0,  6,  3,  4,  1,  32,
   97,  0,  1,  2,  5,  4,   2,  0,  2,  4,  3,  1,  17,
   98,  1,  1,  3,  4,  3,   8,  2,  6,  4,  1,  1,  27,
   99,  0,  2,  1,  2,  3,   6,  0,  4,  1,  1,  1,  16,
  100,  1,  2,  5,  8,  6,  12,  0,  6,  4,  3,  1,  36,
  101,  1,  1,  5,  8,  4,   6,  2,  5,  3,  3,  1,  28,
  102,  1,  1,  5,  8,  2,   6,  2,  6,  4,  3,  1,  28,
  103,  1,  1,  5,  9,  4,   6,  1,  6,  4,  4,  2,  30,
  104,  1,  1,  4,  5,  5,  12,  2,  6,  0,  1,  1,  30,
  105,  1,  1,  3,  7,  5,  12,  2,  6,  4,  4,  1,  36,
  106,  1,  1,  2,  4,  3,  12,  0,  6,  4,  2,  1,  29,
  107,  0,  1,  4,  6,  3,   5,  2,  6,  4,  2,  1,  26,
  108,  0,  1,  1,  2,  4,   4,  0,  6,  2,  1,  1,  18,
  109,  0,  1,  0,  1,  5,   6,  2,  4,  1,  1,  1,  19,
  110,  0,  1,  2,  4,  3,   4,  0,  3,  2,  2,  2,  16,
  111,  0,  1,  1,  3,  5,   4,  0,  6,  4,  2,  2,  22,
  112,  0,  1,  1,  3,  4,   4,  0,  6,  3,  2,  2,  20,
  113,  0,  2,  1,  3,  2,  10,  1,  0,  4,  2,  1,  20,
  114,  1,  2,  4,  6,  2,   8,  3,  6,  4,  2,  1,  29,
  115,  1,  2,  3,  7,  4,   0,  2,  6,  4,  4,  1,  23,
  116,  1,  2,  4,  5,  4,   3,  1,  6,  2,  1,  1,  21,
  117,  1,  2,  5,  9,  2,  12,  0,  6,  4,  4,  1,  33,
  118,  0,  1,  3,  5,  3,   4,  2,  5,  0,  2,  1,  19,
  119,  0,  1,  4,  7,  3,  12,  0,  6,  4,  3,  1,  32,
  120,  0,  1,  0,  1,  5,   4,  0,  6,  1,  1,  1,  17,
  121,  1,  2,  5,  8,  6,   8,  0,  6,  4,  3,  1,  32,
  122,  0,  2,  4,  6,  2,   2,  2,  6,  3,  2,  1,  21,
  123,  1,  1,  2,  3,  5,   8,  0,  6,  4,  1,  1,  26,
  124,  1,  1,  5,  8,  3,  12,  0,  6,  4,  3,  1,  33,
  125,  0,  1,  2,  5,  6,   8,  2,  6,  4,  3,  1,  31,
  126,  1,  2,  4,  5,  2,  10,  2,  6,  4,  3,  1,  29,
  127,  0,  2,  4,  5,  3,   6,  2,  4,  4,  1,  1,  24,
  128,  1,  1,  3,  3,  6,   8,  0,  5,  4,  0,  1,  26,
  129,  0,  1,  1,  3,  4,  10,  1,  6,  2,  2,  1,  26,
  130,  0,  1,  1,  3,  4,   2,  1,  6,  2,  2,  1,  18,
  131,  0,  2,  1,  2,  6,   0,  1,  6,  3,  1,  1,  18,
  132,  0,  1,  1,  3,  2,   0,  0,  3,  1,  2,  1,   9,
  133,  1,  1,  4,  6,  3,   4,  0,  6,  4,  2,  1,  23,
  134,  1,  1,  5,  6,  3,  10,  0,  6,  3,  1,  1,  28,
  135,  1,  1,  5,  8,  4,  10,  0,  6,  4,  3,  2,  32,
  136,  0,  1,  3,  5,  4,   4,  1,  6,  0,  2,  1,  20,
  137,  0,  1,  2,  3,  5,   8,  0,  6,  0,  1,  1,  22,
  138,  0,  1,  2,  3,  5,   0,  4,  4,  2,  3,  2,  18,
  139,  0,  1,  1,  1,  4,  10,  1,  6,  4,  0,  1,  26,
  140,  0,  1,  5,  9,  4,   6,  0,  6,  1,  4,  1,  26,
  141,  0,  1,  1,  2,  3,   2,  2,  6,  3,  1,  1,  18,
  142,  0,  1,  3,  6,  3,   6,  0,  6,  4,  3,  1,  25,
  143,  1,  1,  4,  6,  5,  10,  0,  6,  3,  2,  1,  30,
  144,  1,  2,  2,  5,  3,  12,  1,  6,  4,  3,  1,  31,
  145,  1,  2,  3,  6,  4,  10,  0,  6,  4,  3,  1,  30,
  146,  1,  2,  5,  9,  6,  10,  0,  6,  4,  4,  1,  35,
  147,  1,  2,  2,  5,  5,   8,  2,  5,  4,  3,  1,  29,
  148,  1,  2,  2,  5,  3,  10,  0,  6,  1,  3,  1,  25,
  149,  0,  2,  2,  4,  3,   6,  0,  6,  4,  2,  1,  23,
  150,  0,  1,  4,  5,  3,   4,  6,  5,  1,  1,  1,  24,
  151,  0,  1,  3,  5,  6,   4,  4,  6,  1,  2,  2,  26,
  152,  1,  2,  5,  7,  5,  12,  0,  6,  4,  2,  1,  34,
  153,  0,  1,  2,  4,  2,   6,  3,  5,  4,  2,  1,  24,
  154,  1,  1,  5,  8,  6,  12,  0,  6,  4,  3,  1,  36,
  155,  0,  1,  3,  6,  6,   8,  0,  3,  0,  3,  2,  23,
  156,  0,  1,  0,  2,  5,  12,  6,  4,  0,  2,  2,  29,
  157,  0,  1,  4,  8,  3,   0,  2,  6,  0,  4,  1,  19,
  158,  1,  1,  1,  2,  4,   8,  2,  6,  4,  1,  1,  26,
  159,  0,  1,  2,  3,  5,   8,  0,  6,  3,  1,  1,  25,
  160,  0,  1,  4,  7,  5,   5,  1,  6,  1,  3,  1,  25,
  161,  1,  2,  3,  5,  6,   8,  2,  6,  4,  2,  1,  31,
  162,  0,  1,  0,  0,  5,   8,  2,  6,  1,  0,  1,  22,
  163,  0,  1,  2,  4,  5,   4,  0,  1,  0,  2,  1,  14,
  164,  0,  2,  5,  8,  4,  12,  0,  6,  4,  3,  1,  34,
  165,  0,  1,  4,  4,  5,   2,  0,  5,  3,  0,  1,  19,
  166,  0,  1,  4,  7,  5,  12,  2,  6,  3,  3,  1,  35
)
h2o <- h2o %>%
  mutate(
    moving = factor(moving),
    sex = factor(sex, levels = c(2:1), labels = c("M", "F")),
    y = factor(y, labels = c("fail", "pass"))
  )
```

Here are the pass/fail results of the study by sex.

```{r}
prop.table(table(h2o$sex, h2o$y), margin = 1) %>% round(digits = 4)
```

The pass rates are very different. The Pearson Chi-squared test of equality of two proportions calculates a Chi-Square value of 18.562, p-value < 0.0001, so reject the hypothesis that the pass rates are the same.

```{r}
chisq.test(table(h2o$sex, h2o$y), correct = FALSE)
```

Fit a logistic regression.

```{r}
m <- glm(y ~ sex, data = h2o, family = binomial(link = logit))
anova(m)
```

The likelihood ratio is $G^2 = 18.658$, so reject the null hypothesis of no sex effect - there is statistically significant difference in pass rates.  Here are the coefficient estimators.

```{r}
tidy(m)
```

The odds ratio (females vs males) is 

```{r}
exp(tidy(m)[2, ]$estimate)
```

meaning the odds of a female passing are 0.24 times that of a male.  Or take the inverse to get the male vs female odds ratio, 

```{r}
1 / exp(tidy(m)[2, ]$estimate)
```

meaning the odds of a male passing are 4.24 times that of a female.  `oddsratio::or_glm()` calculates the odds ratio from an increment in the predictor values, in this case, incrementing sex from male to female.

```{r}
oddsratio::or_glm(h2o, m, 1)
```

What if we control for the results of the gravity test (gravity = number of correct answers on test)?  Here is a logistic regression of pass ~ gravity.

```{r}
m <- glm(y ~ gravity, data = h2o, family = binomial(link = logit))
anova(m)
```

The likelihood ratio is $G^2 = 42.2$, so reject H0 that there is no gravity effect - there is a statistically significant difference between the gravity score and the pass rate.  The odds of passing the water level task increase by 2.225 for each additional right answer on gravity.

```{r}
oddsratio::or_glm(h2o, m, incr = list(gravity = 1))
```

```{r fig.height=3, fig.width=5}
augment(m) %>%
  group_by(gravity) %>% 
  summarize(
    observed = mean(as.numeric(y) - 1),
    fitted = mean(exp(.fitted) / (1 + exp(.fitted)))
  ) %>% 
  pivot_longer(cols = -gravity, names_to = "model", values_to = "rate") %>%
  ggplot(aes(x = gravity, y = rate, color = model)) +
  geom_point() +
  scale_color_mf() +
  theme_mf() +
  labs(title = "Water Test Pass Rate", color = "")
```

You can model sex and gravity with and without interaction effects. Here is the model without interaction effects.  $G^2 = 32.3$ so 

```{r}
m <- glm(y ~ sex + gravity, data = h2o, family = binomial(link = "logit"))
anova(m)
```

```{r}
tidy(m)
```

## Multinomial Logistic Regression

The following notes rely heavily on the [PSU STAT 504 course notes](https://online.stat.psu.edu/stat504/node/171/.

Multinomial logistic regression models the odds the multinomial response variable $Y \sim Mult(n, \pi)$ is in level $j$ relative to baseline category $j^*$ for all pairs of categories as a function of $k$ explanatory variables, $X = (X_1, X_2, ... X_k)$. 

$$\log \left( \frac{\pi_{ij}}{\pi_{ij^*}} \right) = x_i^T \beta_j, \hspace{5mm} j \ne j^2$$

Interpet the $k^{th}$ element of $\beta_j$ as the increase in log-odds of falling a response in category $j$ relative to category $j^*$ resulting from a one-unit increase in the $k^{th}$ predictor term, holding the other terms constant.

Multinomial model is a type of GLM.

Here is an example using multinomial logistic regression.  A researcher classified the stomach contents of $n = 219$ alligators according to $r = 5$ categories (*fish, Inv., Rept, Bird, Other*) as a function of covariates *Lake*, *Sex*, and *Size*..

```{r}
gator_dat <- tribble(
  ~profile, ~Gender, ~Size, ~Lake, ~Fish, ~Invertebrate, ~Reptile, ~Bird, ~Other,
  "1", "f", "<2.3", "george",  3, 9, 1, 0, 1,
  "2", "m", "<2.3", "george", 13, 10, 0, 2, 2,
  "3", "f", ">2.3", "george", 8, 1, 0, 0, 1,
  "4", "m", ">2.3", "george", 9, 0, 0, 1, 2,
  "5", "f", "<2.3", "hancock", 16, 3, 2, 2, 3,
  "6", "m", "<2.3", "hancock", 7, 1, 0, 0, 5,
  "7", "f", ">2.3", "hancock", 3, 0, 1, 2, 3,
  "8", "m", ">2.3", "hancock", 4, 0, 0, 1, 2,
  "9", "f", "<2.3", "oklawaha", 3, 9, 1, 0, 2,
  "10", "m", "<2.3", "oklawaha", 2, 2, 0, 0, 1,
  "11", "f", ">2.3", "oklawaha", 0, 1, 0, 1, 0,
  "12", "m", ">2.3", "oklawaha", 13, 7, 6, 0, 0,
  "13", "f", "<2.3", "trafford", 2, 4, 1, 1, 4,
  "14", "m", "<2.3", "trafford", 3, 7, 1, 0, 1,
  "15", "f", ">2.3", "trafford", 0, 1, 0, 0, 0,
  "16", "m", ">2.3", "trafford", 8, 6, 6, 3, 5
)
gator_dat <- gator_dat %>%
  mutate(
    Gender = as_factor(Gender),
    Lake = fct_relevel(Lake, "hancock"),
    Size = as_factor(Size)
  )
```

There are 4 equations to estimate:

$$\log \left( \frac{\pi_j} {\pi_{j^*}} \right) = \beta X$$

where $\pi_{j^*}$ is the probability of fish, the baseline category.  


Run a multivariate logistic regression model with `VGAM::vglm()`.

```{r message=FALSE}
library(VGAM)
```

`vglm()` fits 4 logit models.

```{r}
gator_vglm <- vglm(
  cbind(Bird,Invertebrate,Reptile,Other,Fish) ~ Lake + Size + Gender, 
  data = gator_dat, 
  family = multinomial
)
summary(gator_vglm)
```

The **residual deviance** is 50.2637 on 40 degrees of freedom.  Residual deviance tests the current model fit versus the saturated model. The saturated model, which fits a separate multinomial distribution to each of the 16 profiles (unique combinations of lake, sex and size), has 16 × 4 = 64 parameters. The current model has an intercept, three lake coefficients, one sex coefficient and one size coefficient for each of the four logit equations, for a total of 24 parameters. Therefore, the overall fit statistics have 64 − 24 = 40 degrees of freedom.

```{r}
E <- data.frame(fitted(gator_vglm) * rowSums(gator_dat[, 5:9]))
O <- gator_dat %>% select(Bird, Invertebrate, Reptile, Other, Fish) + .000001
(g2 <- 2 * sum(O * log(O / E)))
```


indicates the model fits okay, but not great. The Residual Deviance of 50.26 with 40 df from the table above output is reasonable, with p-value of 0.1282 and the statistics/df is close to 1 that is 1.256.



## Ordinal Logistic Regression

The following notes rely heavily on [this UVA web page](https://data.library.virginia.edu/fitting-and-interpreting-a-proportional-odds-model/) and [PSU STAT 504 class notes](https://online.stat.psu.edu/stat504/node/171/.

The ordinal logistic regression model is

$$logit[P(Y \le j)] = \alpha_j - \beta X, \hspace{5mm} j \in [1, J-1]$$
where $j \in [1, J-1]$ are the levels of the ordinal outcome variable $Y$.  The proportional odds model assumes there is a common set of slope parameters $\beta$ for the predictors.  The ordinal outcomes are distinguished by the $J-1$ intercepts $\alpha_j$.  The benchmark level is $J$.  "Logit" means log-odds, so $logit[P(Y \le j)] = \log[P(Y \le j) / P(Y \gt j)]$.

Suppose we want to model the probability a respondent holds a political ideology ["Very Liberal", "Slightly Liberal", "Moderate", "Slightly Conservative", "Very Conservative"] given their party affiliation ["Republican", "Democrat"].

```{r}
ideo <- c(
  "Very Liberal",
  "Slightly Liberal",
  "Moderate",
  "Slightly Conservative",
  "Very Conservative"
)
ideo_cnt_rep <- c(30, 46, 148, 84, 99)
ideo_cnt_dem <- c(80, 81, 171, 41, 55)
dat <- data.frame(
  party = factor(rep(c("Rep", "Dem"), c(407, 428)), levels = c("Rep", "Dem")),
  ideo = factor(
    c(rep(ideo, ideo_cnt_rep), rep(ideo, ideo_cnt_dem)), 
    levels = ideo
  )
)
table(dat)
```

Fit a proportional odds model with the `MASS::polr()` function (polr stands for proportional odds linear regression).

```{r}
pom <- MASS::polr(ideo ~ party, data = dat)

summary(pom)
```

The log-odds a Democrat identifies as “Very Liberal” or lower is 
$$logit[P(Y \lt 1)] = -2.4690 - (-0.9745)(1) = -1.4945.$$
Solve $logit[P(Y \le 1)] = \log[P(Y \le 1) / P(Y \gt 1)]$ for $P(Y \le 1)$ to get the probability a Democrat identifies as “Very Liberal” or lower as $P(Y \le 1) = exp(-1.4945) / (1 + exp(-1.4945)) = 0.183$. 


The log odds a Democrat identifies as “Slightly Liberal” or lower is
$$logit[P(Y \lt 2)] = -1.4745 - (-0.9745)(1) = -0.5.$$
The corresponding probability is $P(Y \le 2) = exp(-0.5) / (1 + exp(-0.5)) = 0.378$. 

To get the probability a Democrat identifies as "Slightly Liberal", just subtract the adjacent cumulative probabilities, $P(Y \le 2) - P(Y \le 1) = 0.378 = 0.183 = 0.194$.  That's what's happening when you use the model to predict the level probabilities.

```{r}
predict(pom, newdata = data.frame(party = "Dem"), type = "probs")
```

The baseline for the model was "Republican", so the log-odds a Republicn identifies as "Very Liberal" or lower is 
$$logit[P(Y \lt 1)] = -2.4690 - (-0.9745)(0) = -2.4690$$
with corresponding probability $P(Y \le 1) = exp(-2.4690) / (1 + exp(-2.4690)) = 0.078$.  And so on.

```{r}
predict(pom, newdata = data.frame(party = "Rep"), type = "probs")
```

Here is a manual calculation of the probabilities.  Note the `zeta` variable instead of `beta` because the model fits $\alpha_j - \beta X$ instead of $\alpha_j + \beta X$, and so it uses a new variable $\zeta = -\beta$.  Technically, the model could be expressed $logit[P(Y \le j)] = \alpha_j + \zeta X$.

```{r}
library(stringr)
# cum log odds, dems an reps
dclo <- c(pom$zeta - pom$coefficients)
rclo <- c(pom$zeta - 0)
# cum probs, dems and reps
dcp <- exp(dclo) / (1 + exp(dclo))
rcp <- exp(rclo) / (1 + exp(rclo))
# fix the names and add 1 for "Very..."
names(dcp) <- str_sub(names(dcp), start = 1, end = str_locate(names(dcp), "\\|")[, 1] - 1)
names(rcp) <- str_sub(names(rcp), start = 1, end = str_locate(names(rcp), "\\|")[, 1] - 1)
dcp <- c(dcp, 1)
rcp <- c(rcp, 1)
names(dcp) <- c(names(dcp)[1:4], "Very Conservative")
names(rcp) <- c(names(rcp)[1:4], "Very Conservative")
# Democrat probs
(dp <- dcp - lag(dcp, 1, 0))
(rp <- rcp - lag(rcp, 1, 0))
```

The "proportional odds" part of the proportional odds model is that the ratios of the $J - 1$ odds-ratios are identical for the different levels of the predictors.  Here we have a single predictor, `party`.  The odds ratios for `party = Dem` are

```{r}
(dcp / (1 - dcp)) / (rcp / (1 - rcp))
```

Which is to say, any level of ideology $j$, the estimated odds that a Democrat’s ideology is more liberal $(\lt j)$ rather than more conservative $(\ge j)$ is about 2.65 times a Republicans odds.

The log of these odds ratios is the coefficient estimator for `party`.
```{r}
log((dcp / (1 - dcp)) / (rcp / (1 - rcp)))
```

Always check the assumption of proportional odds. One way to do this is by comparing the proportional odds model with a multinomial logit model, also called an unconstrained baseline logit model. The multinomial logit model models *unordered* responses and fits a slope to each level of the $J – 1$ responses. The proportional odds model is nested in the multinomial model, so can use a likelihood ratio test to see if the models are statistically different.

```{r}
mlm <- nnet::multinom(ideo ~ party, data = dat)
```

Calculate the deviance test statistic $D = -2 loglik(\beta)$.

```{r}
G <- -2 * (logLik(pom)[1] - logLik(mlm)[1])
pchisq(G, df = length(pom$zeta) - 1, lower.tail = FALSE)
```

The p-value is high, so the proportional odds model fits as well as the more complex multinomial logit model.


## Poisson Regression

Poisson models count data, like "traffic tickets per day", or "website hits per day".  The response is an expected *rate* or intensity.  For count data, specify the generalized model, this time with `family = poisson` or `family = quasipoisson`. 

Recall that the probability of achieving a count $y$ when the expected rate is $\lambda$ is distributed 

$$P(Y = y|\lambda) = \frac{e^{-\lambda} \lambda^y}{y!}.$$


The poisson regression model is

$$\lambda = \exp(X \beta).$$ 
 
You can solve this for $y$ to get

$$y = X\beta = \ln(\lambda).$$

That is, the model predicts the log of the response rate.  For a sample of size *n*, the likelihood function is

$$L(\beta; y, X) = \prod_{i=1}^n \frac{e^{-\exp({X_i\beta})}\exp({X_i\beta})^{y_i}}{y_i!}.$$

The log-likelihood is

$$l(\beta) = \sum_{i=1}^n (y_i X_i \beta - \sum_{i=1}^n\exp(X_i\beta) - \sum_{i=1}^n\log(y_i!).$$

Maximizing the log-likelihood has no closed-form solution, so the coefficient estimates are found through interatively reweighted least squares.  

Poisson processes assume the variance of the response variable equals its mean.  "Equals" means the mean and variance are of a similar order of magnitude.  If that assumption does not hold, use the quasi-poisson.  Use Poisson regression for large datasets.  If the predicted counts are much greater than zero (>30), the linear regression will work fine.  Whereas RMSE is not useful for logistic models, it is a good metric in Poisson.


### Example {-}

Dataset `fire` contains response variable `injuries` counting the number of injuries during the month and one explanatory variable, the month `mo`.

```{r}
fire <- read_csv(file = "C:/Users/mpfol/OneDrive/Documents/Data Science/Data/CivilInjury_0.csv")
fire <- fire %>% 
  mutate(mo = as.POSIXlt(`Injury Date`)$mon + 1) %>%
  rename(dt = `Injury Date`,
         injuries = `Total Injuries`)
str(fire)
```

In a situation like this where there the relationship is bivariate, start with a visualization.

```{r}
ggplot(fire, aes(x = mo, y = injuries)) +
  geom_jitter() +
  geom_smooth(method = "glm", method.args = list(family = "poisson")) +
  labs(title = "Injuries by Month")
```


Fit a poisson regression in R using `glm(formula, data, family = poisson)`.  But first, check whether the mean and variance of `injuries` are the same magnitude?  If not, then use `family = quasipoisson`.

```{r}
mean(fire$injuries)
var(fire$injuries)
```

They are of the same magnitude, so fit the regression with `family = poisson`.

```{r}
m2 <- glm(injuries ~ mo, family = poisson, data = fire)
summary(m2)
```

The *predicted* value $\hat{y}$ is the estimated **log** of the response variable, 

$$\hat{y} = X \hat{\beta} = \ln (\lambda).$$

Suppose `mo` is January (mo = `), then the log of `injuries` is $\hat{y} = 0.323787$. Or, more intuitively, the expected count of injuries is $\exp(0.323787) = 1.38$  

```{r}
predict(m2, newdata = data.frame(mo=1))
predict(m2, newdata = data.frame(mo=1), type = "response")
```

Here is a plot of the predicted counts in red.

```{r}
augment(m2, type.predict = "response") %>%
  ggplot(aes(x = mo, y = injuries)) +
  geom_point() +
  geom_point(aes(y = .fitted), color = "red") + 
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Month",
       y = "Injuries",
       title = "Poisson Fitted Line Plot")
```

Evaluate a logistic model fit with an analysis of deviance.  

```{r}
(perf <- glance(m2))
(pseudoR2 <- 1 - perf$deviance / perf$null.deviance)
```

The deviance of the null model (no regressors) is 139.9.  The deviance of the full model is 132.2.  The psuedo-R2 is very low at .05.  How about the RMSE?

```{r}
RMSE(pred = predict(m2, type = "response"), obs = fire$injuries)
```

The average prediction error is about 0.99.  That's almost as much as the variance of `injuries` - i.e., just predicting the mean of `injuries` would be almost as good!  Use the `GainCurvePlot()` function to plot the gain curve.

```{r}
augment(m2, type.predict = "response") %>%
  ggplot(aes(x = injuries, y = .fitted)) +
  geom_point() +
  geom_smooth(method ="lm") +
  labs(x = "Actual",
       y = "Predicted",
       title = "Poisson Fitted vs Actual")
```


```{r}
augment(m2) %>% data.frame() %>% 
  GainCurvePlot(xvar = ".fitted", truthVar = "injuries", title = "Poisson Model")
```

It seems that `mo` was a poor predictor of `injuries`.  

