--- 
title: "My Data Science Notes"
author: "Michael Foley"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    lib_dir: assets
    split_by: section
    config:
      toolbar:
        position: static
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a compendium of notes from classes, tutorials, etc. that I reference from time to time."
---
--- 
title: "My Data Science Notes"
author: "Michael Foley"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    lib_dir: assets
    split_by: section
    config:
      toolbar:
        position: static
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a compendium of notes from classes, tutorials, etc. that I reference from time to time."
---

# Intro {-}

These notes are pulled from various classes, tutorials, books, etc. and are intended for my own consumption.  If you are finding this on the internet, I hope it is useful to you, but you should know that I am just a student and there's a good chance whatever you're reading here is mistaken.  In fact, that should probably be your null hypothesis... or your prior.  Whatever.

<!--chapter:end:index.Rmd-->


# Probability {#probability}

Placeholder


## Principles
## Discrete Distributions {#disc_dist}
### Bernoulli
### Binomial
### Poission
### Multinomial
### Negative-Binomial
#### Examples {-}
### Geometric
#### Examples {-}
### Hypergeometric
#### Example {-}
### Gamma
#### Examples {-}
## Continuous Distributions {#cont_dist}
### Normal
#### Example {-}
#### Example
#### Example
#### Example
#### Normal Approximation to Binomial
#### Example
#### Example
#### From Sample to Population
## Join Distributions
## Likelihood

<!--chapter:end:01-probability.Rmd-->


# Categorical Analysis - Nonmodel {#discrete_analysis}

Placeholder


## Chi-Square Test
## One-Way Tables
### Chi-Square Goodness-of-Fit Test
### Proportion Test
## Two-Way Tables
### Chi-Square Independence Test
### Residuals Analysis
### Difference in Proportions
### Relative Risk
### Odds Ratio
### Partitioning Chi-Square
### Correlation
## K-Way Tables
### Odds Ratio
### Chi-Square Independence Test
#### Mutual Independence
#### Joint Independence
#### Marginal Independence
#### Conditional Independence
#### Homogenous Associations

<!--chapter:end:02-discrete_analysis.Rmd-->


# Continuous Variable Analysis {#continuous_analysis}

Placeholder


### Correlation
#### Pearson's r
#### Spearman's Rho
#### Kendall's Tau

<!--chapter:end:03-continuous_analysis.Rmd-->

```{r include=FALSE}
library(tidyverse)
library(mfstylr)
```

# Experiment Design

These notes are structured from the [PSU STAT 503](https://online.stat.psu.edu/stat503/) course.

## Single Factor

## Blocking

## Nested

## Split Plot

<!--chapter:end:04-experiment_design.Rmd-->


# PART 2: Supervised Machine Learning {-}

Placeholder


## Linear Regression Model
## Parameter Estimation
#### Example
## Model Assumptions
### Linearity
### Multicollinearity
#### Example
#### Example
### Normality
### Equal Variances
## Prediction
#### Example
#### Example 
## Inference
### *t*-Test
#### Example
### *F*-Test
#### Example
## Interpretation
## Model Validation
### Accuracy Metrics
#### R-Squared
#### RMSE
#### RSE
#### MAE
#### Adjusted R-squared
#### AIC
#### AICc
#### BIC
#### Mallows Cp
##### Example
### Cross-Validation
#### Validation Set
##### Example
#### LOOCV
#### K-fold Cross-Validation
#### Repeated K-fold CV
#### Bootstrapping
### Gain Curve
## Reference

<!--chapter:end:06-ols.Rmd-->


# Generalized Linear Models

Placeholder


## Logistic Regression
## Multinomial Logistic Regression
## Ordinal Logistic Regression
### Assumptions
### Modeling 
### Case Study
## Poisson Regression

<!--chapter:end:07-glm.Rmd-->


# Multivariate Statistical Analysis

Placeholder


## Background
## MANOVA
## Repeated Measures
## LDA
## PCA
## Factor Analysis
## Canonical Correlation
## Cluster Analysis

<!--chapter:end:08-multivariate.Rmd-->

# Classification

<!--chapter:end:09-pls.Rmd-->


# Regularization

Placeholder


## Ridge
#### Example {-}
## Lasso
#### Example {-}
## Elastic Net
#### Example {-}
## Model Summary {-}

<!--chapter:end:10-glmnet.Rmd-->

# Decision Trees

Decision trees, also known as classification and regression tree (CART) models, are tree-based methods for supervised machine learning.  Simple *classification trees* and *regression trees* are easy to use and interpret, but are not competitive with the best machine learning methods. However, they form the foundation for **bagged trees**, **random forests**, and **boosted trees** ensemble models, which although less interpretable, are very accurate.

CART models segment the predictor space into $K$ non-overlapping terminal nodes (leaves), $A_1, A_2, \dots, A_K$.  Each node is described by a set of rules which can be used to predict new responses. The predicted value $\hat{y}$ for each node is the mode (classification) or mean (regression).

CART models define the nodes through a *top-down greedy* process called *recursive binary splitting*. The process is *top-down* because it begins at the top of the tree with all observations in a single region and successively splits the predictor space. It is *greedy* because at each splitting step, the best split is made at that particular step without consideration to subsequent splits.

The best split is the predictor variable and cutpoint that minimizes a cost function. For a regression tree, the most common cost function is the sum of squared residuals, 

$$RSS = \sum_{k=1}^K\sum_{i \in A_k}{\left(y_i - \hat{y}_{A_k} \right)^2}.$$

For a classification tree, the most common cost functions are the Gini index, 

$$G = \sum_{c=1}^C{\hat{p}_{kc}(1 - \hat{p}_{kc})},$$

and the entropy (aka information statistic)

$$D = - \sum_{c=1}^C{\hat{p}_{kc} \log \hat{p}_{kc}}$$

where $\hat{p}_{kc}$ is the proportion of training observations in node $k$ that are class $c$.  For a completely *pure* node in a binary tree, $\hat{p} \in \{ 0, 1 \}$ and $G = D = 0$. For a completely *impure* node in a binary tree, $\hat{p} = 0.5$ and $G = 0.5^2 \cdot 2 = 0.25$ and $D = -(0.5 \log(0.5)) \cdot 2 = 0.69$.

CART repeats the splitting process for each of the child nodes until a *stopping criterion* is satisfied, usually when no node size surpasses a predefined maximum, or continued splitting does not improve the model significantly.  CART may also impose a minimum number of observations in each node.

The resulting tree likely over-fits the training data and therefore does not generalize well to test data, so CART *prunes* the tree, minimizing the cross-validated prediction error. Rather than cross-validating every possible subtree to find the one with minimum error, CART uses *cost-complexity pruning*. Cost-complexity is the tradeoff between error (cost) and tree size (complexity) where the tradeoff is quantified with cost-complexity parameter $c_p$.  In the equation below, the cost complexity of the tree $R_{c_p}(T)$ is the sum of its risk (error) plus a "cost complexity" factor $c_p$ multiple of the tree size $|T|$.  

$$R_{c_p}(T) = R(T) + c_p|T|$$

$c_p$ can take on any value from $[0..\infty]$, but it turns out there is an optimal tree for *ranges* of $c_p$ values, so there are only a finite set of *interesting* values for $c_p$ [@James2013] [@Therneau2019] [@Kuhn2016].  A parametric algorithm identifies the interesting $c_p$ values and their associated pruned trees, $T_{c_p}$. CART uses cross-validation to determine which $c_p$ is optimal.


## Classification Tree

A simple classification tree is rarely performed on its own; the bagged, random forest, and gradient boosting methods build on this logic. However, it is good to start here to build understanding. I'll learn by example. Using the `ISLR::OJ` data set, I will use 17 feature variables to predict which brand of orange juice, Citrus Hill (CH) or Minute Maid = (MM), customers `Purchase`.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(caret)
library(rpart)  # classification and regression trees 
library(rpart.plot)  # better formatted plots than the ones in rpart
library(plotROC)  # ROC curves
library(ROCR)

oj_dat <- ISLR::OJ
skimr::skim(oj_dat)
```

I'll split `oj_dat` (n = 1,070) into `oj_train` (80%, n = 857) to fit various models, and `oj_test` (20%, n = 213) to compare their performance on new data.

```{r}
set.seed(12345)
partition <- createDataPartition(y = oj_dat$Purchase, p = 0.8, list = FALSE)
oj_train <- oj_dat[partition, ]
oj_test <- oj_dat[-partition, ]
```

Function `rpart::rpart()` builds a full tree, minimizing the Gini index $G$ by default (`parms = list(split = "gini")`), until the stopping criterion is satisfied.  The default stopping criterion is 

* only attempt a split if the current node has at least `minsplit = 20` observations,
* only accept a split if 
   * the resulting nodes have at least `minbucket = round(minsplit/3)` observations, and 
   * the resulting overall fit improves by `cp = 0.01` (i.e., $\Delta G <= 0.01$).

```{r}
# Use method = "class" for classification, method = "anova" for regression
set.seed(123)
oj_mdl_cart_full <- rpart(formula = Purchase ~ ., data = oj_train, method = "class")
print(oj_mdl_cart_full)
```

The output starts with the root node.  The predicted class at the root is `CH` and this prediction produces 334 errors on the 857 observations for a success rate of 0.61026838 and an error rate of 0.38973162.  The child nodes of node "x" are labeled 2x) and 2x+1), so the child nodes of 1) are 2) and 3), and the child nodes of 2) are 4) and 5).  Terminal nodes are labeled with an asterisk (*).  

Surprisingly, only 3 of the 17 features were used the in full tree: `LoyalCH` (Customer brand loyalty for CH), `PriceDiff` (relative price of MM over CH), and `SalePriceMM` (absolute price of MM).  The first split is at `LoyalCH` = 0.48285.  Here is a diagram of the full (unpruned) tree.

```{r}
rpart.plot(oj_mdl_cart_full, yesno = TRUE)
```

The boxes show the node classification (based on mode), the proportion of observations that are *not* `CH`, and the proportion of observations included in the node. 

`rpart()` not only grew the full tree, it identified the set of cost complexity parameters, and measured the model performance of each corresponding tree using cross-validation. `printcp()` displays the candidate $c_p$ values. You can use this table to decide how to prune the tree.

```{r}
printcp(oj_mdl_cart_full)
```

There are 4 $c_p$ values in this model.  The model with the smallest complexity parameter allows the most splits (`nsplit`).  The highest complexity parameter corresponds to a tree with just a root node.  `rel error` is the error rate relative to the root node.  The root node absolute error is 0.38973162 (the proportion of MM), so its `rel error` is 0.38973162/0.38973162 = 1.0.  That means the absolute error of the full tree (at CP = 0.01) is 0.42814 * 0.38973162 = 0.1669.  You can verify that by calculating the error rate of the predicted values:

```{r}
data.frame(pred = predict(oj_mdl_cart_full, newdata = oj_train, type = "class")) %>%
   mutate(obs = oj_train$Purchase,
          err = if_else(pred != obs, 1, 0)) %>%
   summarize(mean_err = mean(err))
```

Finishing the CP table tour, `xerror` is the relative cross-validated error rate and `xstd` is its standard error. If you want the lowest possible error, then prune to the tree with the smallest relative CV error, $c_p$ = 0.01.  If you want to balance predictive power with simplicity, prune to the smallest tree within 1 SE of the one with the smallest relative error.  The CP table is not super-helpful for finding that tree, so add a column to find it.

```{r}
oj_mdl_cart_full$cptable %>%
   data.frame() %>%
   mutate(
      min_idx = which.min(oj_mdl_cart_full$cptable[, "xerror"]),
      rownum = row_number(),
      xerror_cap = oj_mdl_cart_full$cptable[min_idx, "xerror"] + 
                   oj_mdl_cart_full$cptable[min_idx, "xstd"],
      eval = case_when(rownum == min_idx ~ "min xerror",
                       xerror < xerror_cap ~ "under cap",
                       TRUE ~ "")
   ) %>%
   select(-rownum, -min_idx) 
```

The simplest tree using the 1-SE rule is $c_p = 0.01347305, CV error = `r round(0.4700599*0.38973162, 4)`).  Fortunately, `plotcp()` presents a nice graphical representation of the relationship between `xerror` and `cp`.

```{r}
plotcp(oj_mdl_cart_full, upper = "splits")
```

The dashed line is set at the minimum `xerror` + `xstd`.  The top axis shows the number of splits in the tree.  I'm not sure why the CP values are not the same as in the table (they are close, but not the same).  The figure suggests I should prune to 5 or 3 splits.  I see this curve never really hits a minimum - it is still decreasing at 5 splits.  The default tuning parameter value `cp = 0.01` may be too large, so I'll set it to `cp = 0.001` and start over.

```{r}
set.seed(123)
oj_mdl_cart_full <- rpart(
   formula = Purchase ~ .,
   data = oj_train,
   method = "class",
   cp = 0.001
   )
print(oj_mdl_cart_full)
```

This is a much larger tree.  Did I find a `cp` value that produces a local min?

```{r}
plotcp(oj_mdl_cart_full, upper = "splits")
```

Yes, the min is at CP = 0.011 with 5 splits.  The min + 1 SE is at CP = 0.021 with 3 splits.  I'll prune the tree to 3 splits.

```{r}
oj_mdl_cart <- prune(
   oj_mdl_cart_full,
   cp = oj_mdl_cart_full$cptable[oj_mdl_cart_full$cptable[, 2] == 3, "CP"]
)
rpart.plot(oj_mdl_cart, yesno = TRUE)
```

The most "important" indicator of `Purchase` appears to be `LoyalCH`.  From the **rpart** [vignette](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf) (page 12), 

> "An overall measure of variable importance is the sum of the goodness of split measures for each split for which it was the primary variable, plus goodness (adjusted agreement) for all splits in which it was a surrogate."

Surrogates refer to alternative features for a node to handle missing data. For each split, CART evaluates a variety of alternative "surrogate" splits to use when the feature value for the primary split is NA. Surrogate splits are splits that produce results similar to the original split. 

A variable's importance is the sum of the improvement in the overall Gini (or RMSE) measure produced by the nodes in which it appears. Here is the variable importance for this model.

```{r}
oj_mdl_cart$variable.importance %>% 
   data.frame() %>%
   rownames_to_column(var = "Feature") %>%
   rename(Overall = '.') %>%
   ggplot(aes(x = fct_reorder(Feature, Overall), y = Overall)) +
   geom_pointrange(aes(ymin = 0, ymax = Overall), color = "cadetblue", size = .3) +
   theme_minimal() +
   coord_flip() +
   labs(x = "", y = "", title = "Variable Importance with Simple Classication")
```

`LoyalCH` is by far the most important variable, as expected from its position at the top of the tree, and one level down. 

You can see how the surrogates appear in the model with the `summary()` function.

```{r}
summary(oj_mdl_cart)
```

I'll evaluate the predictions and record the accuracy (correct classification percentage) for comparison to other models. Two ways to evaluate the model are the confusion matrix, and the ROC curve. 


### Holdout Performance

#### Confusion Matrix

Print the confusion matrix with `caret::confusionMatrix()` to see how well does this model performs against the holdout set.

```{r}
oj_preds_cart <- bind_cols(
   predict(oj_mdl_cart, newdata = oj_test, type = "prob"),
   Pred = predict(oj_mdl_cart, newdata = oj_test, type = "class"),
   Purchase = oj_test$Purchase
)

oj_cm_cart <- confusionMatrix(oj_preds_cart$Pred, reference = oj_test$Purchase)
oj_cm_cart
```

The confusion matrix is at the top.  It also includes a lot of statistics.  It's worth getting familiar with the stats.  The model accuracy and 95% CI are calculated from the binomial test.

```{r}
binom.test(x = 113 + 70, n = 213)
```

The "No Information Rate" (NIR) statistic is the class rate for the largest class.  In this case CH is the largest class, so NIR = 130/213 = 0.6103.  "P-Value [Acc > NIR]" is the binomial test that the model accuracy is significantly better than the NIR (i.e., significantly better than just always guessing CH).

```{r}
binom.test(x = 113 + 70, n = 213, p = 130/213, alternative = "greater")
```

The "Accuracy" statistic indicates the model predicts 0.8590 of the observations correctly.  That's good, but less impressive when you consider the prevalence of CH is 0.6103 - you could achieve 61% accuracy just by predicting CH every time. A measure that controls for the prevalence is Cohen's kappa statistic. The kappa statistic is explained [here](https://standardwisdom.com/softwarejournal/2011/12/confusion-matrix-another-single-value-metric-kappa-statistic/). It compares the accuracy to the accuracy of a "random system".  It is defined as

$$\kappa = \frac{Acc - RA}{1-RA}$$

where 

$$RA = \frac{ActFalse \times PredFalse + ActTrue \times PredTrue}{Total \times Total}$$

is the hypothetical probability of a chance agreement.  ActFalse will be the number of "MM" (13 + 70 = 83) and actual true will be the number of "CH" (113 + 17 = 130).  The predicted counts are

```{r}
table(oj_preds_cart$Pred)
```

So, $RA = (83*87 + 130*126) / 213^2 = 0.5202$ and $\kappa = (0.8592 - 0.5202)/(1 - 0.5202) = 0.7064$.  The kappa statistic varies from 0 to 1 where 0 means accurate predictions occur merely by chance, and 1 means the predictions are in perfect agreement with the observations.  In this case, a kappa statistic of 0.7064 is "substantial".  See chart [here](https://www.statisticshowto.datasciencecentral.com/cohens-kappa-statistic/).

The other measures from the `confusionMatrix()` output are various proportions and you can remind yourself of their definitions in the documentation with `?confusionMatrix`.

Visuals are almost always helpful.  Here is a plot of the confusion matrix.

```{r}
plot(oj_test$Purchase, oj_preds_cart$Pred, 
     main = "Simple Classification: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")
```

#### ROC Curve

The ROC (receiver operating characteristics) curve [@Fawcett2005] is another measure of accuracy.  The ROC curve is a plot of the true positive rate (TPR, sensitivity) versus the false positive rate (FPR, 1 - specificity) for a set of thresholds. By default, the threshold for predicting the default classification is 0.50, but it could be any threshold. `precrec::evalmod()` calculates the confusion matrix values from the model using the holdout data set.  The AUC on the holdout set is 0.8848.  `pRoc::plot.roc()`, `plotROC::geom_roc()`, and `yardstick::roc_curve()` are all options for plotting a ROC curve.

```{r}
precrec::evalmod(scores = oj_preds_cart$CH, labels = oj_preds_cart$Purchase, posclass = "CH")

yardstick::roc_curve(oj_preds_cart, Purchase, CH) %>%
  autoplot() +
  labs(
    title = "OJ CART ROC Curve",
    subtitle = "AUC = 0.8848"
  )
```

A few points on the ROC space are helpful for understanding how to use it.  

* The lower left point (0, 0) is the result of *always* predicting "negative" or in this case "MM" if "CH" is taken as the default class. No false positives, but no true positives either.
* The upper right point (1, 1) is the result of *always* predicting "positive" ("CH" here).  You catch all true positives, but miss all the true negatives.
* The upper left point (0, 1) is the result of perfect accuracy.
* The lower right point (1, 0) is the result of perfect imbecility.  You made the exact wrong prediction every time. 
* The 45 degree diagonal is the result of randomly guessing positive (CH) X percent of the time.  If you guess positive 90% of the time and the prevalence is 50%, your TPR will be 90% and your FPR will also be 90%, etc.

The goal is for all nodes to bunch up in the upper left.

Points to the left of the diagonal with a low TPR can be thought of as "conservative" predictors - they only make positive (CH) predictions with strong evidence.  Points to the left of the diagonal with a high TPR can be thought of as "liberal" predictors - they make positive (CH) predictions with weak evidence.

#### Gain Curve

The gain curve plots the cumulative summed true outcome versus the fraction of items seen when sorted by the predicted value. The “wizard” curve is the gain curve when the data is sorted by the true outcome. If the model’s gain curve is close to the wizard curve, then the model predicted the response variable well. The gray area is the “gain” over a random prediction.

130 of the 213 consumers in the holdout set purchased CH.

* The gain curve encountered 77 CH purchasers (59%) within the first 79 observations (37%). 

* It encountered all 130 CH purchasers on the 213th observation (100%).

* The bottom of the gray area is the outcome of a random model. Only half the CH purchasers would be observed within 50% of the observations. The top of the gray area is the outcome of the perfect model, the “wizard curve”. Half the CH purchasers would be observed in 65/213=31% of the observations.

```{r}
yardstick::gain_curve(oj_preds_cart, Purchase, CH) %>%
  autoplot() +
  labs(
    title = "OJ CART Gain Curve"
  )
```


### Caret Approach

I can also fit the model with `caret::train()`. There are two ways to tune hyperparameters in `train()`: 

* set the number of tuning parameter values to consider by setting `tuneLength`, or
* set particular values to consider for each parameter by defining a `tuneGrid`.

I'll build the model using 10-fold cross-validation to optimize the hyperparameter CP. If you have no idea what is the optimal tuning parameter, start with `tuneLength` to get close, then fine-tune with `tuneGrid`.  That's what I'll do.  I'll create a training control object that I can re-use in other model builds.  

```{r}
oj_trControl = trainControl(
   method = "cv",
   number = 10,
   savePredictions = "final",       # save predictions for the optimal tuning parameter
   classProbs = TRUE  # return class probabilities in addition to predicted values
   )
```

Now fit the model.

```{r}
set.seed(1234)
oj_mdl_cart2 <- train(
   Purchase ~ ., 
   data = oj_train, 
   method = "rpart",
   tuneLength = 5,
   metric = "Accuracy",
   trControl = oj_trControl
   )
```

`caret` built a full tree using `rpart`'s default parameters: gini splitting index, at least 20 observations in a node in order to consider splitting it, and at least 6 observations in each node.  Caret then calculated the accuracy for each candidate value of $\alpha$.  Here is the results.

```{r}
print(oj_mdl_cart2)
```

The second `cp` (0.008982036) produced the highest accuracy.  I can drill into the best value of `cp` using a tuning grid.

```{r}
set.seed(1234)
oj_mdl_cart2 <- train(
   Purchase ~ ., 
   data = oj_train, 
   method = "rpart",
   tuneGrid = expand.grid(cp = seq(from = 0.001, to = 0.010, length = 11)),  
   metric='Accuracy',
   trControl = oj_trControl
   )
print(oj_mdl_cart2)
```

The best model is at cp = 0.0082.  Here are the cross-validated accuracies for the candidate cp values.

```{r}
plot(oj_mdl_cart2)
```

Here are the rules in the final model.  

```{r}
oj_mdl_cart2$finalModel
```

```{r}
rpart.plot(oj_mdl_cart2$finalModel)
```

Let's look at the performance on the holdout data set.

```{r}
oj_preds_cart2 <- bind_cols(
   predict(oj_mdl_cart2, newdata = oj_test, type = "prob"),
   Pred = predict(oj_mdl_cart2, newdata = oj_test, type = "raw"),
   Purchase = oj_test$Purchase
)

oj_cm_cart2 <- confusionMatrix(oj_preds_cart2$Pred, oj_test$Purchase)
oj_cm_cart2
```

The accuracy is 0.8451 - a little worse than the 0.8592 from the direct method.

```{r}
precrec::evalmod(scores = oj_preds_cart2$CH, labels = oj_preds_cart2$Purchase, posclass = "CH")

yardstick::roc_curve(oj_preds_cart2, Purchase, CH) %>%
  autoplot() +
  labs(
    title = "OJ CART ROC Curve (caret)",
    subtitle = "AUC = 0.9104"
  )
```

The AUC is 0.9104.

```{r}
yardstick::gain_curve(oj_preds_cart2, Purchase, CH) %>%
  autoplot() +
  labs(title = "OJ CART Gain Curve (caret)")
```

Finally, here is the variable importance plot. Brand loyalty is most important, followed by price difference.

```{r}
plot(varImp(oj_mdl_cart2), main="Variable Importance with CART (caret)")
```

Looks like the manual effort fared best.  Here is a summary the accuracy rates of the two models.

```{r}
scoreboard <- rbind(
   data.frame(model = "Manual Class", Acc = round(oj_cm_cart$overall["Accuracy"], 5)), 
   data.frame(model = "Caret w/tuneGrid", Acc = round(oj_cm_cart2$overall["Accuracy"], 5))
)
scoreboard
```


## Regression Tree

A simple regression tree is built in a manner similar to a simple classification tree, and like the simple classification tree, it is rarely invoked on its own; the bagged, random forest, and gradient boosting methods build on this logic.  I'll learn by example again. Using the `ISLR::Carseats` data set, and predict `Sales` using from the 10 feature variables.

```{r}
carseats_dat <- ISLR::Carseats
skimr::skim(carseats_dat)
```

Split `careseats_dat` (n = 400) into `carseats_train` (80%, n = 321) and `carseats_test` (20%, n = 79).

```{r}
set.seed(12345)
partition <- createDataPartition(y = carseats_dat$Sales, p = 0.8, list = FALSE)
carseats_train <- carseats_dat[partition, ]
carseats_test <- carseats_dat[-partition, ]
```

The first step is to build a full tree, then perform k-fold cross-validation to help select the optimal cost complexity (cp). The only difference here is the `rpart()` parameter `method = "anova"` to produce a regression tree.

```{r}
set.seed(1234)
carseats_mdl_cart_full <- rpart(Sales ~ ., carseats_train, method = "anova")
# carseats_mdl_1 <- rpart(
#    formula = Sales ~ .,
#    data = carseats_train,
#    method = "anova", 
#    xval = 10,
#    model = TRUE  # to plot splits with factor variables.
# )
print(carseats_mdl_cart_full)
```

The predicted `Sales` at the root is the mean `Sales` for the training data set, 7.535950 (values are $000s). The deviance at the root is the SSE, 2567.768. The first split is at `ShelveLoc` = [Bad, Medium] vs Good.  Here is the unpruned tree diagram.

```{r}
rpart.plot(carseats_mdl_cart_full, yesno = TRUE)
```

The boxes show the node predicted value (mean) and the proportion of observations that are in the node (or child nodes). 

`rpart()` grew the full tree, and used cross-validation to test the performance of the possible complexity hyperparameters. `printcp()` displays the candidate cp values. You can use this table to decide how to prune the tree.

```{r}
printcp(carseats_mdl_cart_full)
```

There were 16 possible cp values in this model. The model with the smallest complexity parameter allows the most splits (`nsplit`).  The highest complexity parameter corresponds to a tree with just a root node.  `rel error` is the SSE relative to the root node.  The root node SSE is 2567.76800, so its `rel error` is 2567.76800/2567.76800 = 1.0.  That means the absolute error of the full tree (at CP = 0.01) is 0.30963 * 2567.76800 = 795.058. You can verify that by calculating the SSE of the model predicted values:

```{r}
data.frame(pred = predict(carseats_mdl_cart_full, newdata = carseats_train)) %>%
   mutate(obs = carseats_train$Sales,
          sq_err = (obs - pred)^2) %>%
   summarize(sse = sum(sq_err))
```

Finishing the CP table tour, `xerror` is the cross-validated SSE and `xstd` is its standard error. If you want the lowest possible error, then prune to the tree with the smallest relative SSE (`xerror`).  If you want to balance predictive power with simplicity, prune to the smallest tree within 1 SE of the one with the smallest relative SSE.  The CP table is not super-helpful for finding that tree. I'll add a column to find it.

```{r}
carseats_mdl_cart_full$cptable %>%
   data.frame() %>%
   mutate(min_xerror_idx = which.min(carseats_mdl_cart_full$cptable[, "xerror"]),
          rownum = row_number(),
          xerror_cap = carseats_mdl_cart_full$cptable[min_xerror_idx, "xerror"] + 
             carseats_mdl_cart_full$cptable[min_xerror_idx, "xstd"],
          eval = case_when(rownum == min_xerror_idx ~ "min xerror",
                           xerror < xerror_cap ~ "under cap",
                           TRUE ~ "")) %>%
   select(-rownum, -min_xerror_idx) 
```

Okay, so the simplest tree is the one with CP = 0.02599265 (5 splits).  Fortunately, `plotcp()` presents a nice graphical representation of the relationship between `xerror` and `cp`.

```{r}
plotcp(carseats_mdl_cart_full, upper = "splits")
```

The dashed line is set at the minimum `xerror` + `xstd`.  The top axis shows the number of splits in the tree.  I'm not sure why the CP values are not the same as in the table (they are close, but not the same).  The smallest relative error is at `r carseats_mdl_1$cptable[which.min(carseats_mdl_1$cptable[, "xerror"]), "CP"]`, but the maximum CP below the dashed line (one standard deviation above the minimum error) is at CP = 0.02599265 (5 splits). Use the `prune()` function to prune the tree by specifying the associated cost-complexity `cp`.  

```{r}
carseats_mdl_cart <- prune(
   carseats_mdl_cart_full,
   cp = carseats_mdl_cart_full$cptable[carseats_mdl_cart_full$cptable[, 2] == 5, "CP"]
)
rpart.plot(carseats_mdl_cart, yesno = TRUE)
```

The most "important" indicator of `Sales` is `ShelveLoc`.  Here are the importance values from the model. 

```{r}
carseats_mdl_cart$variable.importance %>% 
   data.frame() %>%
   rownames_to_column(var = "Feature") %>%
   rename(Overall = '.') %>%
   ggplot(aes(x = fct_reorder(Feature, Overall), y = Overall)) +
   geom_pointrange(aes(ymin = 0, ymax = Overall), color = "cadetblue", size = .3) +
   theme_minimal() +
   coord_flip() +
   labs(x = "", y = "", title = "Variable Importance with Simple Regression")
```

The most important indicator of `Sales` is `ShelveLoc`, then `Price`, then `Age`, all of which appear in the final model.  `CompPrice` was also important.

The last step is to make predictions on the validation data set. The root mean squared error ($RMSE = \sqrt{(1/2) \sum{(actual - pred)^2}})$ and mean absolute error ($MAE = (1/n) \sum{|actual - pred|}$) are the two most common measures of predictive accuracy. The key difference is that RMSE punishes large errors more harshly. For a regression tree, set argument `type = "vector"` (or do not specify at all). 

```{r message=FALSE, warning=FALSE}
carseats_preds_cart <- predict(carseats_mdl_cart, carseats_test, type = "vector")

carseats_rmse_cart <- RMSE(
   pred = carseats_preds_cart,
   obs = carseats_test$Sales
)
carseats_rmse_cart
```

The pruning process leads to an average prediction error of 2.363 in the test data set.  Not too bad considering the standard deviation of `Sales` is `r round(sd(carseats_test$Sales), 3)`. Here is a predicted vs actual plot. 

```{r}
data.frame(Predicted = carseats_preds_cart, Actual = carseats_test$Sales) %>%
   ggplot(aes(x = Actual, y = Predicted)) +
   geom_point(alpha = 0.6, color = "cadetblue") +
   geom_abline(intercept = 0, slope = 1, linetype = 2) +
   labs(title = "Carseats CART, Predicted vs Actual")
```

The 6 possible predicted values do a decent job of binning the observations.


### Caret Approach

I can also fit the model with `caret::train()`, specifying `method = "rpart"`. I'll build the model using 10-fold cross-validation to optimize the hyperparameter CP. 

```{r}
carseats_trControl = trainControl(
   method = "cv",
   number = 10,
   savePredictions = "final"       # save predictions for the optimal tuning parameter
)
```


I'll let the model look for the best CP tuning parameter with `tuneLength` to get close, then fine-tune with `tuneGrid`.  

```{r}
set.seed(1234)
carseats_mdl_2 = train(
   Sales ~ ., 
   data = carseats_train, 
   method = "rpart",  # for classification tree
   tuneLength = 5,  # choose up to 5 combinations of tuning parameters (cp)
   metric = "RMSE",  # evaluate hyperparamter combinations with RMSE
   trControl = carseats_trControl
)
print(carseats_mdl_2)
```

The first `cp` (0.04167149) produced the smallest RMSE.  I can drill into the best value of `cp` using a tuning grid.  I'll try that now.

```{r}
myGrid <-  expand.grid(cp = seq(from = 0, to = 0.1, by = 0.01))
carseats_mdl_3 = train(
   Sales ~ ., 
   data = carseats_train, 
   method = "rpart",  # for classification tree
   tuneGrid = myGrid,  # choose up to 5 combinations of tuning parameters (cp)
   metric = "RMSE",  # evaluate hyperparamter combinations with RMSE
   trControl = carseats_trControl
)
print(carseats_mdl_3)
```

It looks like the best performing tree is the unpruned one.

```{r}
plot(carseats_mdl_3)
```

Lets's see the final model.

```{r}
rpart.plot(carseats_mdl_3$finalModel)
```

What were the most important variables?
```{r}
plot(varImp(carseats_mdl_3), main="Variable Importance with Simple Regression")
```

Evaluate the model by making predictions with the test data set.  

```{r}
carseats_mdl_3_preds <- predict(carseats_mdl_3, carseats_test, type = "raw")
data.frame(Actual = carseats_test$Sales, Predicted = carseats_mdl_3_preds) %>%
ggplot(aes(x = Actual, y = Predicted)) +
   geom_point() +
   geom_smooth() +
   geom_abline(slope = 1, intercept = 0) + 
   scale_y_continuous(limits = c(0, 15)) +
   labs(title = "Simple Regression: Predicted vs. Actual")
```

Looks like the model over-estimates at the low end and undestimates at the high end.  Calculate the test data set RMSE.

```{r}
carseats_mdl_3_pruned_rmse <- RMSE(
   pred = carseats_mdl_3_preds,
   obs = carseats_test$Sales
)
carseats_mdl_3_pruned_rmse
```


Caret faired better in this model.  Here is a summary the RMSE values of the two models.

```{r}
rbind(data.frame(model = "Manual ANOVA", 
                 RMSE = round(carseats_mdl_1_pruned_rmse, 5)), 
      data.frame(model = "Caret", 
                 RMSE = round(carseats_mdl_3_pruned_rmse, 5))
)
```


## Bagging

Bootstrap aggregation, or *bagging*, is a general-purpose procedure for reducing the variance of a statistical learning method.  The algorithm constructs *B* regression trees using *B* bootstrapped training sets, and averages the resulting predictions. These trees are grown deep, and are not pruned. Hence each individual tree has high variance, but low bias. Averaging these *B* trees reduces the variance.  For classification trees, bagging takes the "majority vote" for the prediction.  Use a value of *B* sufficiently large that the error has settled down.

To test the model accuracy, the out-of-bag observations are predicted from the models that do not use them.  If B/3 of observations are in-bag, there are *B/3* predictions per observation.  These predictions are averaged for the test prediction.  Again, for classification trees, a majority vote is taken.

The downside to bagging is that it improves accuracy at the expense of interpretability.  There is no longer a single tree to interpret, so it is no longer clear which variables are more important than others.  

Bagged trees are a special case of random forests, so see the next section for an example.


## Random Forests

Random forests improve bagged trees by way of a small tweak that de-correlates the trees.  As in bagging, the algorithm builds a number of decision trees on bootstrapped training samples. But when building these decision trees, each time a split in a tree is considered, a random sample of *mtry* predictors is chosen as split candidates from the full set of *p* predictors.  A fresh sample of *mtry* predictors is taken at each split.  Typically $mtry \sim \sqrt{b}$.  Bagged trees are thus a special case of random forests where *mtry = p*.


#### Bagging Classification Example

Again using the `OJ` data set to predict `Purchase`, this time I'll use the bagging method by specifying `method = "treebag"`.  I'll use `tuneLength = 5` and not worry about `tuneGrid` anymore.  Caret has no hyperparameters to tune with this model. 

```{r}
oj.bag = train(Purchase ~ ., 
               data = oj_train, 
               method = "treebag",  # for bagging
               tuneLength = 5,  # choose up to 5 combinations of tuning parameters
               metric = "ROC",  # evaluate hyperparamter combinations with ROC
               trControl = trainControl(
                 method = "cv",  # k-fold cross validation
                 number = 10,  # k=10 folds
                 savePredictions = "final",       # save predictions for the optimal tuning parameters
                      classProbs = TRUE,  # return class probabilities in addition to predicted values
                      summaryFunction = twoClassSummary  # for binary response variable
                      )
                    )
oj.bag
#plot(oj.bag$)
oj.pred <- predict(oj.bag, oj_test, type = "raw")
plot(oj_test$Purchase, oj.pred, 
     main = "Bagging Classification: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")

(oj.conf <- confusionMatrix(data = oj.pred, 
                            reference = oj_test$Purchase))
oj.bag.acc <- as.numeric(oj.conf$overall[1])
rm(oj.pred)
rm(oj.conf)
#plot(oj.bag$, oj.bag$finalModel$y)
plot(varImp(oj.bag), main="Variable Importance with Simple Classication")
```

#### Random Forest Classification Example

Now I'll try it with the random forest method by specifying `method = "ranger"`.  I'll stick with `tuneLength = 5`.  Caret tunes three hyperparameters: 

* `mtry`: number of randomly selected predictors.  Default is sqrt(p).
* `splitrule`: splitting rule.  For classification, options are "gini" (default) and "extratrees".
* `min.node.size`: minimal node size.  Default is 1 for classification.

```{r}
oj.frst = train(Purchase ~ ., 
               data = oj_train, 
               method = "ranger",  # for random forest
               tuneLength = 5,  # choose up to 5 combinations of tuning parameters
               metric = "ROC",  # evaluate hyperparamter combinations with ROC
               trControl = trainControl(
                 method = "cv",  # k-fold cross validation
                 number = 10,  # 10 folds
                 savePredictions = "final",       # save predictions for the optimal tuning parameter1
                 classProbs = TRUE,  # return class probabilities in addition to predicted values
                 summaryFunction = twoClassSummary  # for binary response variable
                 )
               )
oj.frst
plot(oj.frst)
oj.pred <- predict(oj.frst, oj_test, type = "raw")
plot(oj_test$Purchase, oj.pred, 
     main = "Random Forest Classification: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")

(oj.conf <- confusionMatrix(data = oj.pred, 
                            reference = oj_test$Purchase))
oj.frst.acc <- as.numeric(oj.conf$overall[1])
rm(oj.pred)
rm(oj.conf)
#plot(oj.bag$, oj.bag$finalModel$y)
#plot(varImp(oj.frst), main="Variable Importance with Simple Classication")
```

The model algorithm explains *"ROC was used to select the optimal model using the largest value. The final values used for the model were mtry = 9, splitrule = extratrees and min.node.size = 1."*  You can see the results of tuning grid combinations in the associated plot of ROC AUC vs mtry grouped by splitting rule.

The bagging (accuracy = 0.80751) and random forest (accuracy = 0.81690) models faired pretty well, but the manual classification tree is still in first place.  There's still gradient boosting to investigate!

```{r}
rbind(data.frame(model = "Manual Class", Accuracy = round(oj_cm_cart$overall["Accuracy"], 5)), 
      data.frame(model = "Caret w.tuneGrid", Accuracy = round(oj_model_3_cm$overall["Accuracy"], 5)),
      data.frame(model = "Bagging", Accuracy = round(oj.bag.acc, 5)),
      data.frame(model = "Random Forest", Accuracy = round(oj.frst.acc, 5))
) %>% arrange(desc(Accuracy))
```

#### Bagging Regression Example

Again using the `Carseats` data set to predict `Sales`, this time I'll use the bagging method by specifying `method = "treebag"`.  I'll use `tuneLength = 5` and not worry about `tuneGrid` anymore.  Caret has no hyperparameters to tune with this model. 

```{r}
carseats.bag = train(Sales ~ ., 
               data = carseats_train, 
               method = "treebag",  # for bagging
               tuneLength = 5,  # choose up to 5 combinations of tuning parameters
               metric = "RMSE",  # evaluate hyperparamter combinations with RMSE
               trControl = trainControl(
                 method = "cv",  # k-fold cross validation
                 number = 10,  # 10 folds
                 savePredictions = "final"       # save predictions for the optimal tuning parameter1
                 )
               )
carseats.bag
#plot(carseats.bag$finalModel)
carseats.pred <- predict(carseats.bag, carseats_test, type = "raw")
plot(carseats_test$Sales, carseats.pred, 
     main = "Bagging Regression: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")
abline(0, 1)

(carseats.bag.rmse <- RMSE(pred = carseats.pred,
                           obs = carseats_test$Sales))
rm(carseats.pred)
plot(varImp(carseats.bag), main="Variable Importance with Regression Bagging")
```


#### Random Forest Regression Example

Now I'll try it with the random forest method by specifying `method = "ranger"`.  I'll stick with `tuneLength = 5`.  Caret tunes three hyperparameters: 

* `mtry`: number of randomly selected predictors
* `splitrule`: splitting rule.  For regression, options are "variance" (default), "extratrees", and "maxstat". 
* `min.node.size`: minimal node size

```{r}
carseats.frst = train(Sales ~ ., 
               data = carseats_train, 
               method = "ranger",  # for random forest
               tuneLength = 5,  # choose up to 5 combinations of tuning parameters
               metric = "RMSE",  # evaluate hyperparamter combinations with RMSE
               trControl = trainControl(
                 method = "cv",  # k-fold cross validation
                 number = 10,  # 10 folds
                 savePredictions = "final"       # save predictions for the optimal tuning parameter1
                 )
               )
carseats.frst
plot(carseats.frst)
carseats.pred <- predict(carseats.frst, carseats_test, type = "raw")
plot(carseats_test$Sales, carseats.pred, 
     main = "Random Forest Regression: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")
abline(0, 1)
(carseats.frst.rmse <- RMSE(pred = carseats.pred,
                           obs = carseats_test$Sales))
rm(carseats.pred)
#plot(varImp(carseats.frst), main="Variable Importance with Regression Random Forest")
```

The model algorithm explains *"RMSE was used to select the optimal model using the smallest value. The final values used for the model were mtry = 11, splitrule = variance and min.node.size = 5."*  You can see the results of tuning grid combinations in the associated plot of ROC AUC vs mtry grouped by splitting rule.

The bagging and random forest models faired very well - they took over the first and second place!

```{r}
rbind(data.frame(model = "Manual ANOVA", RMSE = round(carseats_mdl_1_pruned_rmse, 5)), 
      data.frame(model = "ANOVA w.tuneGrid", RMSE = round(carseats_mdl_3_pruned_rmse, 5)),
      data.frame(model = "Bagging", RMSE = round(carseats.bag.rmse, 5)),
      data.frame(model = "Random Forest", RMSE = round(carseats.frst.rmse, 5))
) %>% arrange(RMSE)
```


## Gradient Boosting

Boosting is a method to improve (boost) the week learners sequentially and increase the model accuracy with a combined model. There are several boosting algorithms.  One of the earliest was AdaBoost (adaptive boost).  A more recent innovation is gradient boosting.

Adaboost creates a single split tree (decision stump) then weights the observations by how well the initial tree performed, putting more weight on the difficult observations.  It then creates a second tree using the weights so that it focuses on the difficult observations.  Observations that are difficult to classify receive increasing larger weights until the algorithm identifies a model that correctly classifies them.  The final model returns predictions that are a majority vode. (*I think Adaboost applies only to classification problems, not regressions*). 
 
Gradient boosting generalizes the AdaBoost method, so that the object is to minimize a loss function.  In the case of classification problems, the loss function is the log-loss; for regression problems, the loss function is mean squared error.  The regression trees are addative, so that the successive models can be added together to correct the residuals in the earlier models.  Gradient boosting constructs its trees in a "greedy" manner, meaning it chooses the best splits based on purity scores like Gini or minimizing the loss.  It is common to constrain the weak learners by setting maximum tree size parameters.  Gradient boosting continues until it reaches maximum number of trees or an acceptible error level.  This can result in overfitting, so it is common to employ regularization methods that penalize aspects of the model.

**Tree Constraints**.  In general the more constrained the tree, the more trees need to be grown.  Parameters to optimize include number of trees, tree depth, number of nodes, minimmum observations per split, and minimum improvement to loss.

**Learning Rate**.  Each successive tree can be weighted to slow down the learning rate.  Decreasing the learning rate increases the number of required trees.  Common growth rates are 0.1 to 0.3.

The gradient boosting algorithm fits a shallow tree $T_1$ to the data, $M_1 = T_1$.  Then it fits a tree $T_2$ to the residuals and adds a weighted sum of the tree to the original tree as $M_2 = M_1 + \gamma T_2$.  For regularized boosting, include a learning rate factor $\eta \in (0..1)$, $M_2 = M_1 + \eta \gamma T_2$.  A larger $\eta$ produces faster learning, but risks overfitting.  The process repeats until the residuals are small enough, or until it reaches the maximum iterations.  Because overfitting is a risk, use cross-validation to select the appropriate number of trees (the number of trees producing the lowest RMSE).


#### Gradient Boosting Classification Example

Again using the `OJ` data set to predict `Purchase`, this time I'll use the gradient boosting method by specifying `method = "gbm"`.  I'll use `tuneLength = 5` and not worry about `tuneGrid` anymore.  Caret tunes the following  hyperparameters (see `modelLookup("gbm")`). 

* `n.trees`: number of boosting iterations
* `interaction.depth`: maximum tree depth
* `shrinkage`: shrinkage
* `n.minobsinnode`: mimimum terminal node size

```{r}
oj.gbm <- train(Purchase ~ ., 
               data = oj_train, 
               method = "gbm",  # for bagged tree
               tuneLength = 5,  # choose up to 5 combinations of tuning parameters
               metric = "ROC",  # evaluate hyperparamter combinations with ROC
               trControl = trainControl(
                 method = "cv",  # k-fold cross validation
                 number = 10,  # 10 folds
                 savePredictions = "final",       # save predictions for the optimal tuning parameter1
                      classProbs = TRUE,  # return class probabilities in addition to predicted values
                      summaryFunction = twoClassSummary  # for binary response variable
                      )
                    )
oj.gbm
plot(oj.gbm)
oj.pred <- predict(oj.gbm, oj_test, type = "raw")
plot(oj_test$Purchase, oj.pred, 
     main = "Gradient Boosing Classification: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")

(oj.conf <- confusionMatrix(data = oj.pred, 
                            reference = oj_test$Purchase))
oj.gbm.acc <- as.numeric(oj.conf$overall[1])
rm(oj.pred)
rm(oj.conf)
#plot(oj.bag$, oj.bag$finalModel$y)
#plot(varImp(oj.gbm), main="Variable Importance with Gradient Boosting")
```


#### Gradient Boosting Regression Example

Again using the `Carseats` data set to predict `Sales`, this time I'll use the gradient boosting method by specifying `method = "gbm"`.  I'll use `tuneLength = 5` and not worry about `tuneGrid` anymore.  Caret tunes the following  hyperparameters. 

* `n.trees`: number of boosting iterations (increasing `n.trees` reduces the error on training set, but may lead to over-fitting)
* `interaction.depth`: maximum tree depth (the default six - node tree appears to do an excellent job)
* `shrinkage`: learning rate (reduces the impact of each additional fitted base-learner (tree) by reducing the size of incremental steps and thus penalizes the importance of each consecutive iteration.  The intuition is that it is better to improve a model by taking many small steps than by taking fewer large steps. If one of the boosting iterations turns out to be erroneous, its negative impact can be easily corrected in subsequent steps.)
* `n.minobsinnode`: mimimum terminal node size

```{r}
carseats.gbm <- train(Sales ~ ., 
                      data = carseats_train, 
                      method = "gbm",  # for bagged tree
                      tuneLength = 5,  # choose up to 5 combinations of tuning parameters
                      metric = "RMSE",  # evaluate hyperparamter combinations with ROC
                      trControl = trainControl(
                        method = "cv",  # k-fold cross validation
                        number = 10,  # 10 folds
                        savePredictions = "final",       # save predictions for the optimal tuning parameter1
                        verboseIter = FALSE,
                        returnData = FALSE
                        )
                      )
carseats.gbm
plot(carseats.gbm)
carseats.pred <- predict(carseats.gbm, carseats_test, type = "raw")
plot(carseats_test$Sales, carseats.pred, 
     main = "Gradient Boosing Regression: Predicted vs. Actual",
     xlab = "Actual",
     ylab = "Predicted")
abline(0,1)

(carseats.gbm.rmse <- RMSE(pred = carseats.pred,
                           obs = carseats_test$Sales))
rm(carseats.pred)

#plot(varImp(carseats.gbm), main="Variable Importance with Gradient Boosting")
```

## Summary

Okay, I'm going to tally up the results!  For the classification division, the winner is the manual classification tree!  Gradient boosting made a valiant run at it, but came up just a little short.

```{r}
rbind(data.frame(model = "Manual Class", Acc = round(oj_cm_cart$overall["Accuracy"], 5)), 
      data.frame(model = "Class w.tuneGrid", Acc = round(oj_model_3_cm$overall["Accuracy"], 5)),
      data.frame(model = "Bagging", Acc = round(oj.bag.acc, 5)),
      data.frame(model = "Random Forest", Acc = round(oj.frst.acc, 5)),
      data.frame(model = "Gradient Boosting", Acc = round(oj.gbm.acc, 5))
) %>% arrange(desc(Acc))
```

And now for the regression division, the winnner is... gradient boosting!

```{r}
rbind(data.frame(model = "Manual ANOVA", RMSE = round(carseats_mdl_1_pruned_rmse, 5)), 
      data.frame(model = "ANOVA w.tuneGrid", RMSE = round(carseats_mdl_3_pruned_rmse, 5)),
      data.frame(model = "Bagging", RMSE = round(carseats.bag.rmse, 5)),
      data.frame(model = "Random Forest", RMSE = round(carseats.frst.rmse, 5)),
      data.frame(model = "Gradient Boosting", RMSE = round(carseats.gbm.rmse, 5))
) %>% arrange(RMSE)
```


Here are plots of the ROC curves for all the models (one from each chapter) on the same graph. The ROCR package provides the prediction() and performance() functions which generate the data required for plotting the ROC curve, given a set of predictions and actual (true) values. 
The more "up and to the left" the ROC curve of a model is, the better the model. The AUC performance metric is literally the "Area Under the ROC Curve", so the greater the area under this curve, the higher the AUC, and the better-performing the model is.

```{r}
library(ROCR)
# List of predictions
oj.class.pred <- predict(oj_model_3, oj_test, type = "prob")[,2]
oj.bag.pred <- predict(oj.bag, oj_test, type = "prob")[,2]
oj.frst.pred <- predict(oj.frst, oj_test, type = "prob")[,2]
oj.gbm.pred <- predict(oj.gbm, oj_test, type = "prob")[,2]

preds_list <- list(oj.class.pred, oj.bag.pred, oj.frst.pred, oj.gbm.pred)
#preds_list <- list(oj.class.pred)

# List of actual values (same for all)
m <- length(preds_list)
actuals_list <- rep(list(oj_test$Purchase), m)

# Plot the ROC curves
pred <- prediction(preds_list, actuals_list)
#pred <- prediction(oj.class.pred[,2], oj_test$Purchase)
rocs <- performance(pred, "tpr", "fpr")
plot(rocs, col = as.list(1:m), main = "Test Set ROC Curves")
legend(x = "bottomright", 
       legend = c("Decision Tree", "Bagged Trees", "Random Forest", "GBM"),
       fill = 1:m)
```


## Reference

Penn State University, STAT 508: Applied Data Mining and Statistical Learning, "Lesson 11: Tree-based Methods". [https://newonlinecourses.science.psu.edu/stat508/lesson/11](https://newonlinecourses.science.psu.edu/stat508/lesson/11).

Brownlee, Jason. "Classification And Regression Trees for Machine Learning", Machine Learning Mastery.  [https://machinelearningmastery.com/classification-and-regression-trees-for-machine-learning/](https://machinelearningmastery.com/classification-and-regression-trees-for-machine-learning/). 

Brownlee, Jason. "A Gentle Introduction to the Gradient Boosting Algorithm for Machine Learning", Machine Learning Mastery. [https://machinelearningmastery.com/gentle-introduction-gradient-boosting-algorithm-machine-learning/](https://machinelearningmastery.com/gentle-introduction-gradient-boosting-algorithm-machine-learning/).

[DataCamp: Machine Learning with Tree-Based Models in R](https://campus.datacamp.com/courses/machine-learning-with-tree-based-models-in-r)

[An Introduction to Statistical Learning](http://faculty.marshall.usc.edu/gareth-james/ISL/) by Gareth James, et al.

[SAS Documentation](http://support.sas.com/documentation/cdl/en/stathpug/68163/HTML/default/viewer.htm#stathpug_hpsplit_details01.htm)

[StatMethods: Tree-Based Models](https://www.statmethods.net/advstats/cart.html)

[Machine Learning Plus](https://www.machinelearningplus.com/machine-learning/caret-package/)

[GBM (Boosted Models) Tuning Parameters](https://www.listendata.com/2015/07/gbm-boosted-models-tuning-parameters.html)  from Listen Data

[Harry Southworth](https://github.com/harrysouthworth/gbm/blob/master/demo/bernoulli.R) on GitHub

[Gradient Boosting Classification with GBM in R](https://www.datatechnotes.com/2018/03/classification-with-gradient-boosting.html) in DataTechNotes

Molnar, Christoph. "Interpretable machine learning. A Guide for Making Black Box Models Explainable", 2019. [https://christophm.github.io/interpretable-ml-book/](https://christophm.github.io/interpretable-ml-book/).


<!--chapter:end:11-cart.Rmd-->


# Non-linear Models

Placeholder


## Splines
## MARS
## GAM

<!--chapter:end:12-nonlin.Rmd-->


# Support Vector Machines

Placeholder


## Maximal Margin Classifier
## Support Vector Classifier
## Support Vector Machines
## Example
## Using Caret

<!--chapter:end:13-svm.Rmd-->

# Principal Components Analysis

<!--chapter:end:14-pca.Rmd-->


# Text Mining

Placeholder


## N-Grams
## Converting to and from non-tidy formats
## Example
## stringr package
## Regular Expressions

<!--chapter:end:16-text-mining.Rmd-->


# Survival Analysis

Placeholder


## Basic Concepts
## Survival Curve Estimation
### Kaplan-Meier
#### Log-Rank Test
### Weibull
### Cox

<!--chapter:end:17-survival_analysis.Rmd-->


# Appendix {-}

Placeholder


## Publishing to BookDown {-}
## Shiny Apps {-}
## Packages {-}
### Create a package {-}
### Document Functions with roxygen
### Create Data {-}
### Create Vignette {-}
#### Step 2: Create an R Markdown template {-}

<!--chapter:end:18-appendix.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:19-references.Rmd-->

